---
title: "The Impact of Environmental Variability on Fishers' Harvest Decisions in Chile
  using a Multi-Species Approach"
subtitle: "\\begin{center}{\\color{red}\\Large\\textbf{REALLY EARLY DRAFT – PLEASE DO NOT CITE}}\\end{center}"
author: "Felipe J. Quezada-Escalona"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  bookdown::pdf_document2:
    latex_engine: xelatex
    number_sections: true
    toc: false
    pandoc_args: ["--variable=fontsize:11pt"]
  bookdown::latex_document2: default
bibliography: bibliography.bib
csl: apa.csl # Optional: for APA-style citations
linkcolor: blue
citecolor: blue
urlcolor: blue
link-citations: TRUE 
header-includes:
  - \usepackage{setspace}
  - \onehalfspacing
  - \usepackage{indentfirst}  # Enables first-line indentation
  - \setlength{\parindent}{10pt}  # Adjusts the indentation size
  - \usepackage{authblk}
  - \usepackage{booktabs}
  - \usepackage{caption}
  - \usepackage{dcolumn}
  - \author{Felipe J. Quezada-Escalona}
  - \affil{Departmento de Economía \\ Universidad de Concepción \vspace{-48pt}}
editor_options: 
  markdown: 
    wrap: 72
abstract: "In this paper, we aim to answer how fishing decisions, aggregate catch levels, and the price of marine resources will be affected under different climatic scenarios in the multi-species small pelagic fishery (SPF) in Chile, composed by anchoveta (*Engraulis ringens*), jack mackerel (*Trachurus murphyi*), and sardine (*Strangomera bentincki*), among others. By doing this, we expect to gain a better understanding of how Chilean fishers and fishing communities will adapt to climate change. To address our research question, we will estimate a multi-species harvesting model. This model considers species' economic and biological interrelation to study the effect of climate variability on harvest decisions and substitution between species, and determine the impact of different climatic scenarios on the well-being (e.g., profits) of fishers and fishing communities in Chile. We hypothesize that when fishers have reduced access to a main target species, they will switch to the closest substitute if the expected revenue from targeting this new species exceeds the expected costs. Otherwise, the vessel would decrease fishing effort or even exit the fishery due to the lack of economically viable substitutes. Moreover, we expect that this behavior is heterogeneous depending on the geographical area of operation -- as it determines the availability of other species-- and the gear type used."
---

```{r directorio, message=FALSE, warning=FALSE, include=FALSE}
rm(list = ls())
gc()

# Definir directorios según usuario

usuario <- Sys.info()[["user"]]
# computador <- Sys.info()[["nodename"]]  # Alternativamente puedes usar esto
if (usuario == "felip") {
  dirdata <- "C:/Users/felip/OneDrive - Universidad de Concepción/FONDECYT Iniciacion/Data/"
} else if (usuario == "FACEA") {
  dirdata <- "C:/Users/FACEA/OneDrive - Universidad de Concepción/FONDECYT Iniciacion/Data/"
} else {
  stop("Usuario no reconocido. Defina el directorio correspondiente.")
}
rm(usuario)

```

# Introduction

<!-- ADD COMMENTS FROM EFD! -->

The distribution and abundance of marine resources are changing in response to environmental conditions such as global ocean warming [@Poloczanska2013-qq]. Climate change will shift species distribution in the future, leading to reduced species availability in some areas and increased availability in others [@sumaila2011]. <!-- @sumaila2011: The economic consequences of climate change on fisheries might manifest themselves through changes in the price and value of catches, fishing costs, fishers’ incomes, earnings to fishing companies, discount rates and economic rent (that is, the surplus after all costs, including ‘normal’ profits, have been covered), as well as throughout the global economy. --> The literature that studies fishers’ responses to either changes in fish availability or policies that restrict access to fisheries [e.g., @Stafford2018-pq; @Vasquez_Caballero2023-ip] has identified that they can adopt the following adaptive strategies: (i) reduce or reallocate fishing effort, either to another species or to another location [@Gonzalez-Mon2021-kj]; (ii) continue following the same strategy; or (iii) exit the fishery and find alternative employment [@Powell2022-wj]. Among these strategies, reallocating effort to alternative species has been identified as a potentially effective response to climate change [@Young2018-kk]. Diversification of target species has also been linked to reduced income variability [e.g., @Kasperski2013-jz; @Sethi2014-bn] and greater resilience to both climate shocks [@Cline2017-dp; @Fisher2021-lw] and interannual oceanographic variability [@Aguilera2015-wo; @Finkbeiner2015-bs].

This emphasis on diversification aligns with broader evidence from food-producing sectors. As @sjcruz_jmp highlights, climate variability substantially affects agriculture and fisheries, where income depends heavily on environmental fluctuations (e.g., temperature, rainfall) and market forces (e.g., input costs) [@Kasperski2013-jz; @Carter2018]. With variability expected to reduce productivity, income risk is likely to rise [@Carter2018; @Free2019]. Diversification—whether within a sector (e.g., switching crops or species) or across sectors—is often promoted as an adaptive strategy [@Abbott2023-sb]. However, these strategies can be costly for resource-dependent communities with limited capital and skills [@Cherdchuchai2006; @Ellis2000], and the role of switching costs in shaping diversification remains poorly understood.

In the fisheries context, switching between species requires not only the skills but also the appropriate gear and permits [@Frawley2021-cw; @Powell2022-wj]. Even if these conditions are met, diversification may still be constrained by port infrastructure, markets, and regulations [@Beaudreau2019-xg; @Kasperski2013-jz; @Powell2022-wj]. Therefore, deciding which adaptation strategy to adopt is not straightforward and depends on multiple factors. Moreover, fishers may respond differently to similar circumstances depending on their goals, skills, and preferences [@Zhang2011-wv; @Jardine2020-um; @Powell2022-wj].

In this research, we aim to answer how fishing decisions, aggregate catch levels, and the price of marine resources will be affected under different climatic scenarios in the multi-species small pelagic fishery (SPF) in Chile, composed of anchoveta (*Engraulis ringens*), jack mackerel (*Trachurus murphyi*), sardine (*Strangomera bentincki*), among others. The SPF is the most important in terms of catches in the country, accounting for almost 94% of the total Chilean catch in 2019 [@SUBPESCA2020]. Through this research, we aim to gain a deeper understanding of how Chilean fishers and fishing communities will adapt to climate change. According to @Cheung2010, the Chilean Exclusive Economic Zone (EEZ) is projected to experience one of the largest losses in maximum catch potential due to climate change. <!-- Caviedes, C. N. & Fik, T. J. in Climate Variability, Climate Change and Fisheries (ed. Glantz, M.) 355–375 (Cambridge Univ. Press, 1992): "During the 1997–1998 El Niño event, Chilean and Peruvian pelagic marine landings declined by about 50%, resulting in a drop in fishmeal export values by about US$8.2 billion. This huge drop generated negative economic effects and caused severe hardship (lost jobs, incomes and earnings) in both countries"--> However, the Southeast Pacific remains one of the least studied regions regarding the impacts of climate change on fisheries [@sumaila2011].

To address our research question, we will estimate a multi-species harvesting model based on @Kasperski2015-jm. This model considers species' economic and biological interrelations to study the effect of climate variability on harvest decisions and substitution between species, and to determine the impact of different climatic scenarios on the well-being (e.g., profits) of fishers and fishing communities in Chile. We expect to find significant effects of climate variables on species stock dynamics, the cost of fishing during a trip, and the number of trips a vessel takes. These environmental effects might influence optimal harvest levels and prices in local markets. <!-- We also expect to find significant interrelations between species stock and harvest, and that the composition of the catch will vary depending on the climate scenario we use for future -->
<!-- predictions. -->

Under a changing climate, studying the effect of climatic variability on fishers' harvest decisions and landings is relevant for understanding fishing communities' adaptive capacities and strategies in response to climate change, thereby enabling the design of potential mitigation measures in response to these changes by policymakers. Countries have different institutions, cultures, and norms, leading to differing responses based on the study's location. For this reason, conducting this research based on the Chilean fishing industry is necessary to develop local policies that aim to reduce climate change impacts on fisheries. While there is some literature on the effect of climate change on Chilean fisheries, I am unaware of local-level studies that consider a multiple-species framework and the interrelationship between the local market and fishing decisions seen under a variable climate context.[^1]

[^1]: For the case of Chile, as far as I know, the only article that studies fishers’ behavior using discrete choice modeling is @Pena-Torres2017-gn. This article studies how the El Niño–Southern Oscillation (ENSO) affects fishers’ location choices in the jack mackerel fishery.

Because predator–prey links couple these species, reductions in anchoveta or sardine availability may reflect not only environmental drivers but also changes in predation pressure from jack mackerel [@Alheit2004; @Arancibia2019-FIPA]. Thus, even fishers who do not target jack mackerel can face induced changes in catch rates and revenues through ecosystem feedbacks. We hypothesize that if the availability of a main target species decreases, fishers will switch to the closest substitute when expected revenues (net of switching costs) exceed expected costs; otherwise they may reduce effort or exit the fishery. We also expect cross-fleet spillovers: in Chile, jack mackerel is predominantly harvested by the industrial purse-seine fleet, whereas anchoveta and sardine have substantial artisanal and industrial participation. Shocks in one component can propagate across fleets via both biology and markets, in addition to economic linkages (bycatch constraints, shared gear, and market spillovers). This strengthens the case for a multi-species framework that models joint dynamics and substitution rather than single-species responses. Moreover, we expect that this behavior is heterogeneous depending on the geographical area of operation—as it determines the availability of other species [@Reimer2017-jw]—and the gear type used.


-- ADD the fact that the government define the TAC using sa single species model. This modelling effort allow to move further to a better management of the SPF fishery by using a multispecies approach to obtain optimal TAC

<!-- Los dos primeros, recordó, son especies clave para la cadena trófica marina, ya que alimentan a peces de mayor valor como corvina, lenguado o congrio, hoy cada vez más escasos.
LINK: https://www.diarioconcepcion.cl/economia/2025/10/07/sector-pesquero-nueva-resolucion-de-la-camara-baja-mantiene-la-discrepancia-entre-biobio-y-nuble.html -->


# SPF in Chile

The small pelagic fishery (SPF) in Chile is of critical importance to the national fisheries sector. In 2019, the SPF represented nearly 94% of total national fish landings [@SUBPESCA2020]. The fishery is primarily composed of anchoveta (*Engraulis ringens*), sardine (*Strangomera bentincki*), and jack mackerel (*Trachurus murphyi*). While in the Northern region competition mainly occurs between anchoveta and jack mackerel, in the Central-South region all three species play a major role. This makes the Central-South particularly relevant for the study of species interactions and potential substitution within a multispecies management framework, and it is therefore the focus of this research.

The jack mackerel fishery was initially concentrated in northern Chile, but since the mid-1980s the main fishing grounds have shifted to Central-South Chile, traditionally within 50 nautical miles of the coast [@Pena-Torres2017-gn]. Historically, species in the SPF have been used primarily for fishmeal and fish oil production [@Pena-Torres2017-gn]. In fact, about 85% of jack mackerel landings, on a yearly average between 1987 and 2004, were destined for reduction into fishmeal and fish oil  [@Pena-Torres2017-gn]. Today, several key ports serve as hubs for these activities, including San Antonio, Tomé, Talcahuano, San Vicente, Coronel, Lota, and Corral.

## Status of the stocks
Historically, anchoveta in the Central-South was considered collapsed until 2018, shifted to overexploited status in 2019, and has since 2020 been fished within maximum sustainable yield (MSY) limits. Meanwhile, sardine stocks have generally remained within MSY levels, except in 2021 and 2023 when they were classified as overexploited. Jack mackerel was overexploited until 2018 but has since been harvested within MSY limits. 

## Quota allocation

The Chilean fishing sector is managed primarily through a Total Allowable Catch (TAC; *Cuota Global*), which is divided between the industrial and artisanal sectors. A small share (≈2%) is reserved for research, with additional portions allocated to contingency and human consumption. The TAC is subdivided by region and season, and unused quotas may be reassigned during the fishing year.

Anchoveta and sardine are regulated as a mixed-species fishery: although each has its own quota, substitution between them is permitted. A share of industrial quota is also periodically reassigned to the artisanal sector.

Since 2013, the industrial sector has operated under an individual transferable quota (ITQ) system, known as Transferable Fishing Licenses (*Licencias Transables de Pesca, LTP*). Class A licenses were allocated based on historical catches, while Class B licenses—up to 15% of the industrial fraction—are auctioned, with the first auctions held in 2015. These sealed-bid, first-price auctions aimed to broaden access and limit concentration but have faced challenges such as low participation, difficulties in reflecting economies of scale, and signs of potential coordinated bidding [@peña_torres_2022_MRE].

The artisanal TAC operates under a regulated freedom-to-fish regime, allowing registered vessels to fish without individual quotas, except in areas where access is closed or suspended, in which case authorities may implement management measures. The main measure is the Régimen Artesanal de Extracción (RAE), which allocates the regional artisanal TAC by area, vessel size, landing site (caleta), organization, or individually, in agreement with artisanal fisher organizations. To date, area-based and organization-based allocations are the only observed schemes. Area-based allocations allow registered artisanal vessels in a given area to fish as in open access until the assigned quota is exhausted, while organization-based allocations follow the historical rights of members to distribute the organization’s quota.


- Sardine: RAE in V, VIII Y X regions? What about other species? Open access in anchovy and jack mackerel (only artisanal TAC matter at country level?)


### Chile regionalized fisheries governance framework 

Chile has a regionalize fisheres governance framework, where boat register in one region can not fish in another one. For instance, a recent conflict between the Biobío and Ñuble regions has reignited the debate over the spatial governance of the small pelagic fishery (SPF) in south–central Chile. In late August 2025, the Chilean Chamber of Deputies approved a resolution urging the Government to repeal the authorization that allows vessels from Biobío to operate in the coastal waters of Ñuble. The measure, promoted by local authorities and artisanal organizations from Ñuble, aims to protect local fishing grounds and reduce pressure on nearshore ecosystems. However, representatives from Biobío have warned that such restrictions could have severe economic consequences for the region, given its strong dependence on small pelagic landings. This episode highlights the institutional tensions arising from Chile’s regionalized fisheries governance framework, where jurisdictional boundaries often conflict with the biological and economic interdependencies of fish stocks.

## Other regulations

### Limited entry

Fishery with restricted access to new operators (just artisanal?)


### Biological closures for recruitment

- Jack mackerel is open through all year.
- Sardine and anchovy: In southern-central Chile, December–April (fixed period: January to February).

### Biological closures for reproduction

- Jack mackerel is open through all year.
- Sardine and anchovy: In southern-central Chile, July–October (fixed period: August-September).

- Seasonality? Include quarter dummies? Jack mackerel gather in the first 6 month of the year in shoals, great density in EEZ, then migrate outside 200nm ()

### Minimum size

- Jack mackerel: 26 cm 
- Sardine and anchovy?

### Maximum harvest levels

- All species: Maximum catch limit per vessel owner (LMC) for industrial vessels


```{r load_env_vars, message=FALSE, warning=FALSE, include=FALSE}

env_dt <- readRDS("data/env/EnvCoastDaily_2012_2025_0.125deg.rds")
env_dt_00_11 <- readRDS("data/env/2000-2011/EnvCoastDaily_2000_2011_0.25deg.rds")

```


```{r load_biomass, message=FALSE, warning=FALSE, include=FALSE}

biomass <- readRDS("data/biomass/biomass_dt.rds")

```

```{r load_logbook, message=FALSE, warning=FALSE, include=FALSE}

logbooks <- readRDS("data/logbooks/logbooks.rds")

```



## Switching patterns


```{r harvest_substitution_after, message=FALSE, warning=FALSE, include=FALSE}

library(dplyr)
library(data.table)
library(lubridate)
library(sf)
library(rnaturalearth)
library(ggplot2)
library(tidyr)
library(purrr)

# Build shares per vessel/fleet/year/species
vessel_species_year <- logbooks %>%
  filter(year >= 2019) %>%
  select(COD_BARCO, TIPO_FLOTA, year, NOMBRE_ESPECIE, CAPTURA_RETENIDA) %>%
  group_by(COD_BARCO, TIPO_FLOTA, year, NOMBRE_ESPECIE) %>%
  summarise(total_catch = sum(CAPTURA_RETENIDA, na.rm = TRUE), .groups = "drop") %>%
  group_by(COD_BARCO, TIPO_FLOTA, year) %>%
  mutate(species_share = total_catch / sum(total_catch, na.rm = TRUE)) %>%
  ungroup() %>%
  select(-total_catch) %>%
  complete(COD_BARCO, TIPO_FLOTA, year, NOMBRE_ESPECIE, fill = list(species_share = 0)) %>%
  group_by(COD_BARCO, TIPO_FLOTA, year) %>%
  filter(sum(species_share, na.rm = TRUE) > 0) %>%
  group_by(COD_BARCO, TIPO_FLOTA, NOMBRE_ESPECIE) %>%
  summarise(species_share = mean(species_share, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = NOMBRE_ESPECIE,
    values_from = species_share,
    values_fill = 0
  ) %>%
  # here just add missing columns if they don't exist
  { if(!"ANCHOVETA" %in% names(.)) .$ANCHOVETA <- 0; . } %>%
  { if(!"SARDINA COMUN" %in% names(.)) .$`SARDINA COMUN` <- 0; . } %>%
  { if(!"JUREL" %in% names(.)) .$JUREL <- 0; . } %>%
  mutate(
    Other = 1 - ANCHOVETA - `SARDINA COMUN` - JUREL,
    Anchoveta = ANCHOVETA,
    Sardine = `SARDINA COMUN`,
    JackMackerel = JUREL
  ) %>%
  select(COD_BARCO, TIPO_FLOTA, Anchoveta, Sardine, JackMackerel, Other)

# Strategy function
get_strategy <- function(sardine, jackmackerel, anchoveta, other) {
  species <- c()
  if (sardine > 0.15) species <- c(species, "Sardine")
  if (jackmackerel > 0.15) species <- c(species, "JackMackerel")
  if (anchoveta > 0.15) species <- c(species, "Anchoveta")
  if (other > 0.15) species <- c(species, "Other")
  
  n <- length(species)
  if (n == 0) return("None or negligible")
  if (n == 1) return(paste("Only", species[1]))
  if (n == 2) return(paste0(species[1], " and ", species[2]))
  if (n == 3) return(paste0(species[1], ", ", species[2], " and ", species[3]))
  return(paste0(species, collapse = ", "))
}

# Apply strategy classification
vessel_species_year <- vessel_species_year %>%
  mutate(strategy_After = pmap_chr(
    list(Sardine, JackMackerel, Anchoveta, Other),
    get_strategy
  ))

# Percentages by fleet type
strategy_percent <- vessel_species_year %>%
  group_by(TIPO_FLOTA, strategy_After) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(TIPO_FLOTA) %>%
  mutate(percent = round(100 * n / sum(n), 1)) %>%
  arrange(TIPO_FLOTA, desc(percent))

# Final table vessel-strategy
strategy_after <- vessel_species_year %>%
  select(COD_BARCO, TIPO_FLOTA, strategy_After)

```

```{r harvest_substitution_before, message=FALSE, warning=FALSE, include=FALSE}

# Build shares per vessel/fleet/year/species
vessel_species_year <- logbooks %>%
  filter(year < 2019) %>%
  select(COD_BARCO, TIPO_FLOTA, year, NOMBRE_ESPECIE, CAPTURA_RETENIDA) %>%
  group_by(COD_BARCO, TIPO_FLOTA, year, NOMBRE_ESPECIE) %>%
  summarise(total_catch = sum(CAPTURA_RETENIDA, na.rm = TRUE), .groups = "drop") %>%
  group_by(COD_BARCO, TIPO_FLOTA, year) %>%
  mutate(species_share = total_catch / sum(total_catch, na.rm = TRUE)) %>%
  ungroup() %>%
  select(-total_catch) %>%
  complete(COD_BARCO, TIPO_FLOTA, year, NOMBRE_ESPECIE, fill = list(species_share = 0)) %>%
  group_by(COD_BARCO, TIPO_FLOTA, year) %>%
  filter(sum(species_share, na.rm = TRUE) > 0) %>%
  group_by(COD_BARCO, TIPO_FLOTA, NOMBRE_ESPECIE) %>%
  summarise(species_share = mean(species_share, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = NOMBRE_ESPECIE,
    values_from = species_share,
    values_fill = 0
  ) %>%
  # here just add missing columns if they don't exist
  { if(!"ANCHOVETA" %in% names(.)) .$ANCHOVETA <- 0; . } %>%
  { if(!"SARDINA COMUN" %in% names(.)) .$`SARDINA COMUN` <- 0; . } %>%
  { if(!"JUREL" %in% names(.)) .$JUREL <- 0; . } %>%
  mutate(
    Other = 1 - ANCHOVETA - `SARDINA COMUN` - JUREL,
    Anchoveta = ANCHOVETA,
    Sardine = `SARDINA COMUN`,
    JackMackerel = JUREL
  ) %>%
  select(COD_BARCO, TIPO_FLOTA, Anchoveta, Sardine, JackMackerel, Other)

# Apply strategy classification
vessel_species_year <- vessel_species_year %>%
  mutate(strategy_Before = pmap_chr(
    list(Sardine, JackMackerel, Anchoveta, Other),
    get_strategy
  ))

# Percentages by fleet type
strategy_percent_before <- vessel_species_year %>%
  group_by(TIPO_FLOTA, strategy_Before) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(TIPO_FLOTA) %>%
  mutate(percent = round(100 * n / sum(n), 1)) %>%
  arrange(TIPO_FLOTA, desc(percent))

# Final table vessel-strategy
strategy_before <- vessel_species_year %>%
  select(COD_BARCO, TIPO_FLOTA, strategy_Before)

rm(vessel_species_year)

```

See Figure \@ref(fig:switchStrategy-ART) for strategy transitions. The year 2019 is used as reference as anchoveta and jack mackerel started to recover.


```{r strategy_table_logbooks_ART, message=FALSE, warning=FALSE, include=FALSE, results='asis'}

library(kableExtra)
library(stringr)

# Rename for joining
before <- strategy_percent_before %>%
  rename(strategy = strategy_Before, n_before = n, percent_before = percent)

after <- strategy_percent %>%
  rename(strategy = strategy_After, n_after = n, percent_after = percent)

# Merge strategies
strategy_summary <- full_join(before, after, by = c("strategy", "TIPO_FLOTA")) %>%
  arrange(desc(coalesce(percent_after, 0) + coalesce(percent_before, 0))) %>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

# Filter only ART
strategy_table_art <- strategy_summary %>%
  filter(TIPO_FLOTA == "ART") %>%
  ungroup() %>%
  select(-TIPO_FLOTA)

```


Table \@ref(tab:table-art) for strategy transitions.


```{r table-art, echo=FALSE, results='asis'}

# Create table with grouped columns
strategy_table_art %>%
  select(strategy, n_before, percent_before, n_after, percent_after) %>%
  kable("latex", booktabs = TRUE, digits = 1,
        col.names = c("Strategy", "n", "%", "n", "%"),
        caption = "Comparison of Strategies Before and After -- Small-scale vessels") %>%
  add_header_above(c(" " = 1, "Before" = 2, "After" = 2)) %>%
  kable_styling(font_size = 10)

rm(strategy_table_art)


```


```{r strategy-table-logbooks_IND, message=FALSE, warning=FALSE, include=FALSE, results='asis'}

library(kableExtra)
library(stringr)

# Rename for joining
before <- strategy_percent_before %>%
  rename(strategy = strategy_Before, n_before = n, percent_before = percent)

after <- strategy_percent %>%
  rename(strategy = strategy_After, n_after = n, percent_after = percent)

# Merge strategies
strategy_summary <- full_join(before, after, by = c("strategy", "TIPO_FLOTA")) %>%
  arrange(desc(coalesce(percent_after, 0) + coalesce(percent_before, 0))) %>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

# Filter only IND
strategy_table_ind <- strategy_summary %>%
  filter(TIPO_FLOTA == "IND") %>%
  ungroup() %>%
  select(-TIPO_FLOTA)

rm(strategy_summary, strategy_percent, strategy_percent_before, after, before)

```


Table \@ref(tab:table-ind) for industrial strategy transitions.

```{r table-ind, echo=FALSE, results='asis'}

# Create table with grouped columns
strategy_table_ind %>%
  select(strategy, n_before, percent_before, n_after, percent_after) %>%
  kable("latex", booktabs = TRUE, digits = 1,
        col.names = c("Strategy", "n", "%", "n", "%"),
        caption = "Comparison of Strategies Before and After -- Industrial vessels") %>%
  add_header_above(c(" " = 1, "Before" = 2, "After" = 2)) %>%
  kable_styling(font_size = 10)

rm(strategy_table_ind)

```



```{r switchStrategy-ART, echo=FALSE, fig.height=6, fig.width=8, message=FALSE, warning=FALSE, results='asis', fig.cap='Strategy transitions for small-scale vessels'}

library(dplyr)
library(tidyr)
library(ggplot2)
library(ggalluvial)
library(patchwork)
library(ggrepel)

strategy_transitions <- full_join(strategy_before, strategy_after, by = c("COD_BARCO", "TIPO_FLOTA")) %>%
  mutate(
    strategy_Before = ifelse(is.na(strategy_Before), "No fishing", strategy_Before),
    strategy_After  = ifelse(is.na(strategy_After),  "No fishing", strategy_After)
  ) %>% filter(TIPO_FLOTA == "ART")

transition_counts <- strategy_transitions %>%
  count(strategy_Before, strategy_After)

simplify_focus_strategy <- function(strategy) {
  ifelse(strategy == "Only Sardine", "Sardine",
  ifelse(strategy == "Only Anchoveta", "Anchoveta",
  ifelse(strategy == "Only JackMackerel", "JackMackerel",
  ifelse(grepl("Sardine", strategy) & grepl("Anchoveta", strategy) & grepl("JackMackerel", strategy),
         "Sardine, JackMackerel & Anchoveta",
  ifelse(grepl("Sardine", strategy) & grepl("Anchoveta", strategy), "Sardine & Anchoveta",
  ifelse(grepl("Sardine", strategy) & grepl("JackMackerel", strategy), "Sardine & JackMackerel",
  ifelse(grepl("JackMackerel", strategy) & grepl("Anchoveta", strategy), "JackMackerel & Anchoveta",
  ifelse(strategy == "No fishing", "No fishing", "Other"))))))))
}

strategy_transitions_focus <- strategy_transitions %>%
  mutate(
    strategy_Before_simple = simplify_focus_strategy(strategy_Before),
    strategy_After_simple = simplify_focus_strategy(strategy_After)
  ) 

# %>%
#   filter(strategy_Before_simple %in% c("Sardine", "Anchoveta", "Sardine & Anchoveta", "Sardine & JackMackerel", "JackMackerel & Anchoveta", "Sardine, JackMackerel & Anchoveta","No fishing") |
#            strategy_After_simple %in% c("Sardine", "Anchoveta", "Sardine & Anchoveta", "Sardine & JackMackerel", "JackMackerel & Anchoveta", "Sardine, JackMackerel & Anchoveta","No fishing"))

transition_counts_focus <- strategy_transitions_focus %>%
  count(strategy_Before_simple, strategy_After_simple)


# Base plot
p <- ggplot(transition_counts_focus,
            aes(axis1 = strategy_Before_simple,
                axis2 = strategy_After_simple,
                y = n)) +
  geom_flow(aes(fill = strategy_Before_simple),
            width = 1/12, knot.pos = 0.4, alpha = 0.8) +
  geom_stratum(width = 1/12, fill = "gray95", color = "black") +
  scale_x_discrete(limits = c("2012–2018", "2019–2024"),
                   expand = c(.25, .25)) +
  scale_fill_brewer(type = "qual", palette = "Set2") +
  labs(x = "", y = "Number of Vessels") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.grid = element_blank()
  )

# Extraer posiciones de etiquetas
stratum_data <- ggplot_build(p)$data[[2]]

left_labels <- stratum_data %>%
  filter(x == 1) %>%
  mutate(x = x - 0.05,
         stratum = str_wrap(stratum, width = 20))

right_labels <- stratum_data %>%
  filter(x == 2) %>%
  mutate(x = x + 0.05,
         stratum = str_wrap(stratum, width = 20))

# Agregar etiquetas con repel
p <- p +
  geom_text_repel(
    data = left_labels,
    aes(x = x, y = y, label = stratum),
    inherit.aes = FALSE,
    hjust = 1, nudge_x = -0.1, direction = "y",
    size = 2.5, segment.color = "grey60", segment.size = 0.3
  ) +
  geom_text_repel(
    data = right_labels,
    aes(x = x, y = y, label = stratum),
    inherit.aes = FALSE,
    hjust = 0, nudge_x = 0.1, direction = "y",
    size = 2.5, segment.color = "grey60", segment.size = 0.3
  )

p

ggsave("figs/strategy_transitions_ART.svg", p,
       device = svglite::svglite, width = 10, height = 6, units = "in")

```

```{r switchStrategy-IND, echo=FALSE, fig.cap='Strategy transitions for industrial vessels', fig.height=6, fig.width=8, message=FALSE, warning=FALSE, results='asis'}

library(dplyr)
library(tidyr)
library(ggplot2)
library(ggalluvial)
library(patchwork)
library(ggrepel)

strategy_transitions <- full_join(strategy_before, strategy_after, by = c("COD_BARCO", "TIPO_FLOTA")) %>%
  mutate(
    strategy_Before = ifelse(is.na(strategy_Before), "No fishing", strategy_Before),
    strategy_After  = ifelse(is.na(strategy_After),  "No fishing", strategy_After)
  ) %>% filter(TIPO_FLOTA == "IND")

transition_counts <- strategy_transitions %>%
  count(strategy_Before, strategy_After)

simplify_focus_strategy <- function(strategy) {
  ifelse(strategy == "Only Sardine", "Sardine",
  ifelse(strategy == "Only Anchoveta", "Anchoveta",
  ifelse(strategy == "Only JackMackerel", "JackMackerel",
  ifelse(grepl("Sardine", strategy) & grepl("Anchoveta", strategy) & grepl("JackMackerel", strategy),
         "Sardine, JackMackerel & Anchoveta",
  ifelse(grepl("Sardine", strategy) & grepl("Anchoveta", strategy), "Sardine & Anchoveta",
  ifelse(grepl("Sardine", strategy) & grepl("JackMackerel", strategy), "Sardine & JackMackerel",
  ifelse(grepl("JackMackerel", strategy) & grepl("Anchoveta", strategy), "JackMackerel & Anchoveta",
  ifelse(strategy == "No fishing", "No fishing", "Other"))))))))
}

strategy_transitions_focus <- strategy_transitions %>%
  mutate(
    strategy_Before_simple = simplify_focus_strategy(strategy_Before),
    strategy_After_simple = simplify_focus_strategy(strategy_After)
  ) 

# %>%
#   filter(strategy_Before_simple %in% c("Sardine", "Anchoveta", "Sardine & Anchoveta", "Sardine & JackMackerel", "JackMackerel & Anchoveta", "Sardine, JackMackerel & Anchoveta","No fishing") |
#            strategy_After_simple %in% c("Sardine", "Anchoveta", "Sardine & Anchoveta", "Sardine & JackMackerel", "JackMackerel & Anchoveta", "Sardine, JackMackerel & Anchoveta","No fishing"))# %>%
#   filter(strategy_Before_simple %in% c("Sardine", "Anchoveta", "Sardine & Anchoveta", "Sardine & JackMackerel", "JackMackerel & Anchoveta", "Sardine, JackMackerel & Anchoveta","No fishing") |
#            strategy_After_simple %in% c("Sardine", "Anchoveta", "Sardine & Anchoveta", "Sardine & JackMackerel", "JackMackerel & Anchoveta", "Sardine, JackMackerel & Anchoveta","No fishing"))

transition_counts_focus <- strategy_transitions_focus %>%
  count(strategy_Before_simple, strategy_After_simple)


# Base plot
p <- ggplot(transition_counts_focus,
            aes(axis1 = strategy_Before_simple,
                axis2 = strategy_After_simple,
                y = n)) +
  geom_flow(aes(fill = strategy_Before_simple),
            width = 1/12, knot.pos = 0.4, alpha = 0.8) +
  geom_stratum(width = 1/12, fill = "gray95", color = "black") +
  scale_x_discrete(limits = c("2012–2018", "2019–2024"),
                   expand = c(.25, .25)) +
  scale_fill_brewer(type = "qual", palette = "Set2") +
  labs(x = "", y = "Number of Vessels") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.grid = element_blank()
  )

# Extraer posiciones de etiquetas
stratum_data <- ggplot_build(p)$data[[2]]

left_labels <- stratum_data %>%
  filter(x == 1) %>%
  mutate(x = x - 0.05,
         stratum = str_wrap(stratum, width = 20))

right_labels <- stratum_data %>%
  filter(x == 2) %>%
  mutate(x = x + 0.05,
         stratum = str_wrap(stratum, width = 20))

# Agregar etiquetas con repel
p <- p +
  geom_text_repel(
    data = left_labels,
    aes(x = x, y = y, label = stratum),
    inherit.aes = FALSE,
    hjust = 1, nudge_x = -0.1, direction = "y",
    size = 2.5, segment.color = "grey60", segment.size = 0.3
  ) +
  geom_text_repel(
    data = right_labels,
    aes(x = x, y = y, label = stratum),
    inherit.aes = FALSE,
    hjust = 0, nudge_x = 0.1, direction = "y",
    size = 2.5, segment.color = "grey60", segment.size = 0.3
  )

p

ggsave("figs/strategy_transitions_IND.svg", p,
       device = svglite::svglite, width = 10, height = 6, units = "in")

rm(left_labels, right_labels, strategy_after, strategy_before, strategy_transitions, strategy_transitions_focus, stratum_data, transition_counts, transition_counts_focus, p)

rm(get_strategy, simplify_focus_strategy)

```

## Fishing seasons

```{r harvest_data, message=FALSE, warning=FALSE, include=FALSE}

library(readxl)
library(dplyr)

harvest_IFOP_month <- readRDS("data/harvest/IFOP_month.rds") 
harvest_IFOP <- readRDS("data/harvest/IFOP.rds")
harvest_SERNAPESCA <- readRDS("data/harvest/sernapesca.rds")
harvest_SERNAPESCA_v2 <- readRDS("data/harvest/sernapesca_v2.rds")


#---- IFOP loogbook data ----

harvest_IFOP_logbooks <- logbooks %>% 
  filter(year >= 2012) %>%
  filter(REGION %in% c(5,6,7,8,9,10,14,16)) %>%
  group_by(NOMBRE_ESPECIE, year) %>%
  summarise(
    harvest_IFOP_loogbook_centrosur = sum(CAPTURA_RETENIDA, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  rename(specie = NOMBRE_ESPECIE) %>%
  mutate(harvest_IFOP_loogbook_centrosur = harvest_IFOP_loogbook_centrosur / 1000)


#---- Harvest data merged ---- 
harvest <- left_join(harvest_SERNAPESCA_v2, harvest_SERNAPESCA, by = c("year", "specie")) 
harvest <- left_join(harvest, harvest_IFOP, by = c("year", "specie"))
harvest <- left_join(harvest, harvest_IFOP_logbooks, by = c("year", "specie")) 

rm(harvest_IFOP, harvest_SERNAPESCA, harvest_SERNAPESCA_v2, harvest_IFOP_logbooks)

```

```{r season_SPF, eval=FALSE, fig.align='center', fig.cap="Fishery seasons", fig.pos='ht!', message=FALSE, warning=FALSE, include=FALSE}

months <- c("E", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D")

# Fishing periods
sardine_periods <- data.frame(
  start = c(1, 3,7,10),    
  end = c(3,7,10,12),      
  status = c("Closed", "Open", "Closed", "Open")
)

# Cruise brackets
cruise_brackets <- data.frame(
  cruise = c("1st Cruise (Summer)", "2nd Cruise (Fall)"),
  x_start = c(1, 3),   
  x_end = c(3, 6),      
  y = c(1.05, 1.05)       
)

# Plot with reduced white space
ggplot() +
  geom_segment(data = sardine_periods, aes(x = start, xend = end, y = 1, yend = 1, color = status), size = 4) +
  geom_segment(data = cruise_brackets, aes(x = x_start, xend = x_end, y = y, yend = y), size = 1) +
  geom_segment(data = cruise_brackets, aes(x = x_start, xend = x_start, y = y, yend = y - 0.05), size = 1) +
  geom_segment(data = cruise_brackets, aes(x = x_end, xend = x_end, y = y, yend = y - 0.05), size = 1) +
  geom_text(data = cruise_brackets, aes(x = (x_start + x_end) / 2, y = y + 0.05, label = cruise), size = 3.5) +
  scale_x_continuous(breaks = 1:12, labels = months, expand = c(0, 0)) +
  scale_y_continuous(limits = c(0.9, 1.3), breaks = 1, labels = " ", expand = c(0, 0)) +
  scale_color_manual(values = c("Open" = "green", "Closed" = "red")) +
    labs(x = "Month", y = NULL) +
  theme_minimal() + coord_cartesian(clip = "off") +
theme(
  plot.margin = margin(2, 2, 2, 2),  # very small margins (t, r, b, l)
  axis.title.x = element_text(margin = margin(t = 0)),  # reduce margin below x-axis label
  axis.text.x = element_text(margin = margin(t = 0)),   # reduce space below x-axis ticks
  plot.title = element_text(hjust = 0.5, margin = margin(b = 0))  # reduce margin above title
)

rm(cruise_brackets, sardine_periods)

```

```{r monthlyharvest, echo=FALSE, fig.align='center', fig.cap="Average monthly landings by species (2012-2024; South-Central Chile) ", fig.pos='ht!', message=FALSE, warning=FALSE}

library(dplyr)
library(stringr)

# Diccionario
month_translation <- c(
  "ene" = "jan", "feb" = "feb", "mar" = "mar", "abr" = "apr",
  "may" = "may", "jun" = "jun", "jul" = "jul", "ago" = "aug", 
  "sept" = "sep", "oct" = "oct", "nov" = "nov", 
  "dic" = "dec"
)

# Orden destino
month_order <- c("jan","feb","mar","apr","may","jun","jul","aug","sep","oct","nov","dec")

harvest_summary <- harvest_IFOP_month %>%
  mutate(
    month = str_to_lower(str_trim(month)),                     
    month = stringi::stri_trans_general(month, "Latin-ASCII"), 
    month = dplyr::recode(month, !!!month_translation, .default = month),
    month = factor(month, levels = month_order, ordered = TRUE)
  ) %>%  # Translate to English
  group_by(month, specie) %>%  # Make sure your column is 'species'
  summarise(harvest = mean(total_harvest_IFOP_centrosur, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(month = factor(month, levels = month_order)) %>%
  mutate(specie = dplyr::recode(specie,
                         "ANCHOVETA" = "Anchoveta",
                         "SARDINA COMUN" = "Sardine",
                         "JUREL" = "Jack Mackerel")) 


# Plot with facets by species
ggplot(harvest_summary, aes(x = month, y = harvest, fill = specie)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ specie, scales = "free_y", ncol = 1) +
  labs(x = "Month", y = "Average Harvest (tons)", fill = "Specie") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Paired")

rm(harvest_IFOP_month, harvest_summary, month_order, month_translation, months)


```


```{r monthlyharvestSP, eval=FALSE, fig.align='center', fig.cap="Average monthly landings by species (2012-2024; South-Central Chile) ", fig.pos='ht!', message=FALSE, warning=FALSE, include=FALSE}
library(stringi)

harvest_month_SERNAPESCA <- read.csv(
  file.path(dirdata, "SERNAPESCA", "bd_desembarque.csv"),
  fileEncoding = "Latin1",
  sep = ";") %>%
  rename(specie = especie,
         year = año,
         month = mes) %>%
  mutate(zone = case_when(
    region %in% c("Valparaíso", "Metropolitana" , "O'Higgins", "Maule", "Ñuble", "Bio-bío", "La Araucanía", "Los Ríos", "Los Lagos") ~ "Centro_Sur", 
    region %in% c("Arica y Parinacota", "Tarapacá", "Antofagasta", "Atacama", "Coquimbo" ) ~ "Norte",
    region %in% c("Magallanes", "Aysén") ~ "Extremo_Sur",
    TRUE ~ "No_Especifica")) %>% 
  mutate(specie = toupper(stri_trans_general(specie, "Latin-ASCII"))) %>% 
  filter(specie %in% c("ANCHOVETA", "JUREL", "SARDINA COMUN")) %>% 
  filter(zone %in% c("Centro_Sur"))

# %>%
#   pivot_wider(
#     names_from = zone,
#     values_from = total_harvest_SERNAPESCA_v2,
#     names_glue ="
#   ) 

# English month order
month_order <- c("jan", "feb", "mar", "apr", "may", "jun",
                 "jul", "aug", "sep", "oct", "nov", "dec")


harvest_summary2 <- harvest_month_SERNAPESCA %>%
  mutate(
    month = factor(as.integer(month), levels = 1:12, labels = month_order, ordered = TRUE)
  ) %>%
  group_by(month, year, specie) %>%
  summarise(harvestym = sum(toneladas, na.rm = TRUE), .groups = "drop") %>%
  filter(year >= 2012) %>%
  group_by(month, specie) %>%
  summarise(harvest = mean(harvestym, na.rm = TRUE), .groups = "drop") %>%
  arrange(month) %>%
  mutate(month = factor(month, levels = month_order)) %>%
  mutate(specie = dplyr::recode(specie,
                         "ANCHOVETA" = "Anchoveta",
                         "SARDINA COMUN" = "Sardine",
                         "JUREL" = "Jack Mackerel")) 


# Plot with facets by species
ggplot(harvest_summary2, aes(x = month, y = harvest, fill = specie)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ specie, scales = "free_y", ncol = 1) +
  labs(x = "Month", y = "Average Harvest (tons)", fill = "Specie") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Paired")

rm(harvest_month_SERNAPESCA, harvest_summary2, month_order)

```



# Data and methodology


<!-- SEE HOW FAR VESSEL TRAVEL -- 80 longitud aprox... IM USING 165 nm -->
<!-- USE DIFFERENT LENGHT EQUATIONS? (INTEAD OF GEAR?) -->
<!-- Shift in spatial distribution of jack mackerel out of EEZ during El Niño events -->


To fulfill the research's objectives, and following @Kasperski2015-jm, the research entails five different stages: (i) estimating the annual stock dynamics of each species included in the model, (ii) estimating trip level cost functions, (iii) estimating total annual trips, (iv) estimate the inverse demand model for outputs (i.e., price responses to supply), and (v) conduct numerical optimization to examine how harvest and profits levels evolve over time. The numeral optimization uses estimated parameters from the previous four stages to conduct the optimization procedure.


## Historical data

We use data requested from the Chilean Fisheries Development Institute (Instituto de Fomento Pesquero, IFOP) covering the 2012–2024 period. The dataset includes trip-level microdata, which contain detailed records on vessel identifiers, departure and arrival times, vessel capacity, fleet and gear type, ports of departure and landing, fishery codes, haul timing and location, species composition, retained catch, and trip activity. In addition, we requested annual information on stock abundance and vessel landings by port, county, region, country, and species.  Finally, we use ex-vessel prices, reported monthly or annually by port, county, region, country, and species. These prices reflect those paid by processing plants to fishers at the point of first sale and are obtained through IFOP’s landing surveys, which do not necessarily cover all market transactions.

<!-- How different are SERNAPESCA and IFOP harvest data? (Figura -->
<!-- \@ref(fig:harvestsource)) -->

```{r harvestsource, eval=FALSE, fig.align='center', fig.cap="Desembarques anuales (IFOP vs SERNAPESCA)", fig.pos='ht!', message=FALSE, warning=FALSE, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)

harvest_long <- harvest %>% 
  dplyr::select(c("specie", "year", 
                  "total_harvest_IFOP_centrosur",
                  "harvest_LANCHAS_IFOP_centrosur",
                  "harvest_BOTES_IFOP_centrosur",
                  "harvest_IND_IFOP_centrosur",
                  "total_harvest_all_SERNAPESCA_Centro_Sur",
                  "total_harvest_sernapesca_v2_centro_sur",
                  "harvest_IFOP_loogbook_centrosur")) %>%
  pivot_longer(cols = c("total_harvest_IFOP_centrosur",
                        "harvest_LANCHAS_IFOP_centrosur",
                        "harvest_BOTES_IFOP_centrosur",
                        "harvest_IND_IFOP_centrosur",
                        "total_harvest_all_SERNAPESCA_Centro_Sur",
                        "total_harvest_sernapesca_v2_centro_sur",
                        "harvest_IFOP_loogbook_centrosur"),
               names_to = "source", 
               values_to = "harvest") %>%
  mutate(source = dplyr::recode(source,
                         "harvest_LANCHAS_IFOP_centrosur" = "Lanchas IFOP",
                         "harvest_BOTES_IFOP_centrosur" = "Botes IFOP",
                         "harvest_IND_IFOP_centrosur" = "Industrial IFOP",
                         "total_harvest_IFOP_centrosur" = "Total IFOP",
                         "total_harvest_all_SERNAPESCA_Centro_Sur" = "Total SERNAPESCA",
                         "total_harvest_sernapesca_v2_centro_sur" = "Total SERNAPESCA v2",
                         "harvest_IFOP_loogbook_centrosur" = "Total IFOP (logbooks)")) %>%
  filter(source %in% c("Total IFOP", "Total SERNAPESCA v2", "Total IFOP (logbooks)")) %>% filter(year >= 2012)

# Gráfico
ggplot(harvest_long, aes(x = year, y = harvest, color = source)) +
  geom_line(size = 1) +
  facet_wrap(~specie, scales = "free_y") +
  labs(x = "Anio", 
       y = "Annual landings (tons)", 
       color = "Source") +
  theme_minimal() +
  theme(legend.position = "bottom") 

rm(harvest_long)

```

For the environmental covariates, we use data from the E.U. Copernicus Marine Service Information, accessed through the Copernicus Marine Toolbox API. Salinity, sea surface temperature, and current speed and direction were obtained from the Global Ocean Physics Reanalysis (GLORYS12V1), which provides data at a 1/12° horizontal resolution with 50 vertical levels [@GLORYS12V1]. Wind speed and direction at the surface were obtained from the Global Ocean Hourly Reprocessed Sea Surface Wind and Stress from Scatterometer and Model dataset, available at 0.125° horizontal spatial resolution and hourly frequency [@WIND_GLO_PHY]. Chlorophyll-a concentrations were obtained from the Global Ocean Colour dataset, which provides data at ~4 km horizontal resolution [@GlobColour]. All environmental data were retrieved daily (hourly in the case of winds) for the 2012–2024 period, covering the Chilean Exclusive Economic Zone (EEZ) between 32°S and 41°S (Figure \@ref(fig:figEnvData))
 

```{r figEnvData, echo=FALSE, fig.align='center', fig.cap="Geographical extent of the study area used for environmental covariates limited to the Chilean Exclusive Economic Zone", message=FALSE, warning=FALSE, fig.pos='ht!', out.width='80%'}

knitr::include_graphics("figs/env_data_map.pdf")

```
 

### To be requested

- Diesel cost.
- Permits by vessels
- Quota prices?
  - Auction market but also secondary market if available
    - If no data, maybe intrapolate prices from other auctions?
  - Captures elements of forward-looking behavior and information [@Birkenbach2024]. @reimer2022structural similarly argue that including a quota price captures forward looking behavior and allows one to simplify the dynamic model to a static one.
- Quota by area/fishing organization for Artisanal, and TAC for industrial with ITQ (by vessel?)

  
## Future data for projections

- OracleBio 
  - Unfurtunally, only decadal (e.g., 2040–2050) projections for different scenarios for SST, salinity, currents and chlorophyll (4km resolution)
  - No winds; CMIP6 for winds? (~100 km).
        
  
## Econometrics models

### Stock dynamics

To estimate stock dynamics, we use annual data on species-specific biomass (stock abundance) and vessel landings within Chile’s Exclusive Economic Zone (EEZ). Following @Kasperski2015-jm, the baseline model for the interannual growth of each species $i$ is represented by a discrete logistic function that accounts for intra- and inter-species interactions:

\begin{equation}
x_{i,y+1} + h_{iy} = \underbrace{(1 + r_i)x_{iy} + \eta_i x_{iy}^2}_{R_i(x_{iy})} + \underbrace{\sum_{j \neq i}^{n-1} a_{ij} x_{iy} x_{jy}}_{I_i(x_y)} \quad i=1,\ldots,n
\label{eq1}
\end{equation}
where $x_{iy}$ is the fish stock by species $i=1,\ldots,n$ in year $y$, $n$ is the total number of species, $h_{iy}$ is the annual harvest of species $i$ on year $y$, $r_i$ is the intrinsic growth rate of the resource $i$, $\eta_i$ is a density-dependent factor related to the carrying capacity, and $\alpha_{ij}$ are the interaction parameters between species. The system of $n$ growth equations is estimated simultaneously using seemingly unrelated regression (SUR). Following @Richter2018, we augment \eqref{eq1} by including environmental covariates $Env_{iy}$ that affect the fish stock and an error term $\varepsilon_{iy}$ that captures random recruitment: 
\begin{equation}
x_{i,y+1} + h_{iy} = \underbrace{(1 + r_i)x_{iy} + \eta_i x_{iy}^2}_{R_i(x_{iy})} + \underbrace{\sum_{j \neq i}^{n-1} a_{ij} x_{iy} x_{jy}}_{I_i(x_y)} + \rho_i Env_{iy} + \varepsilon_{iy} \quad i=1,\ldots,n
\label{eq2}
\end{equation} 
where $\rho_i$ are the coefficient for the environmental covariates.

Environmental conditions were summarized annually using sea surface temperature (SST) and chlorophyll-a concentration (CHL), both averaged over the South-Central Chile region within the EEZ. These variables were selected due to their recognized influence on small pelagic productivity and spatial distribution [@Axbard2016]. To capture potential nonlinear responses, both linear and quadratic terms for SST and CHL were included. Early specifications also tested linear and quadratic wind effects, but excluding them did not significantly reduce explanatory power ($\text{F}_{6,24} = 0.36$, $p\text{-value} = 0.89$). SST serves as a proxy for large-scale oceanographic regimes (warm vs. cold phases) that shape recruitment success and species dominance along the Humboldt Current System. Anchovy tends to dominate during cold, nutrient-rich phases, while sardine prevail under warmer conditions [@cahuin2009; @Yáñez2014]. CHL captures interannual variations in primary productivity, which are critical for sustaining the planktonic base of pelagic food webs. As emphasized by @sumaila2011, “Changes in primary productivity and planktonic community structure affect the amount of energy transferred to higher trophic levels and, eventually, the productivity of trophic groups that contribute to fisheries catches” [see also @jennings2008; @cheung2008]. Thus, chlorophyll acts as an indicator of ecosystem energy available to forage fish.

The SUR framework is particularly suited to this multi-species context, as it allows for contemporaneous correlation among residuals that arise from shared environmental shocks and trophic linkages. Cross-species interaction terms are specified between predator (jack mackerel) and prey (anchoveta, sardine) pairs, enabling the model to capture both biological coupling and indirect effects of environmental variability transmitted across species.


- To add here:
  - Some species might be more resistant to changes in SST and CHL
  - Jack Mackerel biomass why doifferent than Anchoveta and Sardine biomass..
  - Sardine and Anchoveta they move together
  - BE CLEAR THAT EACH SPECIES HAVE ITS OWN EQUATION




### Trip level cost functions


  <!-- - 80 and 300 m$^3$ holds sardine and anchovy; More than 800 m$^3$, likely jack mackerel. @Pena-Torres2017-gn dummy for $H \leq 370 \ \text{m}^3$, $370 \ \text{m}^3 < H \leq 790 \ \text{m}^3$, $H > 790 \ \text{m}^3$ ... Also use age of vessel to check technical obsolescence (year trip, year ship was built) -->
  
 - @sumaila2011: "Capital costs, that is, the cost of vessels, fishing gear, processing plants and so on, would be affected by climate change if additional capital for fishing and process ing operations is required to adapt to climate change impacts on the quantity, composition and distribution of fisheries resources65. Changes in migratory routes and fish distribution would affect travel time, which can lead to increases or decreases in fuel and ice cost depending on catch levels and patterns, and the management regime in place."



Ignoring trip subscript, the cost functions vary by vessel
$v=1,\ldots,V_g$ and gear used $g=1,\ldots,G$, where $V_g$ is the number
of observations using gear type $g$, and $G$ is the total number of
available (or observed) gears: \begin{equation}
C_{vg} = \sum_{i=1}^{2n+M+k} \alpha_{g, \mathbf{X}_i} \mathbf{X}_{ivg} + \frac{1}{2} \sum_{i=1}^{2n+M+k} \sum_{j=1}^{2n+M+k} \alpha_{g, \mathbf{X}_i\mathbf{X}_j} \mathbf{X}_{ivg} \mathbf{X}_{jvg}
\label{eq3}
\end{equation} where $C_{vg}= w z_{vg}^*$,
$\mathbf{X}_{vg}=[w;h_{vg};x;Z_v]$, $w$ is a $V_g \times M$ matrix of
variable input prices, $h_{vg}$ is an $V_g \times n$ matrix of harvest
quantities, $x$ is an $V_g \times n$ matrix of given stock levels of the
species of interest, and $Z_v$ is an $V_g \times k$ matrix of given
vessel characteristics. Therefore, $\mathbf{X}_{vg}$ is a
$V_g \times (2n+M+k)$ matrix, and $\mathbf{X}_{ivg}$ represents the
*i*th column of the $\mathbf{X}_{vg}$ matrix.

Together with estimating the restricted cost function, we estimate the
conditional input demand equations. This addition allows an increase in
the degrees of freedom by imposing cross-equation parameter constraints
and allows for the testing of, for instance, jointness in inputs
[@Kasperski2015-jm]. The conditional input demand equations are derived
by Shepard's Lemma: \begin{equation}
\frac{\partial C_{vg}}{\partial w_m} = z^*_{vg,w_m} = \alpha_{g,w_m} + \sum_{j=1}^{2n+M+k} \alpha_{g,w_m,\mathbf{X}_j} \mathbf{X}_{jvg} \quad m=1,\ldots,M.
\label{eq4}
\end{equation}

Similar to stock dynamics, the system of equations formed by \eqref{eq3}
and \eqref{eq4} can be estimated using SUR. To comply with economic
theory, and to reduce even more the number of parameters to estimate,
the following restrictions are imposed when estimating \eqref{eq4}:

1.  Symmetry of the cost function, where \begin{equation*}
      \alpha_{g,\mathbf{X}_i\mathbf{X}_j} = \alpha_{g,\mathbf{X}_j\mathbf{X}_i} \quad \forall \quad i=1,\ldots,(2n + M + k); \ i \neq j; \ g = 1,\ldots, G.
      \end{equation*}

2.  Linear homogeneity in input prices, where \begin{equation*}
      \sum_m^M \alpha_{g,w_m} = 1 \ \text{and} \ \sum_m^M \alpha_{g,w_m,\mathbf{X}_i} = 0 \quad i=1,\ldots,(2n + M + k); \ g = 1,\ldots, G.
      \end{equation*}

Data at the trip level is available upon request from the Chilean Fisheries Research Institute (IFOP), which registers geo-referenced
catch information on the Chilean fleet's fishing operation per trip [see e.g. @Pena-Torres2017-gn and [IFOP data
observatory](ifop.dataobservatory.net)]. As inputs we can use the time
spent at sea during a trip, where the price is the average wage pay to
crew member per hour, and the distance traveled during a trip, where the
price of distance traveled is the diesel cost. Therefore, the total cost
function $C_{vg}= w z_{vg}^*$ for vessel $v$, using gear $g$ in a trip
would be sum of the total cost of distance travelled plus the total cost
of the time spent at sea.

**Note:** *Depending on the type of vessel, this cost should change.
Some vessels are more efficient, other one are more heavy. How to
capture this? The righ hand side has vessel characteristics, so the
effect of harvest would be conditional on vessel characteristics, the
stock levels and input prices. As we only care in the margin how harvest
increase cost, this should be fine. @Kasperski2015-jm mention this
"...no reliable fixed cost information on these vessels exists, but
these should not affect the optimization as economic decisions are made
at the margin. Therefore, this study does not measure true profit, but
rather a proxy based on the net operating rent accruing to vessels in
the fishery."*

To link this function to climate change, we can also include additional
environmental variables $Env$ to $\mathbf{X}_{vg}$ such as wind
intensity and wave conditions in each trip at the harvest location, upon
data availability. Therefore, the augmented $X_{vg}$ matrix becomes
$\mathbf{X}^{'}_{vg} = [w;h_{vg};x;Z_v;Env]$.

MAYBE INCLUDE QUOTA PRICE? **Higher quota prices for depleted stocks
(e.g., GOM cod) reduce incentives to target them.** and **Active leasing
markets for quota (and previously for DAS) allow fishermen to treat
quota as a "priced input" rather than a fixed, exhaustible resource.**


Add
  - liters/hours x hours trip x fuel price
  - crew member x harvest x price x share
  - certification cost per landed ton x landing
  


### Total annual trips

The number of trips a vessel will take in a given year for each gear
type used is assumed to follow a Poison distribution
[@Kasperski2015-jm]: \begin{equation}
Pr\left[T^{*}_{vgy} = t_v\right] = \frac{exp^{-exp(U^{'}_{vg}\beta_g)}exp(U^{'}_{vg}\beta_g)^{t_v}}{t_v !}
\label{eq5}
\end{equation} where $U_{vg}=[p,w,h_{vg},\bar{q},Z_{vg}]$ is a
$(3n+M+k+1)×V_g$ matrix of explanatory variables, $\beta_g$ is a
$(3n+M+k+1)\times1$ matrix of coefficients to be estimated, $t_v$ is the
number of trips taken by vessel $v$ using gear type $g$ in year $y$, and
$\bar{q}$ is the annual quota level. Additionally, we can add the
accumulation of "bad weather days" as an explanatory variable to
incorporate weather conditions into this decision, thus
$U_{vg}=[p,w,h_{vg},\bar{q},Z_{vg}, Env]$


De Dresdner 2014: 
- Regulacion: Numero de dias de vedas / Cuota anual
- Precios: Precio proceso + Precio Diesel (promedio meses de mayor productividad en la pesquería (marzo a junio y octubre a diciembre).
- Agregar Precio quota?


### Inverse demand model 

The price of each species is modeled using an Inverse Almost Ideal Demand System (IADS). The price of a species $i$ in year $y$ is the following:

\begin{equation}
\ln p_{iy} = \sum_j^n \gamma_j \ln h_{j,y} + \gamma_{H} \ln H + \gamma_{FM} \ln P^\text{FishMeal} + \epsilon_{iy}, \quad i = 1,\ldots,n, \ j = 1,\ldots,n.
\label{eq6}
\end{equation}

Note that harvest may be endogenous in this system due to simultaneity. For this reason, the system formed by \eqref{eq6} will be estimated using a Three Stage Least Squares (3SLS) procedure, where h_j{j,y} would be instrumented by variables that affect the fishers' supply function such as SST, Chl, and fuel prices.

<!-- Comment: One concern I have regards the endogeneity of factors determining the outcomes of fishing effort on a fishing
trip (where, when, how long, and for what, and how much), climate conditions (temperature, wave, and wind),
and input and output prices. While Kasperski assumes that total allowable catches are binding to avoid one potential endogeneity, I wonder if this is a good assumption under climate change. What would you need to assume about the regulator’s response to the changing conditions and the setting of the total allowable catches? -->


<!-- No era que el paper estimaba un IAIDS. Sino que al precio ex-vessel a ser un % del FOB justifica que sea una variable exógena para la IAIDS que nosotros estimaremos. Así capturar las desviaciones del mercado internacional se transmitan directamente al precio ex-vessel. El otro aspecto era el siguiente: En las estimaciones MCO para la VIII Región, el Precio de la Harina de Pescado * Dólar presenta un signo negativo y significativo sobre el número agregado de viajes anuales, aunque en la XIV Región el signo es positivo [11 (254, 398). Lo que podría implicar que la reaccion de la flota al FOB es heterogénea por región  
R:  Perfecto. Entonces no estimaron un modelo de precios. Mejor aún, porque entonces la contribución de su trabajo sigue siendo alta. Solo usan FOB como exógena, por lo tanto, nuestra metodología sigue siendo la misma, de estimar sin incluir FOB, luego incluyéndolo, ver qué pasa con el coeficiente de Q, para ver si hay algo más en la determinación de precios a corto plazo. Luego, si hay tiempo, ver si de hecho el precio FOB depende de desembarques chilenos. En cuanto a la heterogeneidad, a mí me da la impresión de que ese texto de hecho está indicando esta endogeneidad que estamos tratando de capturar! Es posible que la XIV no tenga efecto en el FOB, por lo que lo que vemos es un efecto insesgado para esa región, pero que en la VIII, que es una productora grande de pequeños pelágicos, sí tenga un efecto en el precio FOB, por lo tanto, el coeficiente negativo que obtienen de hecho está sesgado por endogeneidad!! Yo usaría ese texto como justificación de su estudio... -->

- DIFERENCIAR ENTRE CONSUMO HUMANO DIRECTO e INDIRECTO???


## Numerical optimization

To obtain the effect of future climate variability on stock, harvest, quota and profits, we conduct numerical optimization for different climate scenarios using the parameters estimates for the stock dynamic, cost functions, total annual trips and inverse demand equations. In each year, a vessel maximizes profits by choosing their optimal number of trips $T_g$ and harvest levels per trip $h_{g\tau}$ given a gear type:
\begin{align}
\max_{h_{gt}, T_g} \quad \pi_{vgt} & = \sum_{\tau=t}^{T_g} \rho^\tau \left\{ P(h) h_{g\tau} - C_g(h_{g\tau} | w, x, Z, Env) \right\} \quad \tau = t,\ldots, T_g \nonumber \\
\textbf{s.t} \quad q_{g,t+1} & = \omega \ast \bar{q} - \sum_{t=1}^{t} h_{gt} \geq 0, \quad t = 1, \dots, T-1, \quad g = 1, \dots, G
\label{eq7}
\end{align}
where $\rho$ is the intra-annual discount factor, $\omega$ is a vector of shares of $\bar{q}$, and $h_{lt}=0$ for all $l\neq g$. The vector of shares is obtained from historical data on harvest. The optimal profit from the maximization problem in \eqref{eq7} is denoted as $\pi_{vgy}^* (p,w,x,Z,\bar{q},\omega, Env)$, and $h_{vgty}^*$ and $T_{vgy}^*$ are the optimal choices harvest per trip and total number of trips in year $y$ for vessel $v$. To obtain the optimal quota level, we
must solve the social-planner optimization problem to maximize the net value of the fishery by choosing the quota levels per year and by species.

Following @Kasperski2015-jm, the optimization problem will be conducted for the next 25 years. I will use different climate scenarios and compare different optimal outcomes between them by using future projections for the environmental variables included in the model.


- Note that the optimal quota would that we would find using historical data might differ from the actual quota set by the regulator as they model is for single species.


## Projections

*From [@Yáñez2014]*:

-   "To project the model, the average structure of catches and temperature (of Antofagasta and the region Niño 3 + 4) for the years 2005, 2006 and 2007 were used as starting point. We consider a linear increase in temperature, taking into account four climate change scenarios based on the scenarios presented by IPCC, designed for the northern part of Chile until 2100.
    1.  The first scenario considers an increase in temperature of 0.034ºC per year [@Fuenzalida2007], similar to that estimated by @Trenberth2007.
    2.  A second scenario, more moderate, of 0,025ºC/year is also proposed by @Fuenzalida2007.
    3.  The third scenario is not considered a significant effect on the area, following the work of @Trenberth2007.
    4.  The fourth scenario is contradictory, indicating a cooling de 0.02 ºC/year [@falvey2009]. It should be noted that according to the work of @Fuenzalida2007 and @falvey2009 the same SST increase (or decrease) were considered for both temperatures (in Antofagasta and in the Niño 3 + 4 region)."

# Results

## Stock biomass model


As shown in Figure \@ref(fig:biomass), biomass levels vary by species, and there is some interrelation between them. It is also clear that these biomass levels are affected by the harvests that occurred during those periods. For instance, in the case of jack mackerel, an abrupt decline in biomass is observed, likely due to a combination of
overexploitation of the resource and unfavorable environmental conditions.
        
```{r biomass, echo=FALSE, fig.align='center', fig.cap="Estimated biomass of small pelagic species in Chile (2000–2024)", fig.pos='ht!', message=FALSE, warning=FALSE}

library(ggplot2)
library(tidyr)
library(scales)
library(viridis)

biomass_long <- biomass %>%
  pivot_longer(cols = c(sardine_biomass, anchoveta_biomass, jurel_biomass_cs, jurel_biomass_cs_intra),
               names_to = "species", 
               values_to = "biomass") %>%
  filter(!is.na(biomass)) %>%
  arrange(species, year)

biomass_long$biomass <- as.numeric(biomass_long$biomass)
biomass_long$year <- as.numeric(biomass_long$year)
biomass_long$species <- dplyr::recode(biomass_long$species,
                               sardine_biomass = "Sardine", 
                               anchoveta_biomass = "Anchoveta", 
                               jurel_biomass_cs = "Jack Mackerel", 
                               jurel_biomass_cs_intra = "Jack Mackerel (intrapolated)")

biomass <- biomass_long %>% filter(species %in% c("Sardine", "Anchoveta", "Jack Mackerel", "Jack Mackerel (intrapolated)"))


# Plot
ggplot(biomass, aes(x = year, y = biomass, color = species, group = species)) +
  geom_smooth(se = TRUE, method = "loess", span = 0.4, linetype = "solid") +
  scale_y_continuous(labels = scales::comma) +  # Or use scale_y_log10(labels = comma) to compress
  labs(x = "Year",
       y = "Biomass (tons)",
       color = "Species") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.y = element_text(angle = 0, hjust = 1),
    legend.position = "right" 
  ) + 
  scale_color_viridis_d(option = "D")


rm(list = c("biomass_long"))

```

Figure \@ref(fig:harvestbiomass) show biomass adding harvest

```{r harvestbiomass, echo=FALSE, fig.align='center', fig.cap="Estimated biomass vs harvest of small pelagic species in CentralSouth Chile (2000–2024)", fig.pos='ht!', message=FALSE, warning=FALSE}

library(tidyverse)

harvest2 <- harvest %>%
  select(c(specie, year, total_harvest_sernapesca_v2_centro_sur)) %>%
  mutate(specie = dplyr::recode(specie,
                          "SARDINA COMUN" = "Sardine",
                          "JUREL" = "Jack Mackerel",
                          "ANCHOVETA" = "Anchoveta"
                          )) %>%
  rename(harvest = total_harvest_sernapesca_v2_centro_sur) %>%
  rename(species = specie)

harvest2 <- bind_rows(
  harvest2,
  harvest2 %>%
    filter(species == "Jack Mackerel") %>%
    mutate(species = "Jack Mackerel (intrapolated)")
)

biomass_harvest <- full_join(biomass, harvest2, by = c("species" = "species", "year" = "year")) %>% arrange(species, year)

biomass_harvest_long <- biomass_harvest %>%
  pivot_longer(cols = c(biomass, harvest),
               names_to = "variable",
               values_to = "value")

ggplot(biomass_harvest_long, aes(x = year, y = value, color = variable)) +
  geom_line(size = 1, na.rm = TRUE) +  
  # geom_point(size = 2, na.rm = TRUE) + 
  # geom_smooth(se = FALSE, span = 0.4, method = "loess", na.rm = TRUE) + # <- smoothing
  facet_wrap(~species, scales = "free_y") +
  labs(x = "Year", y = "Value (tons)", color = "Variable") +
  scale_color_manual(values = c("biomass" = "darkgreen", "harvest" = "steelblue")) +
  theme_minimal() +
  theme(legend.position = "top")

rm(biomass_harvest_long, harvest2)

```



<!-- ESTIMACIONES DE BIOMASA -->



```{r database_for_biomass_est, message=FALSE, warning=FALSE, include=FALSE}

library(tidyverse)
library(janitor)

# --- 1. Limpieza y preparación de datos ----
biomass_harvest_wide <- biomass_harvest %>%
  mutate(species = gsub("[^A-Za-z0-9]", "", tolower(species))) %>%
  arrange(species, year) %>%
  group_by(species) %>%
  mutate(biomass_t1 = lead(biomass)) %>%
  ungroup() %>%
  pivot_wider(
    names_from = species,
    values_from = c(biomass, harvest, biomass_t1),
    names_sep = "_"
  ) %>%
  janitor::clean_names()

# Biomasa total sin cosecha
biomass_harvest_wide <- biomass_harvest_wide %>%
  mutate(
    biomass_t1_sardine_noharvest      = biomass_t1_sardine + harvest_sardine,
    biomass_t1_anchoveta_noharvest    = biomass_t1_anchoveta + harvest_anchoveta,
    biomass_t1_jackmackerel_noharvest = biomass_t1_jackmackerelintrapolated + harvest_jackmackerelintrapolated
  )

# Obtener datos ambientales 

env_dt_year1 <- env_dt %>% 
  mutate(year = year(date)) %>%
  group_by(year) %>% 
  summarize(sst = mean(sst, na.rm = TRUE),
            chl = mean(chl, na.rm = TRUE),
            wind = mean(speed_max, na.rm = TRUE))

env_dt_year2 <- env_dt_00_11 %>% 
  mutate(year = year(date)) %>%
  group_by(year) %>% 
  summarize(sst = mean(sst, na.rm = TRUE),
            chl = mean(chl, na.rm = TRUE),
            wind = mean(speed_max, na.rm = TRUE))

env_dt_year <- rbind(env_dt_year2, env_dt_year1)
biomass_harvest <- left_join(biomass_harvest_wide, env_dt_year, by = "year")

rm(env_dt_year, env_dt_year2, env_dt_year1, env_dt_00_11, biomass_harvest_wide)

```


```{r sur_data, message=FALSE, warning=FALSE, include=FALSE}

library(tidyverse)

sur_df <- biomass_harvest %>%
  select(
    year,
    biomass_t1_sardine_noharvest,
    biomass_t1_anchoveta_noharvest,
    biomass_t1_jackmackerel_noharvest,
    biomass_sardine,
    biomass_anchoveta,
    biomass_jackmackerelintrapolated,
    sst, chl, wind
  ) %>%
  drop_na() %>%
  mutate(
    sst_c  = sst - mean(sst, na.rm = TRUE),
    chl_c  = chl - mean(chl, na.rm = TRUE),
    wind_c = wind - mean(wind, na.rm = TRUE),
    sst_c2 = sst_c^2,
    chl_c2 = chl_c^2,
    wind_c2 = wind_c^2,
    biomass_t1_sardine_noharvest =  biomass_t1_sardine_noharvest / 1e5,
    biomass_t1_anchoveta_noharvest = biomass_t1_anchoveta_noharvest / 1e5,
    biomass_t1_jackmackerel_noharvest = biomass_t1_jackmackerel_noharvest / 1e5,
    biomass_sardine = biomass_sardine / 1e5,
    biomass_anchoveta = biomass_anchoveta / 1e5,
    biomass_jackmackerelintrapolated = biomass_jackmackerelintrapolated / 1e5,
    biomass_sardine_c = biomass_sardine - mean(biomass_sardine, na.rm = TRUE),
    biomass_anchoveta_c = biomass_anchoveta - mean(biomass_anchoveta, na.rm = TRUE),
    biomass_jackmackerelintrapolated_c = biomass_jackmackerelintrapolated - mean(biomass_jackmackerelintrapolated, na.rm = TRUE))
```

```{r sur_data_correlations, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

library(corrr)

# 1. Examine correlations among all standardized predictors (not responses)
corr_df <- sur_df %>%
  select(
    biomass_sardine, biomass_anchoveta, biomass_jackmackerelintrapolated,
    sst_c, sst_c2, chl_c, chl_c2, wind_c, wind_c2
  )
# Correlation matrix
corr_matrix <- correlate(corr_df)
corr_matrix %>% stretch(na.rm = TRUE) %>%
  filter(abs(r) > 0.7) %>%
  arrange(desc(abs(r))) %>%
  print(n = 20)
# Quick heatmap for visualization
library(ggplot2)
library(reshape2)
# Compute correlation matrix
corr_mat <- cor(corr_df, use = "pairwise.complete.obs")
# Melt to long format for ggplot
corr_long <- melt(corr_mat, varnames = c("Var1", "Var2"), value.name = "Correlation")
# Plot
ggplot(corr_long, aes(Var1, Var2, fill = Correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0,
                       limit = c(-1, 1), name = "r") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    panel.grid = element_blank(),
    aspect.ratio = 1
  ) +
  labs(title = "Correlation heatmap for standardized predictors",
       x = NULL, y = NULL)

# 2. Variance Inflation Factor (VIF)
# We'll test VIF within each equation's design matrix
eq_sardine_vif <- lm(
  biomass_t1_sardine_noharvest ~
    biomass_sardine_c + I(biomass_sardine_c^2) +  
    biomass_sardine_c:biomass_anchoveta_c +
    biomass_sardine_c:biomass_jackmackerelintrapolated_c +
    sst_c + sst_c2 + chl_c + chl_c2 + wind_c + wind_c2,
  data = sur_df
)
vif(eq_sardine_vif) %>% sort()

eq_anchoveta_vif <- lm(
  biomass_t1_anchoveta_noharvest ~
    biomass_anchoveta_c + I(biomass_anchoveta_c^2) +
    biomass_sardine_c:biomass_anchoveta_c + 
    biomass_anchoveta_c:biomass_jackmackerelintrapolated_c + 
    sst_c + sst_c2 + chl_c + chl_c2 + wind_c + wind_c2,
  data = sur_df
)
vif(eq_anchoveta_vif) %>% sort()

eq_jack_vif <- lm(
  biomass_t1_jackmackerel_noharvest ~
    biomass_jackmackerelintrapolated_c + I(biomass_jackmackerelintrapolated_c^2) +
    biomass_sardine_c:biomass_jackmackerelintrapolated_c + 
    biomass_anchoveta_c:biomass_jackmackerelintrapolated_c + 
    sst_c + sst_c2 + chl_c + chl_c2 + wind_c + wind_c2,
  data = sur_df
)
vif(eq_jack_vif) %>% sort()

```

```{r sur_estimation_base, message=FALSE, warning=FALSE, include=FALSE}

library(systemfit)

eq_sardine <- biomass_t1_sardine_noharvest ~
  biomass_sardine_c + I(biomass_sardine_c^2) +
  sst_c + sst_c2 + chl_c + chl_c2 + wind_c + wind_c2 +
  biomass_sardine_c:biomass_anchoveta_c +
  biomass_sardine_c:biomass_jackmackerelintrapolated_c

eq_anchoveta <- biomass_t1_anchoveta_noharvest ~
  biomass_anchoveta_c + I(biomass_anchoveta_c^2) +
  sst_c + sst_c2 + chl_c + chl_c2 + wind_c + wind_c2 +
  biomass_sardine_c:biomass_anchoveta_c +
  biomass_anchoveta_c:biomass_jackmackerelintrapolated_c

eq_jack <- biomass_t1_jackmackerel_noharvest ~
  biomass_jackmackerelintrapolated_c + I(biomass_jackmackerelintrapolated_c^2) +
  sst_c + sst_c2 + chl_c + chl_c2 + wind_c + wind_c2 +
  biomass_sardine_c:biomass_jackmackerelintrapolated_c + 
  biomass_anchoveta_c:biomass_jackmackerelintrapolated_c

fit_sur <- systemfit(
  list(
    sardine      = eq_sardine,
    anchoveta    = eq_anchoveta,
    jackmackerel = eq_jack
  ),
  data = sur_df,
  method = "SUR"
)

summary(fit_sur)

```

```{r function_for_testing_SUR, include=FALSE}
# --- Helper to build and fit SUR models with/without selected blocks ----
library(systemfit)

# 1. Builder of formulas given switches
build_eqs <- function(include_wind = TRUE,
                      include_sst = TRUE,
                      include_chl = TRUE,
                      include_interactions = TRUE) {
  


  sard_terms <- c("biomass_sardine_c", "I(biomass_sardine_c^2)",
                  if (include_interactions) c("biomass_sardine_c:biomass_jackmackerelintrapolated_c",
                                              "biomass_sardine_c:biomass_anchoveta_c"),
                  if (include_sst) c("sst_c", "sst_c2"),  
                  if (include_chl) c("chl_c", "chl_c2"), 
                  if (include_wind) c("wind_c", "wind_c2"))
  
  anch_terms <- c("biomass_anchoveta_c", "I(biomass_anchoveta_c^2)", 
                  if (include_interactions) c("biomass_anchoveta_c:biomass_jackmackerelintrapolated_c",
                                                          "biomass_sardine_c:biomass_anchoveta_c"),
                  if (include_sst) c("sst_c", "sst_c2"),  
                  if (include_chl) c("chl_c", "chl_c2"), 
                  if (include_wind) c("wind_c", "wind_c2"))
  jack_terms <- c("biomass_jackmackerelintrapolated_c", "I(biomass_jackmackerelintrapolated_c^2)",
                  if (include_interactions) c("biomass_sardine_c:biomass_jackmackerelintrapolated_c",
                                                          "biomass_anchoveta_c:biomass_jackmackerelintrapolated_c"),
                  if (include_sst) c("sst_c", "sst_c2"),  
                  if (include_chl) c("chl_c", "chl_c2"), 
                  if (include_wind) c("wind_c", "wind_c2"))

  f <- function(y, xs) as.formula(paste(y, "~", paste(xs, collapse = " + ")))

  list(
    sardine   = f("biomass_t1_sardine_noharvest", sard_terms),
    anchoveta = f("biomass_t1_anchoveta_noharvest", anch_terms),
    jack      = f("biomass_t1_jackmackerel_noharvest", jack_terms)
  )
}

# 2. Fit the SUR given switches
fit_sur_with <- function(include_wind = TRUE,
                         include_sst = TRUE,
                         include_chl = TRUE,
                         include_interactions = TRUE,
                         data = sur_df) {
  eqs <- build_eqs(include_wind, include_sst, include_chl, include_interactions)
  systemfit(
    list(
      sardine      = eqs$sardine,
      anchoveta    = eqs$anchoveta,
      jackmackerel = eqs$jack
    ),
    data = data,
    method = "SUR"
  )
}

# 3. Nested F-test between full and restricted systems
nested_F <- function(fit_full, fit_restr) {
  SSR_full  <- sum(sapply(fit_full$eq,  function(m) sum(resid(m)^2)))
  SSR_restr <- sum(sapply(fit_restr$eq, function(m) sum(resid(m)^2)))
  df_full   <- sum(sapply(fit_full$eq,  function(m) df.residual(m)))
  k_full    <- sum(sapply(fit_full$eq,  function(m) length(coef(m))))
  k_restr   <- sum(sapply(fit_restr$eq, function(m) length(coef(m))))
  q <- k_full - k_restr
  F_stat <- ((SSR_restr - SSR_full)/q) / (SSR_full/df_full)
  p_val  <- pf(F_stat, q, df_full, lower.tail = FALSE)
  data.frame(
    Restrictions = q,
    df2 = df_full,
    F = F_stat,
    p_value = p_val
  )
}
```

```{r test_exclusions, message=FALSE, warning=FALSE, include=FALSE}
# 1) Test dropping wind terms
fit_no_wind <- fit_sur_with(include_wind = FALSE,
                            include_sst = TRUE,
                            include_chl = TRUE,
                            include_interactions = TRUE,
                            data = sur_df)

nested_F(fit_sur, fit_no_wind)

fit_no_env <- fit_sur_with(include_wind = FALSE,
                            include_sst = FALSE,
                            include_chl = FALSE,
                            include_interactions = TRUE,
                            data = sur_df)

nested_F(fit_no_wind, fit_no_env)


fit_no_int <- fit_sur_with(include_wind = FALSE,
                            include_sst = TRUE,
                            include_chl = TRUE,
                            include_interactions = FALSE,
                            data = sur_df)

nested_F(fit_sur, fit_no_int)
nested_F(fit_no_wind, fit_no_int)

rm(build_eqs, fit_sur_with, eq_anchoveta, eq_jack, eq_sardine, sur_df)

```



```{r table_sur_final, message=FALSE, warning=FALSE, include=FALSE}
fit_biomasa <- fit_no_wind
# rm(fit_sur, fit_no_int, fit_no_wind)

library(dplyr)
library(tidyr)
library(kableExtra)

# --- 1. Extraer coeficientes, errores y p-values manualmente de cada ecuación ---
get_eq_df <- function(eq_obj, name){
  coefs <- summary(eq_obj)$coefficients
  data.frame(
    term = rownames(coefs),
    estimate = coefs[,1],
    std.error = coefs[,2],
    p.value = coefs[,4],
    model = name,
    row.names = NULL
  )
}

tidy_sur <- bind_rows(
  get_eq_df(fit_biomasa$eq[[1]], "Sardine"),
  get_eq_df(fit_biomasa$eq[[2]], "Anchoveta"),
  get_eq_df(fit_biomasa$eq[[3]], "JackMackerel")
)

# --- 2. Estandarizar nombres de variables e interacciones ---
tidy_sur <- tidy_sur %>%
  mutate(term = case_when(
    term %in% c("biomass_jackmackerelintrapolated_c:biomass_sardine_c",
                "biomass_sardine_c:biomass_jackmackerelintrapolated_c") ~
      "Sardine × Jack mackerel",
    term %in% c("biomass_anchoveta_c:biomass_jackmackerelintrapolated_c",
                "biomass_jackmackerelintrapolated_c:biomass_anchoveta_c") ~
      "Anchoveta × Jack mackerel",
    term %in% c("biomass_sardine_c:biomass_anchoveta_c",
                "biomass_anchoveta_c:biomass_sardine_c") ~
      "Sardine × Anchoveta",
    term == "biomass_sardine_c" ~ "Biomass (t)",
    term == "biomass_anchoveta_c" ~ "Biomass (t)",
    term == "biomass_jackmackerelintrapolated_c" ~ "Biomass (t)",
    term == "I(biomass_sardine_c^2)" ~ "Biomass sq (t)",
    term == "I(biomass_anchoveta_c^2)" ~ "Biomass sq (t)",
    term == "I(biomass_jackmackerelintrapolated_c^2)" ~ "Biomass sq (t)",
    term == "sst_c" ~ "SST",
    term == "sst_c2" ~ "(SST) sq",
    term == "chl_c" ~ "Chlorophyll-a",
    term == "chl_c2" ~ "Chlorophyll-a sq",
    term == "(Intercept)" ~ "Constant",
    TRUE ~ term
  ))

# --- 3. Agregar estrellas de significancia y formatear ---
tidy_sur <- tidy_sur %>%
  mutate(sig = case_when(
    p.value < 0.01 ~ "***",
    p.value < 0.05 ~ "**",
    p.value < 0.1  ~ "*",
    TRUE ~ ""
  ),
  estimate_fmt = sprintf("%.3f%s", estimate, sig)) %>%
  select(model, term, estimate_fmt)

# --- 4. Pivotear a formato tabla (una fila por variable) ---
sur_table <- tidy_sur %>%
  pivot_wider(names_from = model, values_from = estimate_fmt)

# --- 5. Agregar R² ajustado ---
r2_table <- data.frame(
  term = c("R-squared (Adj.)"),
  Sardine = sprintf("%.3f", summary(fit_biomasa$eq[[1]])$adj.r.squared),
  Anchoveta = sprintf("%.3f", summary(fit_biomasa$eq[[2]])$adj.r.squared),
  JackMackerel = sprintf("%.3f", summary(fit_biomasa$eq[[3]])$adj.r.squared)
)

sur_table <- bind_rows(sur_table, r2_table)
saveRDS(sur_table, "tables/table_SUR_stock.rds")

```



```{r include=FALSE}
library(kableExtra)
tab <- kbl(
  sur_table,
  align   = "lccc",
  caption = "\\label{tab:SUR_results}Seemingly Unrelated Regression (SUR) estimates of biomass dynamics for small pelagic species in Central--Southern Chile.",
  col.names = c("Variable", "Sardine", "Anchoveta", "Jack mackerel"),
  format  = "latex",
  booktabs = TRUE,
  escape  = FALSE,      # <- clave para permitir LaTeX en celdas y notas
  linesep = ""          # opcional: elimina espacio extra entre filas
) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    position = "center",
    font_size = 10
  ) %>%
  footnote(
    general = c(
      "Standard errors in parentheses. Significance levels: * p<0.1; ** p<0.05; *** p<0.01.",
      "Residual correlations ($\\\\rho$): Sardine--Anchoveta = -0.44; Sardine--Jack = -0.72; Anchoveta--Jack = +0.70."
    ),
    general_title = "",
    threeparttable = TRUE,
    escape = FALSE     
  ) 


```


Table \ref{tab:SUR_results} reports the Seemingly Unrelated Regression (SUR) estimates for annual biomass dynamics of sardine, anchoveta, and jack mackerel in central–southern Chile. The adjusted $R^2$ values...

Including sea surface temperature (SST) and chlorophyll (CHL) significantly improves model performance , leading us to reject the exclusion of environmental covariates at the 10% significance level (F = 1.908; $p$-value = 0.07).


Some facts:
  - Herrick, S. F. Jr, et al. in Climate Change and Small Pelagic Fish (eds Checkley, D. M., Roy, C., Alheit, J. & Oozeki, Y.) 256–274 (Cambridge Univ. Press, 2009): "Sardine is known to be more productive during warm-water regimes in the California Current ecosystem."
  

```{r insert_SUR_table, results='asis', echo=FALSE}
cat(tab)
```


**TAREA:** ¿Como usar errores robustos en un SUR?

## Trip level costo function
<< NO RESULTS YET >>

## Total Annual trips
<< NO RESULTS YET >>

## Inverse demand model

<< NO RESULTS YET >>

Two undergrad students working on this module for their thesis... Results by July 2026

## Numerical optimization
<< NO RESULTS YET >>

Once I have results for the four models, then I can run optimizations

# Discussion

- From @sumaila2011: "A shift in species’ geographic range will thus affect the distribution and composition of fisheries resources. This may affect fishing operations, the allocation of catch shares and the effectiveness of fisheries management measures"
- For the ITQ fisheries (i.e., industrial jack mackerel?: 
  - The theoretical findings on multispecies harvest patterns in @Birkenbach2020-nh give rise to nuanced hypotheses about how behavior and outcomes will change after the adoption of catch shares. For example, a secure property right to catch fish at any time in the fishing season allows firms to spread the catch of stocks with high prices and downward-sloping demand over a longer fishing season. This minimizes market gluts that steer product toward lower priced frozen markets [@homans2005] and can result in higher prices for those species. Fishermen might also shift their efforts toward lower-priced species with cheaper quota or toward non-catch-share fisheries, intensifying the race to fish for those species during portions of the season [@asche2007; @cunningham2016spillovers].
    - However, we do not include other species than jack mackerel, sardine and anchovy that might be caught by thise fleet. This would require  too expand the model by **N** species, which would increase dimensionality of the model. 
    - We need permits to check if actually this happen (... still problem with Open-Access fisheries that do not require permits)
- Although the present model captures the historical relationships between biomass, temperature, and productivity, future strengthening of coastal winds and warming of the Chilean coast (Fuenzalida et al. 2007; Garreaud et al. 2009) could shift these dynamics, altering species overlap and fishery composition.
- About using CHL for future predictions. 
  - From @sumaila2011: "However, our understanding of the magnitude and direction of climate change effects on primary productivity is uncertain. For example, projections of primary production based on an empirical relationship estimate increases of 0.7–8.1% by 2050 relative to 2000 (ref. 39), whereas outputs from four Earth-system models suggest a possible decrease of 2–20% by 2100 relative to pre-industrial conditions12. The opposite trends and the large regional differences of the estimates highlight the uncertainty in the projection of large-scale changes in primary productivity". 
- About prices [@sumaila2011]:
  - "Everything being equal, when climate change reduces fish supply, fish price should increase, which could be large enough to balance out the loss in gross revenues due to reduced catches. However, consumers may seek substitutes as prices increase, thereby dampening the demand for fish and reducing the potential for price increases."
  - "Ex-vessel revenues can be affected by climate change through effects first on the quantity, quality and distribution of catches, and then, on ex-vessel prices of fish."
- Adaptation  [@sumaila2011]:
  - "When climate change affects the composition and productivity of exploited species in a region, some fishers can adapt by switching target species or gear type81 or by moving to marginally productive areas. Grafton, R. Q. Adaptation to climate change in marine capture fisheries. Mar. Policy 34, 606–615 (2010)."
    - "For example, new fisheries have already developed for several southern species in the UK (for example, red mullet) as these species have started to migrate to the North Sea because of an increase in sea temperature. Beare, D. J. et al. Long-term increase in prevalence of North Sea fishes having southern biogeographic affinities. Mar. Ecol. Prog. Ser. 284, 269–278 (2004)."
    - "The ability of fishers and fishing enterprises to adapt depends on a number of factors, including the mobility of the fishing fleet.       - "At present, the dominant hypothesis is that more technologically advanced fleets, usually located in rich northern countries, are more likely to be better prepared to adapt to climate change by moving to other fishing grounds and by shifting gears. MacNeil, M. A. et al. Transitional states in marine fisheries: adapting to predicted global change. Phil. Trans. R. Soc. B 365, 3753–3763 (2011)."
    
    
- we did not added production risk.... risk preferences? maybe using random coefficients?  Discusse!!!

# Conclusions

NO YET!


## Future research question

- Several other extensions to the model can be incorporated to be improved. For instance, the geographical space where fishermen operate is relevant, as depending on the location chosen and when to participate, the set of potential choices would vary [@Reimer2017-jw]. As I mentioned above, it is possible to extend the stock dynamic model by considering different locations. The model would also require that the participation decision, which is captured by the Poisson model on the annual number of trips, should then consider the decision to participate in a determined fishing ground, connecting the multispecies model of @Kasperski2015-jm to the literature of location choice modeling [e.g., @Dupont1993-jn; @Smith2005-us; @Hicks2020-mz].

  - About the DCM?
    - Including other economic activity would be relevant, as aquaculture is become more relevant in the country, where some fishermen have swith to small-scale aquaculture practice.
    - Outside option...
  

- Does higher quota allocation of jack mackerel, a predator for anchovy and sardine, helps small scale sector actually?




# Repository

The source code for this project is available on
[GitHub](https://github.com/fquezadae/Impact-of-Environmental-Variability-on-Harvest)

# References
