---
title: "The Impact of Environmental Variability on Fishers' Harvest Decisions in Chile
  using a Multi-Species Approach"
author: "Felipe J. Quezada-Escalona"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  bookdown::pdf_document2:
    latex_engine: xelatex
    number_sections: true
    toc: false
    pandoc_args: ["--variable=fontsize:11pt"]
bibliography: bib_paperpile.bib
csl: apa.csl # Optional: for APA-style citations
linkcolor: blue
citecolor: blue
urlcolor: blue
link-citations: TRUE 
header-includes:
  - \usepackage{setspace}
  - \onehalfspacing
  - \usepackage{indentfirst}  # Enables first-line indentation
  - \setlength{\parindent}{10pt}  # Adjusts the indentation size
  - \usepackage{authblk}
  - \author{Felipe J. Quezada-Escalona}
  - \affil{Departmento de Economía \\ Universidad de Concepción \vspace{-48pt}}
editor_options: 
  markdown: 
    wrap: 72
abstract: "In this paper, we aim to answer how fishing decisions, aggregate catch levels, and the price of marine resources will be affected under different climatic scenarios in the multi-species small pelagic fishery (SPF) in Chile, composed by anchoveta (Engraulis ringens), jack mackerel (Trachurus murphyi), sardine (either Sardinops sagax or Strangomera bentincki), among others. By doing this, we expect to understand better how Chilean fishers and fishing communities will adapt to climate change. To address our research question, we will estimate a multi-species harvesting model. This model considers species' economic and biological interrelation to study the effect of climate variability on harvest decisions and substitution between species and determine the impact of different climatic scenarios on the well-being (e.g., profits) of fishers and fishing communities in Chile. We hypothesize that if the availability of a main target species is reduced, fishers will switch to the closest substitute if the expected revenue obtained from targeting this new species is high enough to cover the expected cost. Otherwise, the vessel would decrease fishing efforts or even exit the fishery due to the lack of economically viable substitutes. Moreover, we expect that this behavior is heterogeneous depending on the geographical area of operation -- as it determines the availability of other species-- and the gear type used."
---


# Introduction

The distribution and abundance of marine resources are changing in response to environmental conditions such as global ocean warming
[@Poloczanska2013-qq]. Due to climate change, species distribution is expected to change in the future, reducing species availability in some
areas but increasing in others. The literature that studies fishermen' responses to either changes in fish availability or policies that
restrict access to fisheries [e.g., @Stafford2018-pq; @Vasquez_Caballero2023-ip] has identified that fishers can adopt the
following adaptive strategies: (i) fishermen can reduce or reallocate fishing effort, either to another species or to another location
[@Gonzalez-Mon2021-kj], (ii) keep following the same strategy, or, (iii) in the worst-case scenario, stop fishing entirely and find alternative
employment elsewhere [@Powell2022-wj]. Among all those strategies, reallocating fishing efforts to other alternative species might be an
effective adaptation strategy to climate change [@Young2018-kk]. Diversification of target species has been associated with reducing
income variability [e.g., @Kasperski2013-jz; @Sethi2014-bn] and increasing resilience to both climate shock [@Cline2017-dp;
@Fisher2021-lw] and interannual oceanographic variability [@Aguilera2015-wo; @Finkbeiner2015-bs].

However, switching between species requires fishers to have the skills, the gear, and the permits to do so [@Frawley2021-cw; @Powell2022-wj].
Moreover, even though a fisher may satisfy these requirements, diversification might not be possible [@Beaudreau2019-xg] as it might be
constrained depending on port infrastructure, markets, and regulations [@Kasperski2013-jz; @Powell2022-wj]. Therefore, deciding which
adaptation strategy to take is not straightforward and would depend on many factors. Additionally, fishers might respond differently to an
analogous situation as they have different goals, skills, and preferences [@Zhang2011-wv; @Jardine2020-um; @Powell2022-wj].

In this project, we aim to answer how fishing decisions, aggregate catch levels, and the price of marine resources will be affected under
different climatic scenarios in the multi-species small pelagic fishery (SPF) in Chile, composed by anchoveta (*Engraulis ringens*), jack
mackerel (*Trachurus murphyi*), sardine (either *Sardinops sagax* or *Strangomera bentincki*), among others. The SPF is the most important in
terms of catches in the country, accounting for almost 94% of the total Chilean catch in 2019 [@SUBPESCA2020]. By doing this research, we expect
to understand better how Chilean fishers and fishing communities will adapt to climate change. To address our research question, we will
estimate a multi-species harvesting model based on @Kasperski2015-jm. This model considers species' economic and biological interrelation to
study the effect of climate variability on harvest decisions and substitution between species and determine the impact of different
climatic scenarios on the well-being (e.g., profits) of fishers and fishing communities in Chile.

We hypothesize that if the availability of a main target species is reduced, fishers will switch to the closest substitute if the expected
revenue obtained from targeting this new species is high enough to cover the expected cost. Otherwise, the vessel would decrease fishing efforts
or even exit the fishery due to the lack of economically viable substitutes. Moreover, we expect that this behavior is heterogeneous
depending on the geographical area of operation -- as it determines the availability of other species [@Reimer2017-jw] -- and the gear type
used.

At the end of the project, I expect to find significant effects of climate variables on species stock dynamics, the cost of fishing during
a trip, and the number of trips a vessel would take. The combinations of these environmental effects would be reflected in the optimal harvest
level and the prices seen on the local market. I also expect to find significant interrelations between species stock and harvest, and that
the composition of the catch will vary depending on the climate scenario we use for future predictions.

Under a changing climate, studying the effect of climatic variability on fishers' harvest decisions and landings is relevant for understanding fishing communities' adaptive capacities and strategies in response to climate change, thereby enabling the design of potential mitigation measures in response to these changes by policymakers. Countries have different institutions, cultures, and norms, so responses might differ depending on where the study is conducted. For this reason, conducting this research based on the Chilean fishing industry is necessary to develop local policies that aim to reduce climate change impacts on fisheries. While there is some literature on the effect of climate change on Chilean fisheries, I am unaware of local-level studies that consider a multiple-species framework and the interrelationship between the local market and fishing decisions seen under a variable climate context.[^1]

[^1]: For the case of Chile, as far as I know, the only article that study fishers' behavior using discrete choice modelling is @Pena-Torres2017-gn. This article study how El Niño Southern Oscillation (ENSO) affect fishermen location choices that participate in the Jack Mackerel fishery.

# SPF in Chile

- Composed by anchoveta (*Engraulis ringens*), jack mackerel (*Trachurus murphyi*), sardine (either *Sardinops sagax* or *Strangomera bentincki*), among others. 
- The SPF is the most important in terms of catches in the country, accounting for almost 94% of the total Chilean catch in 2019 [@SUBPESCA2020]. 

- Some facts:
  - Jack mackerel catched monstly by artisanal boats (race for fish)
  - Anchoveta and sardina is a mix species, with a mixed regulation. Separate quota, but if a vessel can´t found a spcies, can use the quota for that species to capture the other one.
  - There is some quota that goes through industrial vessel to artisanal (SERNAPESCA might have the data for this transactions)
  

# Data and methodology

To fulfill the project's objectives, and following @Kasperski2015-jm, the research entails five different stages: (i) estimating the stock
dynamics of each species included in the model, (ii) estimating trip level cost functions, (iii) estimating total annual trips, (iv) estimate
the inverse demand model for outputs (i.e., price responses to supply), and (v) conduct numerical optimization to examine how harvest and
profits levels evolve over time. The numeral optimization uses estimated parameters from the previous four stages to conduct the optimization
procedure.

## Data

- **SOLICITADO A IFOP 2012-2024:** 
  - Stock abundance and vessel landings (annual by port/county/region/country and species)
  - Data at the trip level ([IFOP data observatory?](ifop.dataobservatory.net)]).
  - Ex-vessel prices (monthly or annual by port/county/region/country and species)
  
- **POR SOLICITAR:** 
  - Environmental covariates -- Ask Fabian Tapia, UdeC
    - Sea surface temperature
    - Chlorophyll levels
    - Wind intensity and wave conditions in each trip at the harvest location
    - Bad weather days?
  - Other data?
    - Average wage pay to crew member per hour
    - Diesel cost.
    
    
```{r biomass_data, message=FALSE, warning=FALSE, include=FALSE}

library(readxl)
library(dplyr)

anch_sard_biomass <- read_excel("Data/3. ESTIMACIONES CRUCERO ACUSTICO.xlsx", sheet = "SARDINA-ANCHOVETA")
anch_sard_biomass <- anch_sard_biomass[, c(1, 3, 4)]
anch_sard_biomass <- anch_sard_biomass[-1, ]
colnames(anch_sard_biomass) <- c("year", "sardine_biomass", "anchoveta_biomass")
jackmackerel_biomass <- read_excel("Data/3. ESTIMACIONES CRUCERO ACUSTICO.xlsx", sheet = "JUREL")
jackmackerel_biomass  <- jackmackerel_biomass[, c(1, 3)]
jackmackerel_biomass  <- jackmackerel_biomass[-1, ]
colnames(jackmackerel_biomass) <- c("year", "jackmackerel_biomass")
anch_sard_biomass$year <- as.numeric(anch_sard_biomass$year)
jackmackerel_biomass$year <- as.numeric(jackmackerel_biomass$year)
biomass <- left_join(anch_sard_biomass, jackmackerel_biomass, by = "year")
rm(list = c("anch_sard_biomass", "jackmackerel_biomass"))

```


## Econometrics models

### Stock dynamics

To estimate stock dynamics, I use annual data on stock abundance and vessel landings. Following @Kasperski2015-jm, the growth of each species follows a
discrete logistic function: \begin{equation}
x_{i,y+1} + h_{iy} = \underbrace{(1 + r_i)x_{iy} + \eta_i x_{iy}^2}_{R_i(x_{iy})} + \underbrace{\sum_{j \neq i}^{n-1} a_{ij} x_{iy} x_{jy}}_{I_i(x_y)} \quad i=1,\ldots,n
\label{eq1}
\end{equation}

where $x_{iy}$ is the fish stock by species $i=1,\ldots,n$ in year $y$, $n$ is the total number of species, $h_{iy}$ is the annual harvest of
species $i$ on year $y$, $r_i$ is the intrinsic growth rate of the resource $i$, $\eta_i$ is a density-dependent factor related to the
carrying capacity, and $\alpha_{ij}$ are the interaction parameters between species. The system of $n$ growth equations can be estimated
simultaneously using seemingly unrelated regression (SUR) or other similar approaches.

Following @Richter2018, we can augment \eqref{eq1} by including environmental covariates $Env_{iy}$ that affect the fish stock, such as
sea surface temperature and chlorophyll levels, and an error term $\varepsilon_{iy}$ that captures random recruitment: \begin{equation}
x_{i,y+1} + h_{iy} = \underbrace{(1 + r_i)x_{iy} + \eta_i x_{iy}^2}_{R_i(x_{iy})} + \underbrace{\sum_{j \neq i}^{n-1} a_{ij} x_{iy} x_{jy}}_{I_i(x_y)} + \rho_i Env_{iy} + \varepsilon_{iy} \quad i=1,\ldots,n
\label{eq2}
\end{equation} where $\rho_i$ are the coefficient for the environmental
covariates. The model could also be expanded to different spatial locations conditional on data availability.


As shown in Figure \@ref(fig:biomass), biomass levels vary by species, and there is some interrelation between them. It is also clear that these biomass levels are affected by the harvests that occurred during those periods. For instance, in the case of jack mackerel, an abrupt decline in biomass is observed, likely due to a combination of overexploitation of the resource and unfavorable environmental conditions.

```{r biomass, fig.cap="Estimated biomass of small pelagic species in Chile (2000–2024)", fig.align='center', fig.pos='ht!', echo=FALSE, message=FALSE, warning=FALSE}

library(ggplot2)
library(tidyr)
library(scales)
library(viridis)

biomass_long <- biomass %>%
  pivot_longer(cols = c(sardine_biomass, anchoveta_biomass, jackmackerel_biomass),
               names_to = "species", 
               values_to = "biomass") %>%
  filter(!is.na(biomass)) %>%
  arrange(species, year)

biomass_long$biomass <- as.numeric(biomass_long$biomass)
biomass_long$year <- as.numeric(biomass_long$year)
biomass_long$species <- recode(biomass_long$species,
                               anchoveta_biomass = "Anchoveta",
                               sardine_biomass = "Sardine",
                               jackmackerel_biomass = "Jack Mackerel")
# Plot
ggplot(biomass_long, aes(x = year, y = biomass, color = species, group = species)) +
  geom_smooth(se = TRUE, method = "loess", span = 0.4, linetype = "solid") +
  scale_y_continuous(labels = comma) +  # Or use scale_y_log10(labels = comma) to compress
  labs(x = "Year",
       y = "Biomass (tons)",
       color = "Species") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.y = element_text(angle = 0, hjust = 1),
    legend.position = "right" 
  ) + 
  scale_color_viridis_d(option = "D")

rm(list = c("biomass_long"))

```
    
    




### Trip level cost functions

Ignoring trip subscript, the cost functions vary by vessel $v=1,\ldots,V_g$ and gear used $g=1,\ldots,G$, where $V_g$ is the number
of observations using gear type $g$, and $G$ is the total number of available (or observed) gears: \begin{equation}
C_{vg} = \sum_{i=1}^{2n+M+k} \alpha_{g, \mathbf{X}_i} \mathbf{X}_{ivg} + \frac{1}{2} \sum_{i=1}^{2n+M+k} \sum_{j=1}^{2n+M+k} \alpha_{g, \mathbf{X}_i\mathbf{X}_j} \mathbf{X}_{ivg} \mathbf{X}_{jvg}
\label{eq3}
\end{equation} where $C_{vg}= w z_{vg}^*$,
$\mathbf{X}_{vg}=[w;h_{vg};x;Z_v]$, $w$ is a $V_g \times M$ matrix of variable input prices, $h_{vg}$ is an $V_g \times n$ matrix of harvest
quantities, $x$ is an $V_g \times n$ matrix of given stock levels of the species of
interest, and $Z_v$ is an $V_g \times k$ matrix of given vessel characteristics. Therefore, $\mathbf{X}_{vg}$ is a $V_g \times (2n+M+k)$
matrix, and $\mathbf{X}_{ivg}$ represents the *i*th column of the $\mathbf{X}_{vg}$ matrix.

Together with estimating the restricted cost function, we estimate the conditional input demand equations. This addition allows an increase in
the degrees of freedom by imposing cross-equation parameter constraints and allows for the testing of, for instance, jointness in inputs
[@Kasperski2015-jm]. The conditional input demand equations are derived by Shepard's Lemma: \begin{equation}
\frac{\partial C_{vg}}{\partial w_m} = z^*_{vg,w_m} = \alpha_{g,w_m} + \sum_{j=1}^{2n+M+k} \alpha_{g,w_m,\mathbf{X}_j} \mathbf{X}_{jvg} \quad m=1,\ldots,M.
\label{eq4}
\end{equation}

Similar to stock dynamics, the system of equations formed by \eqref{eq3}
and \eqref{eq4} can be estimated using SUR. To comply with economic
theory, and to reduce even more the number of parameters to estimate, the following restrictions are imposed when estimating \eqref{eq4}:

1.  Symmetry of the cost function, where \begin{equation*}
      \alpha_{g,\mathbf{X}_i\mathbf{X}_j} = \alpha_{g,\mathbf{X}_j\mathbf{X}_i} \quad \forall \quad i=1,\ldots,(2n + M + k); \ i \neq j; \ g = 1,\ldots, G.
      \end{equation*}

2.  Linear homogeneity in input prices, where \begin{equation*}
      \sum_m^M \alpha_{g,w_m} = 1 \ \text{and} \ \sum_m^M \alpha_{g,w_m,\mathbf{X}_i} = 0 \quad i=1,\ldots,(2n + M + k); \ g = 1,\ldots, G.
      \end{equation*}

Data at the trip level is available upon request from the Chilean Fisheries Research Institute (IFOP), which registers geo-referenced
catch information on the Chilean fleet's fishing operation per trip [see e.g. @Pena-Torres2017-gn and [IFOP data observatory](ifop.dataobservatory.net)].
As inputs we can use the time spent at sea during a trip, where the price is the average wage pay to crew member per hour, and the distance
traveled during a trip, where the price of distance traveled is the diesel cost. Therefore, the total cost function $C_{vg}= w z_{vg}^*$ for vessel $v$, using gear $g$ in a trip would be sum of the total cost of distance travelled plus the total cost of the time spent at sea. 

**Note:** *Depending on the type of vessel, this cost should change. Some vessels are more efficient, other one are more heavy. How to capture this? The righ hand side has vessel characteristics, so the effect of harvest would be conditional on vessel characteristics, the stock levels and input prices. As we only care in the margin how harvest increase cost, this should be fine. @Kasperski2015-jm mention this "...no reliable fixed cost information on these vessels exists, but these should not affect the optimization as economic decisions are made at the margin. Therefore, this study does not measure true profit, but rather a proxy based on the net operating rent accruing to vessels in the fishery."*

To link this function to climate change, we can also include additional environmental variables $Env$ to $\mathbf{X}_{vg}$ such as wind intensity and wave conditions in each trip at the harvest location, upon data availability. Therefore, the augmented $X_{vg}$ matrix becomes $\mathbf{X}^{'}_{vg} = [w;h_{vg};x;Z_v;Env]$.


### Total annual trips

The number of trips a vessel will take in a given year for each gear
type used is assumed to follow a Poison distribution
[@Kasperski2015-jm]: \begin{equation}
Pr\left[T^{*}_{vgy} = t_v\right] = \frac{exp^{-exp(U^{'}_{vg}\beta_g)}exp(U^{'}_{vg}\beta_g)^{t_v}}{t_v !}
\label{eq5}
\end{equation} where $U_{vg}=[p,w,h_{vg},\bar{q},Z_{vg}]$ is a
$(3n+M+k+1)×V_g$ matrix of explanatory variables, $\beta_g$ is a
$(3n+M+k+1)\times1$ matrix of coefficients to be estimated, $t_v$ is the
number of trips taken by vessel $v$ using gear type $g$ in year $y$, and
$\bar{q}$ is the annual quota level. Additionally, we can add the
accumulation of "bad weather days" as an explanatory variable to
incorporate weather conditions into this decision, thus $U_{vg}=[p,w,h_{vg},\bar{q},Z_{vg}, Env]$

### Inverse demand model for outputs

The price of each species is modeled using an inverse demand model,
which assumes weak separability between the species into consideration
and other products [@Kasperski2015-jm]. The price of a species $i$ in
year $y$ is the following:

\begin{equation}
p_{iy} = \sum_j^n \gamma_j p_{j,y-1} + \gamma_{h_i} h_{iy} + \epsilon_{iy}, \quad i = 1,\ldots,n, \ j = 1,\ldots,n.
\label{eq6}
\end{equation}

The system formed by \eqref{eq6} can be estimated using maximum likelihood. Note that harvest may be endogenous in this system due to
simultaneity. @Kasperski2015-jm solves this by assuming that the TAC is exogenous, and the catch, in general, is determined by this quota.
We can relax this assumption by considering that all variables in the inverse demand equations are endogenous by estimating a vector autoregressive (VAR) model [@juselius2006cointegrated]. In other words, harvest $h_{vg}$ has its own equations in the system.

<!-- Comment: One concern I have regards the endogeneity of factors determining the outcomes of fishing effort on a fishing
trip (where, when, how long, and for what, and how much), climate conditions (temperature, wave, and wind),
and input and output prices. While Kasperski assumes that total allowable catches are binding to avoid one potential endogeneity, I wonder if this is a good assumption under climate change. What would you need to assume about the regulator’s response to the changing conditions and the setting of the total allowable catches? -->


## Numerical optimization

To obtain the effect of future climate variability on stock, harvest,
quota and profits, we conduct numerical optimization for different
climate scenarios using the parameters estimates for the stock dynamic,
cost functions, total annual trips and inverse demand equations. In each
year, a vessel maximizes profits by choosing their optimal number of
trips $T_g$ and harvest levels per trip $h_{g\tau}$ given a gear type:
\begin{align}
\max_{h_{gt}, T_g} \quad \pi_{vgt} & = \sum_{\tau=t}^{T_g} \rho^\tau \left\{ P(h) h_{g\tau} - C_g(h_{g\tau} | w, x, Z, Env) \right\} \quad \tau = t,\ldots, T_g \nonumber \\
\textbf{s.t} \quad q_{g,t+1} & = \omega \ast \bar{q} - \sum_{t=1}^{t} h_{gt} \geq 0, \quad t = 1, \dots, T-1, \quad g = 1, \dots, G
\label{eq7}
\end{align} where $\rho$ is the intra-annual discount factor, $\omega$ is a vector of shares of $\bar{q}$, and $h_{lt}=0$ for all $l\neq g$. The vector of shares is obtained from historical data on harvest. The
optimal profit from the maximization problem in \eqref{eq7} is denoted as $\pi_{vgy}^* (p,w,x,Z,\bar{q},\omega, Env)$, and $h_{vgty}^*$ and $T_{vgy}^*$ are the optimal choices harvest per trip and total number of
trips in year $y$ for vessel $v$. To obtain the optimal quota level, we must solve the social-planner optimization problem to maximize the net value of the fishery by choosing the quota levels per year and by species.

Following @Kasperski2015-jm, the optimization problem will be conducted for the next 25 years. I will use different climate scenarios and compare different optimal outcomes between them by using future projections for the environmental variables included in the model.

# Results

NO RESULTS YET

# Discussion

## Potential extension of the project

Several other extensions to the model can be incorporated to be
improved. For instance, the geographical space where fishermen operate
is relevant, as depending on the location chosen and when to
participate, the set of potential choices would vary [@Reimer2017-jw].
As I mentioned above, it is possible to extend the stock dynamic model
by considering different locations. The model would also require that
the participation decision, which is captured by the Poisson model on
the annual number of trips, should then consider the decision to
participate in a determined fishing ground, connecting the multi-species
model of @Kasperski2015-jm to the literature of location choice modeling
[e.g., @Dupont1993-jn; @Smith2005-us; @Hicks2020-mz].

# Conclusions

NO CONCLUSION YET

# Repository

The source code for this project is available on
[GitHub](https://github.com/fquezadae/Impact-of-Environmental-Variability-on-Harvest)

# References