---
title: "The Impact of Environmental Variability on Fishers' Harvest Decisions in Chile
  using a Multi-Species Approach"
subtitle: "\\begin{center}{\\color{red}\\Large\\textbf{REALLY EARLY DRAFT – PLEASE DO NOT CITE}}\\end{center}"
author: "Felipe J. Quezada-Escalona"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  bookdown::pdf_document2:
    latex_engine: xelatex
    number_sections: true
    toc: false
    pandoc_args: ["--variable=fontsize:11pt"]
  bookdown::latex_document2: default
bibliography: bibliography.bib
csl: apa.csl # Optional: for APA-style citations
linkcolor: blue
citecolor: blue
urlcolor: blue
link-citations: TRUE 
header-includes:
  - \usepackage{setspace}
  - \onehalfspacing
  - \usepackage{indentfirst}  # Enables first-line indentation
  - \setlength{\parindent}{10pt}  # Adjusts the indentation size
  - \usepackage{authblk}
  - \usepackage{booktabs}
  - \usepackage{caption}
  - \usepackage{dcolumn}
  - \usepackage{adjustbox}
  - \author{Felipe J. Quezada-Escalona}
  - \affil{Departmento de Economía \\ Universidad de Concepción \vspace{-48pt}}
editor_options: 
  markdown: 
    wrap: 72
abstract: "In this paper, we aim to answer how fishing decisions, aggregate catch levels, and the price of marine resources will be affected under different climatic scenarios in the multi-species small pelagic fishery (SPF) in Chile, composed by anchoveta (*Engraulis ringens*), jack mackerel (*Trachurus murphyi*), and sardine (*Strangomera bentincki*), among others. By doing this, we expect to gain a better understanding of how Chilean fishers and fishing communities will adapt to climate change. To address our research question, we will estimate a multi-species harvesting model. This model considers species' economic and biological interrelation to study the effect of climate variability on harvest decisions and substitution between species, and determine the impact of different climatic scenarios on the well-being (e.g., profits) of fishers and fishing communities in Chile. We hypothesize that when fishers have reduced access to a main target species, they will switch to the closest substitute if the expected revenue from targeting this new species exceeds the expected costs. Otherwise, the vessel would decrease fishing effort or even exit the fishery due to the lack of economically viable substitutes. Moreover, we expect that this behavior is heterogeneous depending on the geographical area of operation -- as it determines the availability of other species-- and the gear type used."
---

```{r directorio, message=FALSE, warning=FALSE, include=FALSE}
rm(list = ls())
gc()

# Definir directorios según usuario

usuario <- Sys.info()[["user"]]
# computador <- Sys.info()[["nodename"]]  # Alternativamente puedes usar esto
if (usuario == "felip") {
  dirdata <- "C:/Users/felip/OneDrive - Universidad de Concepción/FONDECYT Iniciacion/Data/"
} else if (usuario == "FACEA") {
  dirdata <- "C:/Users/FACEA/OneDrive - Universidad de Concepción/FONDECYT Iniciacion/Data/"
} else if (usuario == "Felipe") {
  dirdata <- "D:/OneDrive - Universidad de Concepción/FONDECYT Iniciacion/Data/"
} else {
  stop("Usuario no reconocido. Defina el directorio correspondiente.")
}
rm(usuario)

```

# Introduction

<!-- ADD COMMENTS FROM EFD! -->

The distribution and abundance of marine resources are changing in
response to environmental conditions such as global ocean warming
[@Poloczanska2013-qq]. Climate change will shift species distribution in
the future, leading to reduced species availability in some areas and
increased availability in others [@sumaila2011].
<!-- @sumaila2011: The economic consequences of climate change on fisheries might manifest themselves through changes in the price and value of catches, fishing costs, fishers’ incomes, earnings to fishing companies, discount rates and economic rent (that is, the surplus after all costs, including ‘normal’ profits, have been covered), as well as throughout the global economy. -->
The literature that studies fishers’ responses to either changes in fish
availability or policies that restrict access to fisheries [e.g.,
@Stafford2018-pq; @Vasquez_Caballero2023-ip] has identified that they
can adopt the following adaptive strategies: (i) reduce or reallocate
fishing effort, either to another species or to another location
[@Gonzalez-Mon2021-kj]; (ii) continue following the same strategy; or
(iii) exit the fishery and find alternative employment [@Powell2022-wj].
Among these strategies, reallocating effort to alternative species has
been identified as a potentially effective response to climate change
[@Young2018-kk]. Diversification of target species has also been linked
to reduced income variability [e.g., @Kasperski2013-jz; @Sethi2014-bn]
and greater resilience to both climate shocks [@Cline2017-dp;
@Fisher2021-lw] and interannual oceanographic variability
[@Aguilera2015-wo; @Finkbeiner2015-bs].

This emphasis on diversification aligns with broader evidence from
food-producing sectors. As @sjcruz_jmp highlights, climate variability
substantially affects agriculture and fisheries, where income depends
heavily on environmental fluctuations (e.g., temperature, rainfall) and
market forces (e.g., input costs) [@Kasperski2013-jz; @Carter2018]. With
variability expected to reduce productivity, income risk is likely to
rise [@Carter2018; @Free2019]. Diversification—whether within a sector
(e.g., switching crops or species) or across sectors—is often promoted
as an adaptive strategy [@Abbott2023-sb]. However, these strategies can
be costly for resource-dependent communities with limited capital and
skills [@Cherdchuchai2006; @Ellis2000], and the role of switching costs
in shaping diversification remains poorly understood.

In the fisheries context, switching between species requires not only
the skills but also the appropriate gear and permits [@Frawley2021-cw;
@Powell2022-wj]. Even if these conditions are met, diversification may
still be constrained by port infrastructure, markets, and regulations
[@Beaudreau2019-xg; @Kasperski2013-jz; @Powell2022-wj]. Therefore,
deciding which adaptation strategy to adopt is not straightforward and
depends on multiple factors. Moreover, fishers may respond differently
to similar circumstances depending on their goals, skills, and
preferences [@Zhang2011-wv; @Jardine2020-um; @Powell2022-wj].

In this research, we aim to answer how fishing decisions, aggregate
catch levels, and the price of marine resources will be affected under
different climatic scenarios in the multi-species small pelagic fishery
(SPF) in Chile, composed of anchoveta (*Engraulis ringens*), jack
mackerel (*Trachurus murphyi*), sardine (*Strangomera bentincki*), among
others. The SPF is the most important in terms of catches in the
country, accounting for almost 94% of the total Chilean catch in 2019
[@SUBPESCA2020]. Through this research, we aim to gain a deeper
understanding of how Chilean fishers and fishing communities will adapt
to climate change. According to @Cheung2010, the Chilean Exclusive
Economic Zone (EEZ) is projected to experience one of the largest losses
in maximum catch potential due to climate change.
<!-- Caviedes, C. N. & Fik, T. J. in Climate Variability, Climate Change and Fisheries (ed. Glantz, M.) 355–375 (Cambridge Univ. Press, 1992): "During the 1997–1998 El Niño event, Chilean and Peruvian pelagic marine landings declined by about 50%, resulting in a drop in fishmeal export values by about US$8.2 billion. This huge drop generated negative economic effects and caused severe hardship (lost jobs, incomes and earnings) in both countries"-->
However, the Southeast Pacific remains one of the least studied regions
regarding the impacts of climate change on fisheries [@sumaila2011].

To address our research question, we will estimate a multi-species
harvesting model based on @Kasperski2015-jm. This model considers
species' economic and biological interrelations to study the effect of
climate variability on harvest decisions and substitution between
species, and to determine the impact of different climatic scenarios on
the well-being (e.g., profits) of fishers and fishing communities in
Chile. We expect to find significant effects of climate variables on
species stock dynamics, the cost of fishing during a trip, and the
number of trips a vessel takes. These environmental effects might
influence optimal harvest levels and prices in local markets. 

Take into consideration the economic interaction between species is relevant to include how firms adjust to changes in policies, as relative prices between species might change. Also, cost complementarities between trageiting multispcies should be consideredn as it might be cheaper to harvest two species instead of one... In this paper we allow for economic interaction between species by allowing vessel to have multiple-output production and the output price to be dependent on other species, similar to @Kasperski2015-jm.
<!-- We also expect to find significant interrelations between species stock and harvest, and that the composition of the catch will vary depending on the climate scenario we use for future -->
<!-- predictions. -->

Under a changing climate, studying the effect of climatic variability on
fishers' harvest decisions and landings is relevant for understanding
fishing communities' adaptive capacities and strategies in response to
climate change, thereby enabling the design of potential mitigation
measures in response to these changes by policymakers. Countries have
different institutions, cultures, and norms, leading to differing
responses based on the study's location. For this reason, conducting
this research based on the Chilean fishing industry is necessary to
develop local policies that aim to reduce climate change impacts on
fisheries. While there is some literature on the effect of climate
change on Chilean fisheries, I am unaware of local-level studies that
consider a multiple-species framework and the interrelationship between
the local market and fishing decisions seen under a variable climate
context.[^1]

[^1]: For the case of Chile, as far as I know, the only article that
    studies fishers’ behavior using discrete choice modeling is
    @Pena-Torres2017-gn. This article studies how the El Niño–Southern
    Oscillation (ENSO) affects fishers’ location choices in the jack
    mackerel fishery.

Because predator–prey links couple these species, reductions in
anchoveta or sardine availability may reflect not only environmental
drivers but also changes in predation pressure from jack mackerel
[@Alheit2004; @Arancibia2019-FIPA]. Thus, even fishers who do not target
jack mackerel can face induced changes in catch rates and revenues
through ecosystem feedbacks. We hypothesize that if the availability of
a main target species decreases, fishers will switch to the closest
substitute when expected revenues (net of switching costs) exceed
expected costs; otherwise they may reduce effort or exit the fishery. We
also expect cross-fleet spillovers: in Chile, jack mackerel is
predominantly harvested by the industrial purse-seine fleet, whereas
anchoveta and sardine have substantial artisanal and industrial
participation. Shocks in one component can propagate across fleets via
both biology and markets, in addition to economic linkages (bycatch
constraints, shared gear, and market spillovers). This strengthens the
case for a multi-species framework that models joint dynamics and
substitution rather than single-species responses. Moreover, we expect
that this behavior is heterogeneous depending on the geographical area
of operation—as it determines the availability of other species
[@Reimer2017-jw]—and the gear type used.

-- ADD the fact that the government define the TAC using sa single
species model. This modelling effort allow to move further to a better
management of the SPF fishery by using a multispecies approach to obtain
optimal TAC

```{=html}
<!-- Los dos primeros, recordó, son especies clave para la cadena trófica marina, ya que alimentan a peces de mayor valor como corvina, lenguado o congrio, hoy cada vez más escasos.
LINK: https://www.diarioconcepcion.cl/economia/2025/10/07/sector-pesquero-nueva-resolucion-de-la-camara-baja-mantiene-la-discrepancia-entre-biobio-y-nuble.html -->
```

# SPF in Chile

The small pelagic fishery (SPF) in Chile is of critical importance to
the national fisheries sector. In 2019, the SPF represented nearly 94%
of total national fish landings [@SUBPESCA2020]. The fishery is
primarily composed of anchoveta (*Engraulis ringens*), sardine
(*Strangomera bentincki*), and jack mackerel (*Trachurus murphyi*).
While in the Northern region competition mainly occurs between anchoveta
and jack mackerel, in the Central-South region all three species play a
major role. This makes the Central-South particularly relevant for the
study of species interactions and potential substitution within a
multispecies management framework, and it is therefore the focus of this
research.

The jack mackerel fishery was initially concentrated in northern Chile,
but since the mid-1980s the main fishing grounds have shifted to
Central-South Chile, traditionally within 50 nautical miles of the coast
[@Pena-Torres2017-gn]. Historically, species in the SPF have been used
primarily for fishmeal and fish oil production [@Pena-Torres2017-gn]. In
fact, about 85% of jack mackerel landings, on a yearly average between
1987 and 2004, were destined for reduction into fishmeal and fish oil
[@Pena-Torres2017-gn]. Today, several key ports serve as hubs for these
activities, including San Antonio, Tomé, Talcahuano, San Vicente,
Coronel, Lota, and Corral.



"La pesquería de sardina común (Strangomera bentincki) y anchoveta (Engraulis ringens)
de la zona centro sur de Chile se caracteriza por ser una pesquería mixta. A pesar de contener
especies distintas, éstas conviven y se reproducen en un mismo hábitat. Las artes de pesca
utilizadas para capturar estas especies no permiten diferenciarlas." [@dresdner2013]

"La pesquería de sardina común y anchoveta ha sido por largos años una de las pesquerías
más importante de Chile. La disminución paulatina de los desembarques de jurel (Trachurus
murphyi) ha llevado a la sardina común a posicionarse como el principal recurso pelágico
extraído, seguido por la anchoveta" [@dresdner2013]


## Status of the stocks

Historically, anchoveta in the Central-South was considered collapsed
until 2018, shifted to overexploited status in 2019, and has since 2020
been fished within maximum sustainable yield (MSY) limits. Meanwhile,
sardine stocks have generally remained within MSY levels, except in 2021
and 2023 when they were classified as overexploited. Jack mackerel was
overexploited until 2018 but has since been harvested within MSY limits.

## Quota allocation

The Chilean fishing sector is managed primarily through a Total Allowable Catch (TAC; *Cuota Global*), which is divided between the industrial and artisanal sectors. A small share (≈2%) is reserved for research, with additional portions allocated to contingency and human consumption. The TAC is subdivided by region and season, and unused quotas may be reassigned during the fishing year.

Anchoveta and sardine are regulated as a mixed-species fishery: although each has its own quota, substitution between them is permitted. A share
of industrial quota is also periodically reassigned to the artisanal sector.


Since 2013, the industrial sector has operated under an individual transferable quota (ITQ) system, known as Transferable Fishing Licenses
(*Licencias Transables de Pesca, LTP*). Class A licenses were allocated based on historical catches, while Class B licenses—up to 15% of the
industrial fraction—are auctioned, with the first auctions held in 2015. These sealed-bid, first-price auctions aimed to broaden access and limit
concentration but have faced challenges such as low participation,
difficulties in reflecting economies of scale, and signs of potential
coordinated bidding [@peña_torres_2022_MRE].

The artisanal TAC operates under a regulated freedom-to-fish regime,
allowing registered vessels to fish without individual quotas, except in
areas where access is closed or suspended, in which case authorities may
implement management measures. The main measure is the Régimen Artesanal
de Extracción (RAE), which allocates the regional artisanal TAC by area,
vessel size, landing site (caleta), organization, or individually, in
agreement with artisanal fisher organizations. To date, area-based and
organization-based allocations are the only observed schemes. Area-based
allocations allow registered artisanal vessels in a given area to fish
as in open access until the assigned quota is exhausted, while
organization-based allocations follow the historical rights of members
to distribute the organization’s quota.

-   Sardine: RAE in V, VIII Y X regions? What about other species? Open
    access in anchovy and jack mackerel (only artisanal TAC matter at
    country level?)

### Chile regionalized fisheries governance framework

Chile has a regionalize fisheres governance framework, where boat
register in one region can not fish in another one. For instance, a
recent conflict between the Biobío and Ñuble regions has reignited the
debate over the spatial governance of the small pelagic fishery (SPF) in
south–central Chile. In late August 2025, the Chilean Chamber of
Deputies approved a resolution urging the Government to repeal the
authorization that allows vessels from Biobío to operate in the coastal
waters of Ñuble. The measure, promoted by local authorities and
artisanal organizations from Ñuble, aims to protect local fishing
grounds and reduce pressure on nearshore ecosystems. However,
representatives from Biobío have warned that such restrictions could
have severe economic consequences for the region, given its strong
dependence on small pelagic landings. This episode highlights the
institutional tensions arising from Chile’s regionalized fisheries
governance framework, where jurisdictional boundaries often conflict
with the biological and economic interdependencies of fish stocks.

## Other regulations

### Limited entry

Fishery with restricted access to new operators

### Biological closures for recruitment

-   Jack mackerel is open through all year.
-   Sardine and anchovy: In southern-central Chile, December–March
    (fixed period: January to February).

### Biological closures for reproduction

-   Jack mackerel is open through all year.

-   Sardine and anchovy: In southern-central Chile, July–October (fixed
    period: August-September).

-   Seasonality? Include quarter dummies? Jack mackerel gather in the
    first 6 month of the year in shoals, great density in EEZ, then
    migrate outside 200nm ()

### Minimum size

-   Jack mackerel: 26 cm
-   Sardine and anchovy?

### Maximum harvest levels

-   All species: Maximum catch limit per vessel owner (LMC) for industrial vessels, based on the industrial share of the TAC.

```{r load_env_vars, message=FALSE, warning=FALSE, include=FALSE}

env_dt <- readRDS(paste0(dirdata, "Environmental/env/EnvCoastDaily_2012_2025_0.125deg.rds"))
env_dt_00_11 <- readRDS(paste0(dirdata, "Environmental/env/2000-2011/EnvCoastDaily_2000_2011_0.25deg.rds"))

```

```{r load_biomass, message=FALSE, warning=FALSE, include=FALSE}

biomass <- readRDS("data/biomass/biomass_dt.rds")

```

```{r load_logbook, message=FALSE, warning=FALSE, include=FALSE}

logbooks <- readRDS("data/logbooks/logbooks.rds")

```

## Switching patterns

ARE THIS SPF TRIPS?????

```{r harvest_substitution_after, message=FALSE, warning=FALSE, include=FALSE}

library(dplyr)
library(data.table)
library(lubridate)
library(sf)
library(rnaturalearth)
library(ggplot2)
library(tidyr)
library(purrr)

# Build shares per vessel/fleet/year/species
vessel_species_year <- logbooks %>%
  filter(year >= 2019) %>%
  select(COD_BARCO, TIPO_FLOTA, year, NOMBRE_ESPECIE, CAPTURA_RETENIDA) %>%
  group_by(COD_BARCO, TIPO_FLOTA, year, NOMBRE_ESPECIE) %>%
  summarise(total_catch = sum(CAPTURA_RETENIDA, na.rm = TRUE), .groups = "drop") %>%
  group_by(COD_BARCO, TIPO_FLOTA, year) %>%
  mutate(species_share = total_catch / sum(total_catch, na.rm = TRUE)) %>%
  ungroup() %>%
  select(-total_catch) %>%
  complete(COD_BARCO, TIPO_FLOTA, year, NOMBRE_ESPECIE, fill = list(species_share = 0)) %>%
  group_by(COD_BARCO, TIPO_FLOTA, year) %>%
  filter(sum(species_share, na.rm = TRUE) > 0) %>%
  group_by(COD_BARCO, TIPO_FLOTA, NOMBRE_ESPECIE) %>%
  summarise(species_share = mean(species_share, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = NOMBRE_ESPECIE,
    values_from = species_share,
    values_fill = 0
  ) %>%
  # here just add missing columns if they don't exist
  { if(!"ANCHOVETA" %in% names(.)) .$ANCHOVETA <- 0; . } %>%
  { if(!"SARDINA COMUN" %in% names(.)) .$`SARDINA COMUN` <- 0; . } %>%
  { if(!"JUREL" %in% names(.)) .$JUREL <- 0; . } %>%
  mutate(
    Other = 1 - ANCHOVETA - `SARDINA COMUN` - JUREL,
    Anchoveta = ANCHOVETA,
    Sardine = `SARDINA COMUN`,
    JackMackerel = JUREL
  ) %>%
  select(COD_BARCO, TIPO_FLOTA, Anchoveta, Sardine, JackMackerel, Other)

# Strategy function
get_strategy <- function(sardine, jackmackerel, anchoveta, other) {
  species <- c()
  if (sardine > 0.15) species <- c(species, "Sardine")
  if (jackmackerel > 0.15) species <- c(species, "JackMackerel")
  if (anchoveta > 0.15) species <- c(species, "Anchoveta")
  if (other > 0.15) species <- c(species, "Other")
  
  n <- length(species)
  if (n == 0) return("None or negligible")
  if (n == 1) return(paste("Only", species[1]))
  if (n == 2) return(paste0(species[1], " and ", species[2]))
  if (n == 3) return(paste0(species[1], ", ", species[2], " and ", species[3]))
  return(paste0(species, collapse = ", "))
}

# Apply strategy classification
vessel_species_year <- vessel_species_year %>%
  mutate(strategy_After = pmap_chr(
    list(Sardine, JackMackerel, Anchoveta, Other),
    get_strategy
  ))

# Percentages by fleet type
strategy_percent <- vessel_species_year %>%
  group_by(TIPO_FLOTA, strategy_After) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(TIPO_FLOTA) %>%
  mutate(percent = round(100 * n / sum(n), 1)) %>%
  arrange(TIPO_FLOTA, desc(percent))

# Final table vessel-strategy
strategy_after <- vessel_species_year %>%
  select(COD_BARCO, TIPO_FLOTA, strategy_After)

```

```{r harvest_substitution_before, message=FALSE, warning=FALSE, include=FALSE}

# Build shares per vessel/fleet/year/species
vessel_species_year <- logbooks %>%
  filter(year < 2019) %>%
  select(COD_BARCO, TIPO_FLOTA, year, NOMBRE_ESPECIE, CAPTURA_RETENIDA) %>%
  group_by(COD_BARCO, TIPO_FLOTA, year, NOMBRE_ESPECIE) %>%
  summarise(total_catch = sum(CAPTURA_RETENIDA, na.rm = TRUE), .groups = "drop") %>%
  group_by(COD_BARCO, TIPO_FLOTA, year) %>%
  mutate(species_share = total_catch / sum(total_catch, na.rm = TRUE)) %>%
  ungroup() %>%
  select(-total_catch) %>%
  complete(COD_BARCO, TIPO_FLOTA, year, NOMBRE_ESPECIE, fill = list(species_share = 0)) %>%
  group_by(COD_BARCO, TIPO_FLOTA, year) %>%
  filter(sum(species_share, na.rm = TRUE) > 0) %>%
  group_by(COD_BARCO, TIPO_FLOTA, NOMBRE_ESPECIE) %>%
  summarise(species_share = mean(species_share, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = NOMBRE_ESPECIE,
    values_from = species_share,
    values_fill = 0
  ) %>%
  # here just add missing columns if they don't exist
  { if(!"ANCHOVETA" %in% names(.)) .$ANCHOVETA <- 0; . } %>%
  { if(!"SARDINA COMUN" %in% names(.)) .$`SARDINA COMUN` <- 0; . } %>%
  { if(!"JUREL" %in% names(.)) .$JUREL <- 0; . } %>%
  mutate(
    Other = 1 - ANCHOVETA - `SARDINA COMUN` - JUREL,
    Anchoveta = ANCHOVETA,
    Sardine = `SARDINA COMUN`,
    JackMackerel = JUREL
  ) %>%
  select(COD_BARCO, TIPO_FLOTA, Anchoveta, Sardine, JackMackerel, Other)

# Apply strategy classification
vessel_species_year <- vessel_species_year %>%
  mutate(strategy_Before = pmap_chr(
    list(Sardine, JackMackerel, Anchoveta, Other),
    get_strategy
  ))

# Percentages by fleet type
strategy_percent_before <- vessel_species_year %>%
  group_by(TIPO_FLOTA, strategy_Before) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(TIPO_FLOTA) %>%
  mutate(percent = round(100 * n / sum(n), 1)) %>%
  arrange(TIPO_FLOTA, desc(percent))

# Final table vessel-strategy
strategy_before <- vessel_species_year %>%
  select(COD_BARCO, TIPO_FLOTA, strategy_Before)

rm(vessel_species_year)

```

See Figure \@ref(fig:switchStrategy-ART) for strategy transitions. The
year 2019 is used as reference as anchoveta and jack mackerel started to
recover.

```{r strategy_table_logbooks_ART, message=FALSE, warning=FALSE, include=FALSE, results='asis'}

library(kableExtra)
library(stringr)

# Rename for joining
before <- strategy_percent_before %>%
  rename(strategy = strategy_Before, n_before = n, percent_before = percent)

after <- strategy_percent %>%
  rename(strategy = strategy_After, n_after = n, percent_after = percent)

# Merge strategies
strategy_summary <- full_join(before, after, by = c("strategy", "TIPO_FLOTA")) %>%
  arrange(desc(coalesce(percent_after, 0) + coalesce(percent_before, 0))) %>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

# Filter only ART
strategy_table_art <- strategy_summary %>%
  filter(TIPO_FLOTA == "ART") %>%
  ungroup() %>%
  select(-TIPO_FLOTA)

```

Table \@ref(tab:table-art) for strategy transitions.

```{r table-art, echo=FALSE, results='asis'}

# Create table with grouped columns
strategy_table_art %>%
  select(strategy, n_before, percent_before, n_after, percent_after) %>%
  kable("latex", booktabs = TRUE, digits = 1,
        col.names = c("Strategy", "n", "%", "n", "%"),
        caption = "Comparison of Strategies Before and After -- Small-scale vessels") %>%
  add_header_above(c(" " = 1, "Before" = 2, "After" = 2)) %>%
  kable_styling(font_size = 10)

rm(strategy_table_art)


```

```{r strategy-table-logbooks_IND, message=FALSE, warning=FALSE, include=FALSE, results='asis'}

library(kableExtra)
library(stringr)

# Rename for joining
before <- strategy_percent_before %>%
  rename(strategy = strategy_Before, n_before = n, percent_before = percent)

after <- strategy_percent %>%
  rename(strategy = strategy_After, n_after = n, percent_after = percent)

# Merge strategies
strategy_summary <- full_join(before, after, by = c("strategy", "TIPO_FLOTA")) %>%
  arrange(desc(coalesce(percent_after, 0) + coalesce(percent_before, 0))) %>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

# Filter only IND
strategy_table_ind <- strategy_summary %>%
  filter(TIPO_FLOTA == "IND") %>%
  ungroup() %>%
  select(-TIPO_FLOTA)

rm(strategy_summary, strategy_percent, strategy_percent_before, after, before)

```

Table \@ref(tab:table-ind) for industrial strategy transitions.

```{r table-ind, echo=FALSE, results='asis'}

# Create table with grouped columns
strategy_table_ind %>%
  select(strategy, n_before, percent_before, n_after, percent_after) %>%
  kable("latex", booktabs = TRUE, digits = 1,
        col.names = c("Strategy", "n", "%", "n", "%"),
        caption = "Comparison of Strategies Before and After -- Industrial vessels") %>%
  add_header_above(c(" " = 1, "Before" = 2, "After" = 2)) %>%
  kable_styling(font_size = 10)

rm(strategy_table_ind)

```

```{r switchStrategy-ART, echo=FALSE, fig.height=6, fig.width=8, message=FALSE, warning=FALSE, results='asis', fig.cap='Strategy transitions for small-scale vessels'}

library(dplyr)
library(tidyr)
library(ggplot2)
library(ggalluvial)
library(patchwork)
library(ggrepel)

strategy_transitions <- full_join(strategy_before, strategy_after, by = c("COD_BARCO", "TIPO_FLOTA")) %>%
  mutate(
    strategy_Before = ifelse(is.na(strategy_Before), "No fishing", strategy_Before),
    strategy_After  = ifelse(is.na(strategy_After),  "No fishing", strategy_After)
  ) %>% filter(TIPO_FLOTA == "ART")

transition_counts <- strategy_transitions %>%
  count(strategy_Before, strategy_After)

simplify_focus_strategy <- function(strategy) {
  ifelse(strategy == "Only Sardine", "Sardine",
  ifelse(strategy == "Only Anchoveta", "Anchoveta",
  ifelse(strategy == "Only JackMackerel", "JackMackerel",
  ifelse(grepl("Sardine", strategy) & grepl("Anchoveta", strategy) & grepl("JackMackerel", strategy),
         "Sardine, JackMackerel & Anchoveta",
  ifelse(grepl("Sardine", strategy) & grepl("Anchoveta", strategy), "Sardine & Anchoveta",
  ifelse(grepl("Sardine", strategy) & grepl("JackMackerel", strategy), "Sardine & JackMackerel",
  ifelse(grepl("JackMackerel", strategy) & grepl("Anchoveta", strategy), "JackMackerel & Anchoveta",
  ifelse(strategy == "No fishing", "No fishing", "Other"))))))))
}

strategy_transitions_focus <- strategy_transitions %>%
  mutate(
    strategy_Before_simple = simplify_focus_strategy(strategy_Before),
    strategy_After_simple = simplify_focus_strategy(strategy_After)
  ) 

# %>%
#   filter(strategy_Before_simple %in% c("Sardine", "Anchoveta", "Sardine & Anchoveta", "Sardine & JackMackerel", "JackMackerel & Anchoveta", "Sardine, JackMackerel & Anchoveta","No fishing") |
#            strategy_After_simple %in% c("Sardine", "Anchoveta", "Sardine & Anchoveta", "Sardine & JackMackerel", "JackMackerel & Anchoveta", "Sardine, JackMackerel & Anchoveta","No fishing"))

transition_counts_focus <- strategy_transitions_focus %>%
  count(strategy_Before_simple, strategy_After_simple)


# Base plot
p <- ggplot(transition_counts_focus,
            aes(axis1 = strategy_Before_simple,
                axis2 = strategy_After_simple,
                y = n)) +
  geom_flow(aes(fill = strategy_Before_simple),
            width = 1/12, knot.pos = 0.4, alpha = 0.8) +
  geom_stratum(width = 1/12, fill = "gray95", color = "black") +
  scale_x_discrete(limits = c("2012–2018", "2019–2024"),
                   expand = c(.25, .25)) +
  scale_fill_brewer(type = "qual", palette = "Set2") +
  labs(x = "", y = "Number of Vessels") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.grid = element_blank()
  )

# Extraer posiciones de etiquetas
stratum_data <- ggplot_build(p)$data[[2]]

left_labels <- stratum_data %>%
  filter(x == 1) %>%
  mutate(x = x - 0.05,
         stratum = str_wrap(stratum, width = 20))

right_labels <- stratum_data %>%
  filter(x == 2) %>%
  mutate(x = x + 0.05,
         stratum = str_wrap(stratum, width = 20))

# Agregar etiquetas con repel
p <- p +
  geom_text_repel(
    data = left_labels,
    aes(x = x, y = y, label = stratum),
    inherit.aes = FALSE,
    hjust = 1, nudge_x = -0.1, direction = "y",
    size = 2.5, segment.color = "grey60", segment.size = 0.3
  ) +
  geom_text_repel(
    data = right_labels,
    aes(x = x, y = y, label = stratum),
    inherit.aes = FALSE,
    hjust = 0, nudge_x = 0.1, direction = "y",
    size = 2.5, segment.color = "grey60", segment.size = 0.3
  )

p

ggsave("figs/strategy_transitions_ART.svg", p,
       device = svglite::svglite, width = 10, height = 6, units = "in")

```

```{r switchStrategy-IND, echo=FALSE, fig.cap='Strategy transitions for industrial vessels', fig.height=6, fig.width=8, message=FALSE, warning=FALSE, results='asis'}

library(dplyr)
library(tidyr)
library(ggplot2)
library(ggalluvial)
library(patchwork)
library(ggrepel)

strategy_transitions <- full_join(strategy_before, strategy_after, by = c("COD_BARCO", "TIPO_FLOTA")) %>%
  mutate(
    strategy_Before = ifelse(is.na(strategy_Before), "No fishing", strategy_Before),
    strategy_After  = ifelse(is.na(strategy_After),  "No fishing", strategy_After)
  ) %>% filter(TIPO_FLOTA == "IND")

transition_counts <- strategy_transitions %>%
  count(strategy_Before, strategy_After)

simplify_focus_strategy <- function(strategy) {
  ifelse(strategy == "Only Sardine", "Sardine",
  ifelse(strategy == "Only Anchoveta", "Anchoveta",
  ifelse(strategy == "Only JackMackerel", "JackMackerel",
  ifelse(grepl("Sardine", strategy) & grepl("Anchoveta", strategy) & grepl("JackMackerel", strategy),
         "Sardine, JackMackerel & Anchoveta",
  ifelse(grepl("Sardine", strategy) & grepl("Anchoveta", strategy), "Sardine & Anchoveta",
  ifelse(grepl("Sardine", strategy) & grepl("JackMackerel", strategy), "Sardine & JackMackerel",
  ifelse(grepl("JackMackerel", strategy) & grepl("Anchoveta", strategy), "JackMackerel & Anchoveta",
  ifelse(strategy == "No fishing", "No fishing", "Other"))))))))
}

strategy_transitions_focus <- strategy_transitions %>%
  mutate(
    strategy_Before_simple = simplify_focus_strategy(strategy_Before),
    strategy_After_simple = simplify_focus_strategy(strategy_After)
  ) 

# %>%
#   filter(strategy_Before_simple %in% c("Sardine", "Anchoveta", "Sardine & Anchoveta", "Sardine & JackMackerel", "JackMackerel & Anchoveta", "Sardine, JackMackerel & Anchoveta","No fishing") |
#            strategy_After_simple %in% c("Sardine", "Anchoveta", "Sardine & Anchoveta", "Sardine & JackMackerel", "JackMackerel & Anchoveta", "Sardine, JackMackerel & Anchoveta","No fishing"))# %>%
#   filter(strategy_Before_simple %in% c("Sardine", "Anchoveta", "Sardine & Anchoveta", "Sardine & JackMackerel", "JackMackerel & Anchoveta", "Sardine, JackMackerel & Anchoveta","No fishing") |
#            strategy_After_simple %in% c("Sardine", "Anchoveta", "Sardine & Anchoveta", "Sardine & JackMackerel", "JackMackerel & Anchoveta", "Sardine, JackMackerel & Anchoveta","No fishing"))

transition_counts_focus <- strategy_transitions_focus %>%
  count(strategy_Before_simple, strategy_After_simple)


# Base plot
p <- ggplot(transition_counts_focus,
            aes(axis1 = strategy_Before_simple,
                axis2 = strategy_After_simple,
                y = n)) +
  geom_flow(aes(fill = strategy_Before_simple),
            width = 1/12, knot.pos = 0.4, alpha = 0.8) +
  geom_stratum(width = 1/12, fill = "gray95", color = "black") +
  scale_x_discrete(limits = c("2012–2018", "2019–2024"),
                   expand = c(.25, .25)) +
  scale_fill_brewer(type = "qual", palette = "Set2") +
  labs(x = "", y = "Number of Vessels") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.grid = element_blank()
  )

# Extraer posiciones de etiquetas
stratum_data <- ggplot_build(p)$data[[2]]

left_labels <- stratum_data %>%
  filter(x == 1) %>%
  mutate(x = x - 0.05,
         stratum = str_wrap(stratum, width = 20))

right_labels <- stratum_data %>%
  filter(x == 2) %>%
  mutate(x = x + 0.05,
         stratum = str_wrap(stratum, width = 20))

# Agregar etiquetas con repel
p <- p +
  geom_text_repel(
    data = left_labels,
    aes(x = x, y = y, label = stratum),
    inherit.aes = FALSE,
    hjust = 1, nudge_x = -0.1, direction = "y",
    size = 2.5, segment.color = "grey60", segment.size = 0.3
  ) +
  geom_text_repel(
    data = right_labels,
    aes(x = x, y = y, label = stratum),
    inherit.aes = FALSE,
    hjust = 0, nudge_x = 0.1, direction = "y",
    size = 2.5, segment.color = "grey60", segment.size = 0.3
  )

p

ggsave("figs/strategy_transitions_IND.svg", p,
       device = svglite::svglite, width = 10, height = 6, units = "in")

rm(left_labels, right_labels, strategy_after, strategy_before, strategy_transitions, strategy_transitions_focus, stratum_data, transition_counts, transition_counts_focus, p)

rm(get_strategy, simplify_focus_strategy)

```

## Fishing seasons

```{r harvest_data, message=FALSE, warning=FALSE, include=FALSE}

library(readxl)
library(dplyr)

harvest_IFOP_month <- readRDS("data/harvest/IFOP_month.rds") 
harvest_IFOP <- readRDS("data/harvest/IFOP.rds")
harvest_SERNAPESCA <- readRDS("data/harvest/sernapesca.rds")
harvest_SERNAPESCA_v2 <- readRDS("data/harvest/sernapesca_v2.rds")


#---- IFOP loogbook data ----

harvest_IFOP_logbooks <- logbooks %>% 
  filter(year >= 2012) %>%
  filter(REGION %in% c(5,6,7,8,9,10,14,16)) %>%
  group_by(NOMBRE_ESPECIE, year) %>%
  summarise(
    harvest_IFOP_loogbook_centrosur = sum(CAPTURA_RETENIDA, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  rename(specie = NOMBRE_ESPECIE) %>%
  mutate(harvest_IFOP_loogbook_centrosur = harvest_IFOP_loogbook_centrosur / 1000)


#---- Harvest data merged ---- 
harvest <- left_join(harvest_SERNAPESCA_v2, harvest_SERNAPESCA, by = c("year", "specie")) 
harvest <- left_join(harvest, harvest_IFOP, by = c("year", "specie"))
harvest <- left_join(harvest, harvest_IFOP_logbooks, by = c("year", "specie")) 

rm(harvest_IFOP, harvest_SERNAPESCA, harvest_SERNAPESCA_v2, harvest_IFOP_logbooks)

```

```{r season_SPF, eval=FALSE, fig.align='center', fig.cap="Fishery seasons", fig.pos='ht!', message=FALSE, warning=FALSE, include=FALSE}

months <- c("E", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D")

# Fishing periods
sardine_periods <- data.frame(
  start = c(1, 3,7,10),    
  end = c(3,7,10,12),      
  status = c("Closed", "Open", "Closed", "Open")
)

# Cruise brackets
cruise_brackets <- data.frame(
  cruise = c("1st Cruise (Summer)", "2nd Cruise (Fall)"),
  x_start = c(1, 3),   
  x_end = c(3, 6),      
  y = c(1.05, 1.05)       
)

# Plot with reduced white space
ggplot() +
  geom_segment(data = sardine_periods, aes(x = start, xend = end, y = 1, yend = 1, color = status), size = 4) +
  geom_segment(data = cruise_brackets, aes(x = x_start, xend = x_end, y = y, yend = y), size = 1) +
  geom_segment(data = cruise_brackets, aes(x = x_start, xend = x_start, y = y, yend = y - 0.05), size = 1) +
  geom_segment(data = cruise_brackets, aes(x = x_end, xend = x_end, y = y, yend = y - 0.05), size = 1) +
  geom_text(data = cruise_brackets, aes(x = (x_start + x_end) / 2, y = y + 0.05, label = cruise), size = 3.5) +
  scale_x_continuous(breaks = 1:12, labels = months, expand = c(0, 0)) +
  scale_y_continuous(limits = c(0.9, 1.3), breaks = 1, labels = " ", expand = c(0, 0)) +
  scale_color_manual(values = c("Open" = "green", "Closed" = "red")) +
    labs(x = "Month", y = NULL) +
  theme_minimal() + coord_cartesian(clip = "off") +
theme(
  plot.margin = margin(2, 2, 2, 2),  # very small margins (t, r, b, l)
  axis.title.x = element_text(margin = margin(t = 0)),  # reduce margin below x-axis label
  axis.text.x = element_text(margin = margin(t = 0)),   # reduce space below x-axis ticks
  plot.title = element_text(hjust = 0.5, margin = margin(b = 0))  # reduce margin above title
)

rm(cruise_brackets, sardine_periods)

```

```{r monthlyharvest, echo=FALSE, fig.align='center', fig.cap="Average monthly landings by species (2012-2024; South-Central Chile) ", fig.pos='ht!', message=FALSE, warning=FALSE}

library(dplyr)
library(stringr)

# Diccionario
month_translation <- c(
  "ene" = "jan", "feb" = "feb", "mar" = "mar", "abr" = "apr",
  "may" = "may", "jun" = "jun", "jul" = "jul", "ago" = "aug", 
  "sept" = "sep", "oct" = "oct", "nov" = "nov", 
  "dic" = "dec"
)

# Orden destino
month_order <- c("jan","feb","mar","apr","may","jun","jul","aug","sep","oct","nov","dec")

harvest_summary <- harvest_IFOP_month %>%
  mutate(
    month = str_to_lower(str_trim(month)),                     
    month = stringi::stri_trans_general(month, "Latin-ASCII"), 
    month = dplyr::recode(month, !!!month_translation, .default = month),
    month = factor(month, levels = month_order, ordered = TRUE)
  ) %>%  # Translate to English
  group_by(month, specie) %>%  # Make sure your column is 'species'
  summarise(harvest = mean(total_harvest_IFOP_centrosur, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(month = factor(month, levels = month_order)) %>%
  mutate(specie = dplyr::recode(specie,
                         "ANCHOVETA" = "Anchoveta",
                         "SARDINA COMUN" = "Sardine",
                         "JUREL" = "Jack Mackerel")) 


# Plot with facets by species
ggplot(harvest_summary, aes(x = month, y = harvest, fill = specie)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ specie, scales = "free_y", ncol = 1) +
  labs(x = "Month", y = "Average Harvest (tons)", fill = "Specie") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Paired")

rm(harvest_IFOP_month, harvest_summary, month_order, month_translation, months)


```

```{r monthlyharvestSP, eval=FALSE, fig.align='center', fig.cap="Average monthly landings by species (2012-2024; South-Central Chile) ", fig.pos='ht!', message=FALSE, warning=FALSE, include=FALSE}
library(stringi)

harvest_month_SERNAPESCA <- read.csv(
  file.path(dirdata, "SERNAPESCA", "bd_desembarque.csv"),
  fileEncoding = "Latin1",
  sep = ";") %>%
  rename(specie = especie,
         year = año,
         month = mes) %>%
  mutate(zone = case_when(
    region %in% c("Valparaíso", "Metropolitana" , "O'Higgins", "Maule", "Ñuble", "Bio-bío", "La Araucanía", "Los Ríos", "Los Lagos") ~ "Centro_Sur", 
    region %in% c("Arica y Parinacota", "Tarapacá", "Antofagasta", "Atacama", "Coquimbo" ) ~ "Norte",
    region %in% c("Magallanes", "Aysén") ~ "Extremo_Sur",
    TRUE ~ "No_Especifica")) %>% 
  mutate(specie = toupper(stri_trans_general(specie, "Latin-ASCII"))) %>% 
  filter(specie %in% c("ANCHOVETA", "JUREL", "SARDINA COMUN")) %>% 
  filter(zone %in% c("Centro_Sur"))

# %>%
#   pivot_wider(
#     names_from = zone,
#     values_from = total_harvest_SERNAPESCA_v2,
#     names_glue ="
#   ) 

# English month order
month_order <- c("jan", "feb", "mar", "apr", "may", "jun",
                 "jul", "aug", "sep", "oct", "nov", "dec")


harvest_summary2 <- harvest_month_SERNAPESCA %>%
  mutate(
    month = factor(as.integer(month), levels = 1:12, labels = month_order, ordered = TRUE)
  ) %>%
  group_by(month, year, specie) %>%
  summarise(harvestym = sum(toneladas, na.rm = TRUE), .groups = "drop") %>%
  filter(year >= 2013) %>%
  group_by(month, specie) %>%
  summarise(harvest = mean(harvestym, na.rm = TRUE), .groups = "drop") %>%
  arrange(month) %>%
  mutate(month = factor(month, levels = month_order)) %>%
  mutate(specie = dplyr::recode(specie,
                         "ANCHOVETA" = "Anchoveta",
                         "SARDINA COMUN" = "Sardine",
                         "JUREL" = "Jack Mackerel")) 


# Plot with facets by species
ggplot(harvest_summary2, aes(x = month, y = harvest, fill = specie)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ specie, scales = "free_y", ncol = 1) +
  labs(x = "Month", y = "Average Harvest (tons)", fill = "Specie") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Paired")

rm(harvest_month_SERNAPESCA, harvest_summary2, month_order)

```

# Data and methodology

To fulfill the research's objectives, and following @Kasperski2015-jm,
the research entails five different stages: (i) estimating the annual
stock dynamics of each species included in the model, (ii) estimating
trip level cost functions, (iii) estimating total annual trips, (iv)
estimate the inverse demand model for outputs (i.e., price responses to
supply), and (v) conduct numerical optimization to examine how harvest
and profits levels evolve over time. The numeral optimization uses
estimated parameters from the previous four stages to conduct the
optimization procedure.

<!-- Following the bioeconomic literature on regulated fisheries (e.g. Richter et al., 2018), we model a fishery composed of multiple fleet segments that differ in cost structure, harvesting technology, and catch quality, which in turn implies segment-specific output prices (Asche et al., 2015). -->


## Historical data

We use data requested from the Chilean Fisheries Development Institute
(Instituto de Fomento Pesquero, IFOP) covering the 2013–2024 period. The
dataset includes trip-level microdata, which contain detailed records on
vessel identifiers, departure and arrival times, vessel capacity, fleet
and gear type, ports of departure and landing, fishery codes, haul
timing and location, species composition, retained catch, and trip
activity. In addition, we requested annual information on stock
abundance and vessel landings by port, county, region, country, and
species. Finally, we use ex-vessel prices, reported monthly or annually
by port, county, region, country, and species. These prices reflect
those paid by processing plants to fishers at the point of first sale
and are obtained through IFOP’s landing surveys, which do not
necessarily cover all market transactions.

<!-- How different are SERNAPESCA and IFOP harvest data? (Figura -->

<!-- @ref(fig:harvestsource)) -->

```{r harvestsource, eval=FALSE, fig.align='center', fig.cap="Desembarques anuales (IFOP vs SERNAPESCA)", fig.pos='ht!', message=FALSE, warning=FALSE, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)

harvest_long <- harvest %>% 
  dplyr::select(c("specie", "year", 
                  "total_harvest_IFOP_centrosur",
                  "harvest_LANCHAS_IFOP_centrosur",
                  "harvest_BOTES_IFOP_centrosur",
                  "harvest_IND_IFOP_centrosur",
                  "total_harvest_all_SERNAPESCA_Centro_Sur",
                  "total_harvest_sernapesca_v2_centro_sur",
                  "harvest_IFOP_loogbook_centrosur")) %>%
  pivot_longer(cols = c("total_harvest_IFOP_centrosur",
                        "harvest_LANCHAS_IFOP_centrosur",
                        "harvest_BOTES_IFOP_centrosur",
                        "harvest_IND_IFOP_centrosur",
                        "total_harvest_all_SERNAPESCA_Centro_Sur",
                        "total_harvest_sernapesca_v2_centro_sur",
                        "harvest_IFOP_loogbook_centrosur"),
               names_to = "source", 
               values_to = "harvest") %>%
  mutate(source = dplyr::recode(source,
                         "harvest_LANCHAS_IFOP_centrosur" = "Lanchas IFOP",
                         "harvest_BOTES_IFOP_centrosur" = "Botes IFOP",
                         "harvest_IND_IFOP_centrosur" = "Industrial IFOP",
                         "total_harvest_IFOP_centrosur" = "Total IFOP",
                         "total_harvest_all_SERNAPESCA_Centro_Sur" = "Total SERNAPESCA",
                         "total_harvest_sernapesca_v2_centro_sur" = "Total SERNAPESCA v2",
                         "harvest_IFOP_loogbook_centrosur" = "Total IFOP (logbooks)")) %>%
  filter(source %in% c("Total IFOP", "Total SERNAPESCA v2", "Total IFOP (logbooks)")) %>% filter(year >= 2012)

# Gráfico
ggplot(harvest_long, aes(x = year, y = harvest, color = source)) +
  geom_line(size = 1) +
  facet_wrap(~specie, scales = "free_y") +
  labs(x = "Anio", 
       y = "Annual landings (tons)", 
       color = "Source") +
  theme_minimal() +
  theme(legend.position = "bottom") 

rm(harvest_long)

```

For the environmental covariates, we use data from the E.U. Copernicus Marine Service Information, accessed through the Copernicus Marine Toolbox API. Salinity, sea surface temperature, and current speed and direction were obtained from the Global Ocean Physics Reanalysis (GLORYS12V1), which provides data at a 1/12° horizontal resolution with 50 vertical levels [@GLORYS12V1]. Wind speed and direction at the surface were obtained from the Global Ocean Hourly Reprocessed Sea Surface Wind and Stress from Scatterometer and Model dataset, available at 0.125° horizontal spatial resolution and hourly frequency [@WIND_GLO_PHY]. Chlorophyll-a concentrations were obtained from the Global Ocean Colour dataset, which provides data at \~4 km horizontal resolution [@GlobColour]. All environmental data were retrieved daily (hourly in the case of winds) for the 2013–2024 period, covering the Chilean Exclusive Economic Zone (EEZ) between 32°S and 41°S (Figure \@ref(fig:figEnvData))

To address the years in which no acoustic survey was conducted in south-central Chile—or in which insufficient information was available to produce a stock assessment—we imputed missing jack mackerel biomass values using an auxiliary generalized linear model (GLM) with a Gamma distribution and a log link. The model specifies the log of expected biomass in the south-central region as a nonlinear function of (i) northern jack mackerel biomass, (ii) its squared term, (iii) the squared biomass index estimated by the South Pacific Regional Fisheries Management Organisation (OROP-PS) within the exclusive economic zones of Chile, Ecuador, and Peru, and (iv) an interaction between northern biomass and the SPRFMO biomass index. We estimated five alternative specifications and selected the preferred one based on the Akaike Information Criterion and the squared correlation between observed and fitted values—used here as a pseudo-R² measure. The selected model provides an excellent fit (pseudo-R² $\approx$ 0.97). In two years, however, the model produced extreme and biologically implausible predictions; in those cases, we replaced the estimated values with NA to avoid propagating outliers in subsequent analyses.
  

```{r figEnvData, echo=FALSE, fig.align='center', fig.cap="Geographical extent of the study area used for environmental covariates limited to the Chilean Exclusive Economic Zone", message=FALSE, warning=FALSE, fig.pos='ht!', out.width='80%'}

knitr::include_graphics("figs/env_data_map.pdf")

```

### To be requested

-   Diesel cost.
-   Permits by vessels
-   Quota prices?
    -   Auction market but also secondary market if available
        -   If no data, maybe intrapolate prices from other auctions?
    -   Captures elements of forward-looking behavior and information
        [@Birkenbach2024]. @reimer2022structural similarly argue that
        including a quota price captures forward looking behavior and
        allows one to simplify the dynamic model to a static one.
-   Quota by area/fishing organization for Artisanal, and TAC for
    industrial with ITQ (by vessel?)

## Future data for projections

-   OracleBio
    -   Unfurtunally, only decadal (e.g., 2040–2050) projections for
        different scenarios for SST, salinity, currents and chlorophyll
        (4km resolution)
    -   No winds; CMIP6 for winds? (\~100 km).

## Econometrics models


### Stock dynamics

To estimate stock dynamics, we use annual data on species-specific biomass (stock abundance) and vessel landings within Chile’s Exclusive Economic Zone (EEZ). Following @Kasperski2015-jm, the baseline model for the interannual growth of each species $i$ is represented by a discrete logistic function that accounts for intra- and inter-species interactions:

\begin{equation}
x_{i,y+1} + h_{iy} = \underbrace{(1 + r_i)x_{iy} + \eta_i x_{iy}^2}_{R_i(x_{iy})} + \underbrace{\sum_{j \neq i}^{n-1} a_{ij} x_{iy} x_{jy}}_{I_i(x_y)} \quad i=1,\ldots,n
\label{eq1}
\end{equation} 

where $x_{iy}$ is the biomass of species $i$ in year $y$, $n$ is the total number of species, $h_{iy}$ is the annual harvest, $r_i$ is the intrinsic growth rate, $\eta_i$ is the density-dependent parameter, and $\alpha_{ij}$ captures pairwise species interactions. The system of $n$ growth equations is
estimated simultaneously using Seemingly Unrelated Regression (SUR). Following @Richter2018, we augment \eqref{eq1} by including environmental covariates $Env_{iy}$ that affect stock dynamics, along with an error term $\varepsilon_{iy}$ representing stochastic recruitment:

\begin{equation}
x_{i,y+1} + h_{iy} = \underbrace{(1 + r_i)x_{iy} + \eta_i x_{iy}^2}_{R_i(x_{iy})} + \underbrace{\sum_{j \neq i}^{n-1} a_{ij} x_{iy} x_{jy}}_{I_i(x_y)} + \rho_i Env_{iy} + \varepsilon_{iy} \quad i=1,\ldots,n
\label{eq2}
\end{equation} 

where $\rho_i$ denotes the environmental response parameters. Importantly, each species has its own equation, allowing for species-specific growth, density dependence, and environmental sensitivities.

Environmental conditions were summarized annually using sea surface temperature (SST) and chlorophyll-a concentration (CHL), both averaged over the South-Central Chile region within the EEZ. These variables were selected due to their recognized influence on small pelagic productivity and spatial distribution [@Axbard2016]. To capture nonlinear responses, we include both linear and quadratic terms for SST and CHL. Early specifications also tested wind effects, but excluding them did not reduce explanatory power ($\text{F}_{6,21} = 0.919$, $p\text{-value} = 0.509$). 

SST serves as a proxy for large-scale oceanographic regimes (warm vs. cold phases) that shape recruitment success along the Humboldt Current System. Anchoveta typically dominates during cold, nutrient-rich phases, whereas sardine tends to increase under warmer regimes [@cahuin2009; @Yáñez2014]. CHL reflects interannual variation in primary productivity, which forms the energetic base sustaining small pelagics. As emphasized by @sumaila2011, “Changes in primary productivity and planktonic community structure affect the amount of energy transferred to higher trophic levels and, eventually, the productivity of trophic groups that contribute to fisheries catches” [see also @cheung2008; @jennings2008].

Figure \@ref(fig:biomass) shows that anchoveta, sardine, and jack mackerel exhibit distinct biomass trajectories over time, with no evidence of strong interannual co-movement across species. Biomass levels differ markedly between them, and the figure suggests only limited interrelation alongside clear divergences in year-to-year dynamics. Harvest pressure is also visibly associated with these fluctuations; for instance, jack mackerel experienced an abrupt decline during the late 2000s, consistent with the combined effects of intense fishing pressure and unfavorable environmental conditions.

```{r biomass, echo=FALSE, fig.align='center', fig.cap="Estimated biomass of small pelagic species in Chile (2000–2024)", fig.pos='ht!', message=FALSE, warning=FALSE}

library(ggplot2)
library(tidyr)
library(scales)
library(viridis)

biomass_long <- biomass %>%
  pivot_longer(cols = c(sardine_biomass, anchoveta_biomass, jurel_biomass_cs, jurel_cs_interp_primary),
               names_to = "species", 
               values_to = "biomass") %>%
  filter(!is.na(biomass)) %>%
  arrange(species, year)

biomass_long$biomass <- as.numeric(biomass_long$biomass)
biomass_long$year <- as.numeric(biomass_long$year)
biomass_long$species <- dplyr::recode(biomass_long$species,
                               sardine_biomass = "Sardine", 
                               anchoveta_biomass = "Anchoveta", 
                               jurel_biomass_cs = "Jack Mackerel", 
                               jurel_cs_interp_primary = "Jack Mackerel (intrapolated)")

biomass2 <- biomass_long %>% filter(species %in% c("Sardine", "Anchoveta", "Jack Mackerel", "Jack Mackerel (intrapolated)"))

biomass2 <- biomass2 %>% filter(species %in% c("Sardine", "Anchoveta", "Jack Mackerel (intrapolated)")) %>%
  mutate(species = ifelse(species == "Jack Mackerel (intrapolated)", "Jack Mackerel", species))

# Plot
ggplot(biomass2, aes(x = year, y = biomass, color = species, group = species)) +
  geom_smooth(se = TRUE, method = "loess", span = 0.4, linetype = "solid") +
  scale_y_continuous(labels = scales::comma) +  # Or use scale_y_log10(labels = comma) to compress
  labs(x = "Year",
       y = "Biomass (tons)",
       color = "Species") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.y = element_text(angle = 0, hjust = 1),
    legend.position = "right" 
  ) + 
  scale_color_viridis_d(option = "D")

rm(biomass_long, biomass2)

```

The SUR framework is particularly suited to this multi-species setting, as it allows for contemporaneous correlation among residuals that naturally arise from shared non-observed environmental shocks, trophic linkages, and imperfectly observed ecological processes. Cross-species interaction terms are specified between predator (jack mackerel) and prey (anchoveta, sardine) pairs, enabling the model to capture both direct biological coupling and environmental spillovers transmitted across species. We estimate a three-equation SUR system using the SEM framework implemented in lavaan, freely allowing the error terms across biomass equations to be correlated. The model is estimated via robust maximum likelihood (MLR), yielding SUR coefficients with Huber–White robust standard errors.

### Trip level cost functions

Two different estimations (one for industrial, 1 for small scale- both use purse seine...): **"Trip-level costs are modeled as a function of expected effort and segment-specific parameters, reflecting differences in technology and operational costs across fleet segments."**

Just usea real SPF trips (consider 95% of total revenue...)
DO I HAVE DATA IN CREW SIZE???? Opportunity cost... (paid by time at sea x hourly wage) -- MAYBE USE TIPO EMBARCACION TO LINK CREW SIZE!??? USE BODEGA SIZE TO LINK CREW SIZE???
Fixed cost excluded. Should not affect as econ decision made at the margins (optimization behavior)
hard to use share of revenue, as how they implement this depend on each operator

<!-- - 80 and 300 m$^3$ holds sardine and anchovy; More than 800 m$^3$, likely jack mackerel. @Pena-Torres2017-gn dummy for $H \leq 370 \ \text{m}^3$, $370 \ \text{m}^3 < H \leq 790 \ \text{m}^3$, $H > 790 \ \text{m}^3$ ... Also use age of vessel to check technical obsolescence (year trip, year ship was built) -->

-   @sumaila2011: "Capital costs, that is, the cost of vessels, fishing
    gear, processing plants and so on, would be affected by climate
    change if additional capital for fishing and processing operations
    is required to adapt to climate change impacts on the quantity,
    composition and distribution of fisheries resources65. Changes in
    migratory routes and fish distribution would affect travel time,
    which can lead to increases or decreases in fuel and ice cost
    depending on catch levels and patterns, and the management regime in
    place."

We estimate restricted cost function for both small-scale and industrical vessel. The inputs that we assume fixed are vessel characteristics and the population of each species. 

Ignoring trip subscript, the quadratic approximation of the cost functions vary by vessel
$v=1,\ldots,V_g$ and gear used $g=1,\ldots,G$, where $V_g$ is the number
of observations by fleet $g$, and $G$ is the total number of
fleets: \begin{equation}
C_{vg} = \sum_{i=1}^{2n+M+k} \alpha_{g, \mathbf{X}_i} \mathbf{X}_{ivg} + \frac{1}{2} \sum_{i=1}^{2n+M+k} \sum_{j=1}^{2n+M+k} \alpha_{g, \mathbf{X}_i\mathbf{X}_j} \mathbf{X}_{ivg} \mathbf{X}_{jvg}
\label{eq3}
\end{equation} where $C_{vg}= w z_{vg}^*$,
$\mathbf{X}_{vg}=[w;h_{vg};x;Z_v]$, $w$ is a $V_g \times M$ matrix of
variable input prices, $h_{vg}$ is an $V_g \times n$ matrix of harvest
quantities, $x$ is an $V_g \times n$ matrix of given stock levels of the
species of interest, and $Z_v$ is an $V_g \times k$ matrix of given
vessel characteristics. Therefore, $\mathbf{X}_{vg}$ is a
$V_g \times (2n+M+k)$ matrix, and $\mathbf{X}_{ivg}$ represents the
*i*th column of the $\mathbf{X}_{vg}$ matrix.

Together with estimating the restricted cost function, we estimate the
conditional input demand equations. This addition allows an increase in
the degrees of freedom by imposing cross-equation parameter constraints
and allows for the testing of, for instance, jointness in inputs
[@Kasperski2015-jm]. The conditional input demand equations are derived
by Shepard's Lemma: \begin{equation}
\frac{\partial C_{vg}}{\partial w_m} = z^*_{vg,w_m} = \alpha_{g,w_m} + \sum_{j=1}^{2n+M+k} \alpha_{g,w_m,\mathbf{X}_j} \mathbf{X}_{jvg} \quad m=1,\ldots,M.
\label{eq4}
\end{equation}

Similar to stock dynamics, the system of equations formed by \eqref{eq3}
and \eqref{eq4} can be estimated using SUR. To comply with economic
theory, and 

To ensure that the estimated cost function is consistent with microeconomic theory, we impose the standard regularity conditions of symmetry and linear homogeneity in input prices.


1.  Symmetry of the cost function, where \begin{equation*}
      \alpha_{g,\mathbf{X}_i\mathbf{X}_j} = \alpha_{g,\mathbf{X}_j\mathbf{X}_i} \quad \forall \quad i=1,\ldots,(2n + M + k); \ i \neq j; \ g = 1,\ldots, G.
      \end{equation*}

2.  Linear homogeneity in input prices, where \begin{equation*}
      \sum_m^M \alpha_{g,w_m} = 1 \ \text{and} \ \sum_m^M \alpha_{g,w_m,\mathbf{X}_i} = 0 \quad i=1,\ldots,(2n + M + k); \ g = 1,\ldots, G.
      \end{equation*}
      
These restrictions reduce even more the number of parameters to estimate in \eqref{eq4}. Symmetry guarantees that the function is twice differentiable with consistent mixed partial derivatives, allowing it to be interpreted as a valid dual cost function. Linear homogeneity (adding-up) ensures that the cost function is homogeneous of degree one in input prices, so that proportional changes in all prices translate into proportional changes in total cost. These restrictions ensure theoretical coherence and prevent the model from being a purely statistical approximation.

A potential concern is the endogeneity between harvest and costs. However, following Kasperski (2015), we assume that vessel optimization occurs in two stages. In the first stage, vessels choose the cost-minimizing combination of inputs for any given vector of outputs. In the second stage, they choose their optimal output mix to maximize profits. Under this sequential decision process, outputs are treated as exogenous when estimating the restricted cost function, because they are taken as given in the underlying cost-minimization problem.

Data at the trip level is available upon request from the Chilean Fisheries Research Institute (IFOP), which registers geo-referenced catch information on the Chilean fleet's fishing operation per trip [see e.g. @Pena-Torres2017-gn and [IFOP data observatory](ifop.dataobservatory.net)]. As inputs we can use the time spent at sea during a trip, where the price is the average wage pay to crew member per hour, and the distance traveled during a trip, where the price of distance traveled is the diesel cost. Therefore, the total cost function $C_{vg}= w z_{vg}^*$ for vessel $v$, using gear $g$ in a trip would be sum of the total cost of distance travelled plus the total cost of the time spent at sea.

**Note:** *Depending on the type of vessel, this cost should change.
Some vessels are more efficient, other one are more heavy. How to
capture this? The righ hand side has vessel characteristics, so the
effect of harvest would be conditional on vessel characteristics, the
stock levels and input prices. As we only care in the margin how harvest
increase cost, this should be fine. @Kasperski2015-jm mention this
"...no reliable fixed cost information on these vessels exists, but
these should not affect the optimization as economic decisions are made
at the margin. Therefore, this study does not measure true profit, but
rather a proxy based on the net operating rent accruing to vessels in
the fishery."*

To link this function to climate change, we can also include additional
environmental variables $Env$ to $\mathbf{X}_{vg}$ such as wind
intensity and wave conditions in each trip at the harvest location, upon
data availability. Therefore, the augmented $X_{vg}$ matrix becomes
$\mathbf{X}^{'}_{vg} = [w;h_{vg};x;Z_v;Env]$.

MAYBE INCLUDE QUOTA PRICE? **Higher quota prices for depleted stocks
(e.g., GOM cod) reduce incentives to target them.** and **Active leasing
markets for quota (and previously for DAS) allow fishermen to treat
quota as a "priced input" rather than a fixed, exhaustible resource.**

Add - liters/hours x hours trip x fuel price - crew member x harvest x
price x share - certification cost per landed ton x landing


### Total annual trips

We model the annual number of fishing trips taken by vessel $v$ in year $y$ as a count process following @Kasperski2015-jm. Since the available logbook data only record purse-seine operations, the unit of observation is a vessel–year. Separate effort models are estimated for the industrial and artisanal fleets, allowing fishing activity to respond differently to economic conditions and regulatory constraints across sectors. This specification explicitly accommodates technological heterogeneity across fleet segments, reflecting differences in vessel capacity, operating scale, and production technology.

\begin{equation}
T_{vy} \sim \text{Poisson}(\lambda_{vy}), \qquad 
\lambda_{vy} = \exp\!\left(U_{vy}'\beta\right),
\label{eq:poisson_trips}
\end{equation}

where $T_{vy}$ denotes the total number of purse-seine trips recorded for vessel $v$ in year $y$. The vector of explanatory variables $U_{vy}$ includes output prices by species, input prices, vessel-level allocated harvest, aggregate total allowable catches (TACs), fixed vessel characteristics, and operating conditions:

\begin{equation}
U_{vy}=\big[p_{sy},\, w_y,\, H^{alloc}_{vy},\, \bar{Q}_y,\, Z_v,\, Env_y\big].
\label{eq:U_trips}
\end{equation}

Output prices $p_{sy}$ correspond to species-specific processed fish prices, while $w_y$ captures key variable input costs, including diesel prices. Consistent with @dresdner2013, prices are measured as annual averages over peak fishing months to reflect economically relevant conditions faced by vessels when planning annual effort.

Because the objective of the model is to characterize vessels’ effort responses to changes in prices, quotas, and environmental conditions within the small pelagic fishery, we restrict attention to SPF trips recorded in logbooks. These trips account for approximately 95\% of observed fishing revenue and therefore capture the primary economic margin through which vessels adjust annual fishing effort.

Quota shares do not enter Eq.~\eqref{eq:poisson_trips} directly. Instead, shares are used to translate annual TACs into vessel-level allocated harvests, which then enter the trip equation. Let $\omega_{vs}$ denote vessel $v$'s historical share of landings of species $s$ within the purse-seine fleet. Given a candidate TAC $\bar{Q}_{sy}$ for species $s$ in year $y$, vessel-level allocated harvest is defined as:

\begin{equation}
H^{alloc}_{vy,s}=\omega_{vs}\,\bar{Q}_{sy},
\qquad
H^{alloc}_{vy}=\sum_{s} H^{alloc}_{vy,s}.
\label{eq:Halloc}
\end{equation}

This construction is essential for the forward-looking simulation framework. When the social planner chooses aggregate TAC paths $\{\bar{Q}_{sy}\}$, Eq. \eqref{eq:Halloc} generates internally consistent vessel-level harvest allocations that can be combined with predicted trips and trip-level cost functions to compute total harvesting costs.

The vector $Env_y$ captures operating and regulatory constraints, including the number of days closed to fishing due to seasonal bans (vedas), as well as environmental conditions that affect the feasibility of fishing operations, such as the number of bad weather days.

Quota prices are not included explicitly in the trip equation. Instead, the economic scarcity of quota is captured implicitly through aggregate TAC levels and vessel-level allocated harvest. In the simulation stage, quota scarcity generates shadow values that affect optimal harvest and effort decisions through profits rather than through an explicit quota price term.

Trips, harvest, and prices may be jointly determined within the year, implying potential endogeneity in reduced-form trip regressions. However, the purpose of Eq. \eqref{eq:poisson_trips} is not causal identification but rather to provide an empirically grounded behavioral relationship that maps changes in prices, quotas, and environmental conditions into expected fishing effort for use in the bioeconomic simulation.



### Inverse demand model

The price of each species is modeled using an Inverse Almost Ideal
Demand System (IADS). The price of a species $i$ in year $y$ is the
following:

\begin{equation}
\ln p_{iy} = \sum_j^n \gamma_j \ln h_{j,y} + \gamma_{H} \ln H + \gamma_{FM} \ln P^\text{FishMeal} + \epsilon_{iy}, \quad i = 1,\ldots,n, \ j = 1,\ldots,n.
\label{eq6}
\end{equation}

Note that harvest may be endogenous in this system due to simultaneity.
For this reason, the system formed by \eqref{eq6} will be estimated
using a Three Stage Least Squares (3SLS) procedure, where h_j{j,y} would
be instrumented by variables that affect the fishers' supply function
such as SST, Chl, and fuel prices.

```{=html}
<!-- Comment: One concern I have regards the endogeneity of factors determining the outcomes of fishing effort on a fishing
trip (where, when, how long, and for what, and how much), climate conditions (temperature, wave, and wind),
and input and output prices. While Kasperski assumes that total allowable catches are binding to avoid one potential endogeneity, I wonder if this is a good assumption under climate change. What would you need to assume about the regulator’s response to the changing conditions and the setting of the total allowable catches? -->
```

```{=html}
<!-- No era que el paper estimaba un IAIDS. Sino que al precio ex-vessel a ser un % del FOB justifica que sea una variable exógena para la IAIDS que nosotros estimaremos. Así capturar las desviaciones del mercado internacional se transmitan directamente al precio ex-vessel. El otro aspecto era el siguiente: En las estimaciones MCO para la VIII Región, el Precio de la Harina de Pescado * Dólar presenta un signo negativo y significativo sobre el número agregado de viajes anuales, aunque en la XIV Región el signo es positivo [11 (254, 398). Lo que podría implicar que la reaccion de la flota al FOB es heterogénea por región  
R:  Perfecto. Entonces no estimaron un modelo de precios. Mejor aún, porque entonces la contribución de su trabajo sigue siendo alta. Solo usan FOB como exógena, por lo tanto, nuestra metodología sigue siendo la misma, de estimar sin incluir FOB, luego incluyéndolo, ver qué pasa con el coeficiente de Q, para ver si hay algo más en la determinación de precios a corto plazo. Luego, si hay tiempo, ver si de hecho el precio FOB depende de desembarques chilenos. En cuanto a la heterogeneidad, a mí me da la impresión de que ese texto de hecho está indicando esta endogeneidad que estamos tratando de capturar! Es posible que la XIV no tenga efecto en el FOB, por lo que lo que vemos es un efecto insesgado para esa región, pero que en la VIII, que es una productora grande de pequeños pelágicos, sí tenga un efecto en el precio FOB, por lo tanto, el coeficiente negativo que obtienen de hecho está sesgado por endogeneidad!! Yo usaría ese texto como justificación de su estudio... -->
```

-   DIFERENCIAR ENTRE CONSUMO HUMANO DIRECTO e INDIRECTO? Ricardo y Arnol usan solo información de procesamiento. Darle una vuelta a esto
- NECESITO MODELO ANUAL??? SIMPLE CON REZAGOS DE CANTIDAD??? O MAS SIMPLE SIGUIENDO A KASPERSKI PERO INSTRUMENTANDO LA CANTIDAD???


- SEPARAR MODELO POR FLOTA IND Y ART???
  - "On the demand side, we assume an inverse demand structure where the output price depends on the segment that has landed the catch because of differences in quality, but also on the total catch of all boats, due to an adverse effect of aggregate landings on the sales price."





## Numerical optimization

To obtain the effect of future climate variability on stock, harvest,
quota and profits, we conduct numerical optimization for different
climate scenarios using the parameters estimates for the stock dynamic,
cost functions, total annual trips and inverse demand equations. In each
year, a vessel maximizes profits by choosing their optimal number of
trips $T$ and optimal portfolio of harvest on each trip $h_{\tau}$:
\begin{align}
\max_{h_{gt}, T_g} \quad \pi_{vgt} & = \sum_{\tau=t}^{T_g} \rho^\tau \left\{ P(h) h_{g\tau} - C_g(h_{g\tau} | w, x, Z, Env) \right\} \quad \tau = t,\ldots, T_g \nonumber \\
\textbf{s.t} \quad q_{g,t+1} & = \omega \ast \bar{q} - \sum_{t=1}^{t} h_{gt} \geq 0, \quad t = 1, \dots, T-1, \quad g = 1, \dots, G
\label{eq7}
\end{align} where $\rho$ is the intra-annual discount factor, $\omega$
is a vector of shares of $\bar{q}$, and $h_{lt}=0$ for all $l\neq g$.
The vector of shares is obtained from historical data on harvest. The
optimal profit from the maximization problem in \eqref{eq7} is denoted
as $\pi_{vgy}^* (p,w,x,Z,\bar{q},\omega, Env)$, and $h_{vgty}^*$ and
$T_{vgy}^*$ are the optimal choices harvest per trip and total number of
trips in year $y$ for vessel $v$. To obtain the optimal quota level, we
must solve the social-planner optimization problem to maximize the net
value of the fishery by choosing the quota levels per year and by
species.

Following @Kasperski2015-jm, the optimization problem will be conducted
for the next 25 years. I will use different climate scenarios and
compare different optimal outcomes between them by using future
projections for the environmental variables included in the model.

-   Note that the optimal quota would that we would find using
    historical data might differ from the actual quota set by the
    regulator as they model is for single species.
  
- We assume that quota is not tradable, but allocated individually to each vessel. This avoid to consider other vessle participation decision on vessel yeraly maximization process [@Kasperski2015-jm].





## Projections


According @Kasperski2015-jm, the model presented in this project is more suitable for long-term projections of fish population and profits than intra-anuial impacts of short-term management action. Therefore, is suitable to studyn long term effect of climate variability. However, based on Aufhammer, because we are using short-term variability instead of decadal vsariability, then our model is onlpy able to capture the impact of climate without adaptation.


*From [@Yáñez2014]*:

-   "To project the model, the average structure of catches and
    temperature (of Antofagasta and the region Niño 3 + 4) for the years
    2005, 2006 and 2007 were used as starting point. We consider a
    linear increase in temperature, taking into account four climate
    change scenarios based on the scenarios presented by IPCC, designed
    for the northern part of Chile until 2100.
    1.  The first scenario considers an increase in temperature of
        0.034ºC per year [@Fuenzalida2007], similar to that estimated by
        @Trenberth2007.
    2.  A second scenario, more moderate, of 0,025ºC/year is also
        proposed by @Fuenzalida2007.
    3.  The third scenario is not considered a significant effect on the
        area, following the work of @Trenberth2007.
    4.  The fourth scenario is contradictory, indicating a cooling de
        0.02 ºC/year [@falvey2009]. It should be noted that according to
        the work of @Fuenzalida2007 and @falvey2009 the same SST
        increase (or decrease) were considered for both temperatures (in
        Antofagasta and in the Niño 3 + 4 region)."

# Results


## Stock biomass model

<!-- Figure \@ref(fig:harvestbiomass) shows biomass and harvest. -->

```{r harvestbiomass, fig.align='center', fig.cap="Estimated biomass vs harvest of small pelagic species in Central-South Chile (2000–2024)", fig.pos='ht!', message=FALSE, warning=FALSE, include=FALSE}

library(tidyverse)

harvest2 <- harvest %>%
  select(specie, year, total_harvest_sernapesca_v2_centro_sur) %>%
  mutate(specie = dplyr::recode(specie,
    "SARDINA COMUN" = "Sardine",
    "JUREL"         = "Jack Mackerel",
    "ANCHOVETA"     = "Anchoveta"
  )) %>%
  rename(harvest = total_harvest_sernapesca_v2_centro_sur,
         species = specie)

# Build biomass long from the new biomass_dt.rds structure
biomass_long <- biomass %>%
  select(year, sardine_biomass, anchoveta_biomass,
         jurel_cs_interp_primary) %>%
  pivot_longer(
    cols = c(sardine_biomass, anchoveta_biomass, jurel_cs_interp_primary),
    names_to = "species", values_to = "biomass"
  ) %>%
  mutate(species = dplyr::recode(species,
    sardine_biomass         = "Sardine",
    anchoveta_biomass       = "Anchoveta",
    jurel_cs_interp_primary = "Jack Mackerel"
  )) %>%
  filter(!is.na(biomass))

biomass_harvest <- full_join(biomass_long, harvest2,
                             by = c("species", "year")) %>%
  arrange(species, year)

biomass_harvest_long <- biomass_harvest %>%
  pivot_longer(cols = c(biomass, harvest),
               names_to = "variable", values_to = "value")

ggplot(biomass_harvest_long, aes(x = year, y = value, color = variable)) +
  geom_line(linewidth = 1, na.rm = TRUE) +
  facet_wrap(~species, scales = "free_y") +
  labs(x = "Year", y = "Value (tons)", color = "Variable") +
  scale_color_manual(values = c("biomass" = "darkgreen", "harvest" = "steelblue")) +
  theme_minimal() +
  theme(legend.position = "top")

rm(biomass_harvest_long, harvest2, biomass_long)

```

<!-- ============================================================ -->
<!-- BIOMASS ESTIMATION: DATA PREPARATION                          -->
<!-- ============================================================ -->

```{r database_for_biomass_est, message=FALSE, warning=FALSE, include=FALSE}

library(tidyverse)
library(janitor)

# ---- 1. Build wide biomass + harvest dataset ----

biomass_wide <- biomass %>%
  select(year, sardine_biomass, anchoveta_biomass,
         jurel_biomass_cs, jurel_cs_interp_primary, jurel_cs_source) %>%
  mutate(
    jurel_main = jurel_cs_interp_primary,
    jurel_obs  = jurel_biomass_cs
  )

# Harvest data (wide)
harvest_wide <- harvest %>%
  select(specie, year, total_harvest_sernapesca_v2_centro_sur) %>%
  pivot_wider(names_from = specie,
              values_from = total_harvest_sernapesca_v2_centro_sur,
              names_prefix = "h_") %>%
  janitor::clean_names()

biomass_harvest_wide <- left_join(biomass_wide, harvest_wide, by = "year")

# ---- 2. Build dependent variable: biomass_{t+1} + harvest_t ----

biomass_harvest_wide <- biomass_harvest_wide %>%
  arrange(year) %>%
  mutate(
    sardine_t1      = lead(sardine_biomass),
    anchoveta_t1    = lead(anchoveta_biomass),
    jurel_main_t1   = lead(jurel_main),
    jurel_obs_t1    = lead(jurel_obs),
    y_sardine   = sardine_t1   + h_sardina_comun,
    y_anchoveta = anchoveta_t1 + h_anchoveta,
    y_jurel     = jurel_main_t1 + h_jurel,
    y_jurel_obs = jurel_obs_t1  + h_jurel
  )

# ---- 3. Environmental data (annual averages) ----

env_dt_year1 <- env_dt %>%
  mutate(year = year(date)) %>%
  group_by(year) %>%
  summarize(sst  = mean(sst, na.rm = TRUE),
            chl  = mean(chl, na.rm = TRUE),
            wind = mean(speed_max, na.rm = TRUE),
            .groups = "drop")

env_dt_year2 <- env_dt_00_11 %>%
  mutate(year = year(date)) %>%
  group_by(year) %>%
  summarize(sst  = mean(sst, na.rm = TRUE),
            chl  = mean(chl, na.rm = TRUE),
            wind = mean(speed_max, na.rm = TRUE),
            .groups = "drop")

env_dt_year <- bind_rows(env_dt_year2, env_dt_year1)

biomass_harvest_wide <- left_join(biomass_harvest_wide, env_dt_year, by = "year")

rm(env_dt_year, env_dt_year2, env_dt_year1, biomass_wide, harvest_wide)

```

```{r sur_data, message=FALSE, warning=FALSE, include=FALSE}

# ---- 4. Build SUR estimation datasets ----

scale_b <- 1e5

# --- Main spec: 3-species with interpolated jurel (N ~ 23) ---
sur_main <- biomass_harvest_wide %>%
  filter(!is.na(y_sardine), !is.na(y_anchoveta), !is.na(y_jurel),
         !is.na(sardine_biomass), !is.na(anchoveta_biomass), !is.na(jurel_main),
         !is.na(sst), !is.na(chl)) %>%
  mutate(
    y_s = y_sardine   / scale_b,
    y_a = y_anchoveta / scale_b,
    y_j = y_jurel     / scale_b,
    b_s = sardine_biomass   / scale_b,
    b_a = anchoveta_biomass / scale_b,
    b_j = jurel_main        / scale_b,
    b_s_c = b_s - mean(b_s),
    b_a_c = b_a - mean(b_a),
    b_j_c = b_j - mean(b_j),
    b_s_c2 = b_s_c^2,
    b_a_c2 = b_a_c^2,
    b_j_c2 = b_j_c^2,
    sa = b_s_c * b_a_c,
    sj = b_s_c * b_j_c,
    aj = b_a_c * b_j_c,
    sst_c  = sst - mean(sst),
    chl_c  = chl - mean(chl),
    sst_c2 = sst_c^2,
    chl_c2 = chl_c^2
  )

cat("Main spec (interpolated jurel): N =", nrow(sur_main), "\n")

# --- Robustness 1: 3-species with observed jurel only (N ~ 12) ---
sur_obs <- biomass_harvest_wide %>%
  filter(!is.na(y_sardine), !is.na(y_anchoveta), !is.na(y_jurel_obs),
         !is.na(sardine_biomass), !is.na(anchoveta_biomass), !is.na(jurel_obs),
         !is.na(sst), !is.na(chl)) %>%
  mutate(
    y_s = y_sardine    / scale_b,
    y_a = y_anchoveta  / scale_b,
    y_j = y_jurel_obs  / scale_b,
    b_s = sardine_biomass   / scale_b,
    b_a = anchoveta_biomass / scale_b,
    b_j = jurel_obs         / scale_b,
    b_s_c = b_s - mean(b_s), b_a_c = b_a - mean(b_a), b_j_c = b_j - mean(b_j),
    b_s_c2 = b_s_c^2, b_a_c2 = b_a_c^2, b_j_c2 = b_j_c^2,
    sa = b_s_c * b_a_c, sj = b_s_c * b_j_c, aj = b_a_c * b_j_c,
    sst_c = sst - mean(sst), chl_c = chl - mean(chl),
    sst_c2 = sst_c^2, chl_c2 = chl_c^2
  )

cat("Robustness 1 (observed jurel only): N =", nrow(sur_obs), "\n")

# --- Robustness 2: 2-species SUR, jurel as exogenous (N ~ 23) ---
sur_2sp <- biomass_harvest_wide %>%
  filter(!is.na(y_sardine), !is.na(y_anchoveta),
         !is.na(sardine_biomass), !is.na(anchoveta_biomass),
         !is.na(sst), !is.na(chl)) %>%
  mutate(
    y_s = y_sardine   / scale_b,
    y_a = y_anchoveta / scale_b,
    b_s = sardine_biomass   / scale_b,
    b_a = anchoveta_biomass / scale_b,
    b_j_exog = coalesce(jurel_main / scale_b, 0),
    b_s_c = b_s - mean(b_s), b_a_c = b_a - mean(b_a),
    b_j_exog_c = b_j_exog - mean(b_j_exog),
    b_s_c2 = b_s_c^2, b_a_c2 = b_a_c^2,
    sa = b_s_c * b_a_c,
    sst_c = sst - mean(sst), chl_c = chl - mean(chl),
    sst_c2 = sst_c^2, chl_c2 = chl_c^2
  )

cat("Robustness 2 (2-species, jurel exogenous): N =", nrow(sur_2sp), "\n")

```

<!-- ============================================================ -->
<!-- SUR ESTIMATION                                                -->
<!-- ============================================================ -->

```{r sur_estimation, message=FALSE, warning=FALSE, include=FALSE}

library(lavaan)

# ==================================================================
# MAIN SPECIFICATION: Parsimonious (no interactions, no CHL^2)
# 6 params per equation, ~17 df with N=23
# Chosen based on AIC comparison: this spec has lowest AIC
# ==================================================================

model_main <- '
  y_s ~ 1 + b_s_c + b_s_c2 + sst_c + sst_c2 + chl_c
  y_a ~ 1 + b_a_c + b_a_c2 + sst_c + sst_c2 + chl_c
  y_j ~ 1 + b_j_c + b_j_c2 + sst_c + sst_c2 + chl_c
'

fit_main <- sem(model_main, data = sur_main, estimator = "MLR")

# ==================================================================
# ALTERNATIVE SPECS FOR MODEL SELECTION
# ==================================================================

# --- With cross-species interactions ---
model_interactions <- '
  y_s ~ 1 + b_s_c + b_s_c2 + sst_c + sst_c2 + chl_c + sa + sj
  y_a ~ 1 + b_a_c + b_a_c2 + sst_c + sst_c2 + chl_c + sa + aj
  y_j ~ 1 + b_j_c + b_j_c2 + sst_c + sst_c2 + chl_c + sj + aj
'

fit_interactions <- sem(model_interactions, data = sur_main, estimator = "MLR")

# --- With CHL^2 ---
model_chl2 <- '
  y_s ~ 1 + b_s_c + b_s_c2 + sst_c + sst_c2 + chl_c + chl_c2
  y_a ~ 1 + b_a_c + b_a_c2 + sst_c + sst_c2 + chl_c + chl_c2
  y_j ~ 1 + b_j_c + b_j_c2 + sst_c + sst_c2 + chl_c + chl_c2
'

fit_chl2 <- sem(model_chl2, data = sur_main, estimator = "MLR")

# --- No environmental variables ---
model_no_env <- '
  y_s ~ 1 + b_s_c + b_s_c2
  y_a ~ 1 + b_a_c + b_a_c2
  y_j ~ 1 + b_j_c + b_j_c2
'

fit_no_env <- sem(model_no_env, data = sur_main, estimator = "MLR")

# --- Full kitchen sink (interactions + CHL^2) ---
model_full <- '
  y_s ~ 1 + b_s_c + b_s_c2 + sst_c + sst_c2 + chl_c + chl_c2 + sa + sj
  y_a ~ 1 + b_a_c + b_a_c2 + sst_c + sst_c2 + chl_c + chl_c2 + sa + aj
  y_j ~ 1 + b_j_c + b_j_c2 + sst_c + sst_c2 + chl_c + chl_c2 + sj + aj
'

fit_full <- sem(model_full, data = sur_main, estimator = "MLR")

# ==================================================================
# MODEL SELECTION TABLE (AIC / BIC)
# ==================================================================

model_sel <- tibble(
  Specification = c(
    "No environment",
    "Main (SST + SST² + CHL)",
    "Main + CHL²",
    "Main + interactions",
    "Full (interactions + CHL²)"
  ),
  K_per_eq = c(3, 6, 7, 8, 10),
  AIC = c(AIC(fit_no_env), AIC(fit_main), AIC(fit_chl2),
          AIC(fit_interactions), AIC(fit_full)),
  BIC = c(BIC(fit_no_env), BIC(fit_main), BIC(fit_chl2),
          BIC(fit_interactions), BIC(fit_full))
) %>%
  mutate(delta_AIC = AIC - min(AIC),
         delta_BIC = BIC - min(BIC))

cat("\n====== MODEL SELECTION ======\n")

model_sel <- model_sel %>%
  mutate(Specification = c(
    "No environment",
    "Main (SST + SST2 + CHL)",
    "Main + CHL2",
    "Main + interactions",
    "Full (interactions + CHL2)"
  ))

print(model_sel, width = 120)

```

Table \ref{tab:model_selection} summarizes the model selection exercise. Five candidate specifications were compared, ranging from a baseline with only own-biomass terms to a full model including quadratic chlorophyll-a and cross-species interaction terms. The preferred specification---which includes own biomass, its square, SST, SST$^2$, and chlorophyll-a---yields the lowest AIC and BIC. Adding quadratic CHL terms or cross-species interactions does not improve fit; in fact, both AIC and BIC increase when these terms are included. Wind variables were excluded in an earlier specification search, as a joint exclusion test failed to reject their omission. The preferred specification contains six parameters per equation, yielding approximately 17 residual degrees of freedom with $N=23$ observations.

Table \ref{tab:SUR_results} reports the SUR estimates of annual biomass dynamics for sardine, anchoveta, and jack mackerel. The system is estimated via robust maximum likelihood (MLR), yielding Huber--White standard errors and allowing unrestricted contemporaneous correlation across equations.

All three species exhibit strong interannual persistence and significant density dependence, consistent with discrete logistic growth. However, the species differ markedly in their sensitivity to environmental conditions.

Sardine biomass displays the strongest environmental response. Chlorophyll-a enters positively and is highly significant ($\hat{\rho}_{CHL} = 80.99$, $p < 0.001$), indicating that years with higher primary productivity favor sardine growth. The quadratic SST term is also significant and positive ($\hat{\rho}_{SST^2} = 54.74$, $p = 0.001$), suggesting a convex response to temperature anomalies---sardine biomass increases more sharply under larger SST deviations from the mean. This pattern is consistent with evidence from the Humboldt Current System, where sardine tends to exhibit smoother population changes associated with warmer or transitional oceanographic regimes [@cahuin2009; @Yáñez2014].

Anchoveta biomass is dominated by density dependence, with the strongest own-biomass coefficient ($\hat{r} = 1.11$, $p < 0.001$) and a highly significant quadratic term ($\hat{\eta} = -0.202$, $p < 0.001$). SST has a marginally significant negative effect ($p = 0.087$), but neither SST$^2$ nor CHL are significant. This asymmetric environmental response---where anchoveta reacts primarily to temperature rather than productivity---is consistent with the ecological literature. Anchoveta in the Humboldt Current is known to respond sharply to short-term cold, nutrient-rich conditions, dominating during cold phases of decadal-scale oceanographic variability [@cahuin2009; @Yáñez2014; @Alheit2004].

Jack mackerel shows strong persistence ($\hat{r} = 1.06$, $p < 0.001$) but a distinct environmental signature. The quadratic SST term is large and highly significant ($\hat{\rho}_{SST^2} = -56.49$, $p = 0.001$), indicating a clear thermal optimum: biomass declines when temperatures deviate substantially from the long-run mean in either direction. Neither linear SST nor CHL are individually significant, suggesting that jack mackerel growth is less sensitive to annual fluctuations in upwelling productivity than sardine. This weaker coupling to coastal environmental variability is consistent with jack mackerel's broader spatial distribution, deeper foraging behavior, and transboundary stock structure within the Southeast Pacific [@Pena-Torres2017-gn].

These contrasting environmental responses have direct implications for climate projections. Under warming scenarios, sardine may benefit from increased thermal variability if accompanied by sustained productivity, whereas anchoveta could face reduced biomass if warming erodes the cold, nutrient-rich conditions it depends on. Jack mackerel, with its thermal optimum response, would be adversely affected by both warming and cooling extremes.

Cross-species interactions were tested but did not improve model fit (Table \ref{tab:model_selection}). The residual covariances across equations are also statistically insignificant, suggesting that shared unobserved shocks---beyond those captured by SST and CHL---do not generate strong contemporaneous co-movement across species. This does not rule out ecological interactions (e.g., predator--prey dynamics between jack mackerel and small pelagics), but it indicates that such effects are not detectable in the annual aggregate biomass data given the available sample size. The robustness check that includes interaction terms (Table \ref{tab:SUR_interactions}) confirms that their coefficients are individually and jointly insignificant.

We conduct three robustness checks. Table \ref{tab:SUR_rob_obs} restricts the jack mackerel series to observed (non-interpolated) years, reducing the sample to `r nrow(sur_obs)` observations and requiring a more parsimonious specification without SST$^2$. Table \ref{tab:SUR_rob_2sp} estimates a two-species SUR for sardine and anchoveta only, treating jack mackerel biomass as an exogenous covariate, which increases the sample to `r nrow(sur_2sp)` observations. Table \ref{tab:SUR_interactions} augments the main specification with cross-species interaction terms. The core findings---strong density dependence, positive CHL effects on sardine, and a thermal optimum for jack mackerel---are qualitatively robust across all specifications, lending confidence that the results are not driven by the interpolation procedure.


```{r sur_robustness_specs, message=FALSE, warning=FALSE, include=FALSE}

# ==================================================================
# ROBUSTNESS 1: Observed jurel only (N ~ 12)
# Even more parsimonious: drop SST^2 to preserve df
# 4 params per eq, ~8 df
# ==================================================================

model_rob1 <- '
  y_s ~ 1 + b_s_c + b_s_c2 + sst_c + chl_c
  y_a ~ 1 + b_a_c + b_a_c2 + sst_c + chl_c
  y_j ~ 1 + b_j_c + b_j_c2 + sst_c + chl_c
'

fit_rob1 <- sem(model_rob1, data = sur_obs, estimator = "MLR")

# ==================================================================
# ROBUSTNESS 2: Two-species SUR, jurel as exogenous (N ~ 23)
# Same env spec as main, jurel biomass as additional covariate
# ==================================================================

model_rob2 <- '
  y_s ~ 1 + b_s_c + b_s_c2 + sst_c + sst_c2 + chl_c + b_j_exog_c
  y_a ~ 1 + b_a_c + b_a_c2 + sst_c + sst_c2 + chl_c + b_j_exog_c
'

fit_rob2 <- sem(model_rob2, data = sur_2sp, estimator = "MLR")

# ==================================================================
# ROBUSTNESS 3: Main spec with interactions (to show they don't matter)
# ==================================================================

fit_rob3 <- fit_interactions  # already estimated above

```

<!-- ============================================================ -->
<!-- BUILD TABLES                                                  -->
<!-- ============================================================ -->

```{r build_table_helper, message=FALSE, warning=FALSE, include=FALSE}
library(lavaan)
library(dplyr)
library(stargazer)

# Helper: extract lavaan estimates and build patched lm objects for stargazer

build_stargazer_inputs <- function(fit_sem, data, eq_names, dep_vars) {
  pe <- parameterEstimates(fit_sem, standardized = FALSE)
  pe_reg <- pe %>% filter(op == "~")
  pe_int <- pe %>% filter(op == "~1")

  results <- list()

  for (i in seq_along(eq_names)) {
    yname <- eq_names[i]
    dep   <- dep_vars[i]

    reg_i <- pe_reg %>% filter(lhs == yname)
    int_i <- pe_int %>% filter(lhs == yname)

    # Build formula and fit lm (for stargazer structure)
    rhs <- paste(reg_i$rhs, collapse = " + ")
    form <- as.formula(paste(dep, "~", rhs))
    lm_i <- lm(form, data = data)

    # Patch coefficients with lavaan estimates
    beta <- reg_i$est; names(beta) <- reg_i$rhs
    beta0 <- int_i$est; names(beta0) <- "(Intercept)"
    lm_i$coefficients["(Intercept)"] <- beta0
    lm_i$coefficients[names(beta)] <- beta

    # Build SE vector
    se_vec <- rep(NA_real_, length(coef(lm_i)))
    names(se_vec) <- names(coef(lm_i))
    se_vec["(Intercept)"] <- int_i$se
    se_beta <- reg_i$se; names(se_beta) <- reg_i$rhs
    se_vec[names(se_beta)] <- se_beta

    results[[i]] <- list(lm = lm_i, se = se_vec)
  }

  results
}

# Build inputs for all specifications
res_main <- build_stargazer_inputs(
  fit_main, sur_main,
  eq_names = c("y_s", "y_a", "y_j"),
  dep_vars = c("y_s", "y_a", "y_j")
)

res_rob1 <- build_stargazer_inputs(
  fit_rob1, sur_obs,
  eq_names = c("y_s", "y_a", "y_j"),
  dep_vars = c("y_s", "y_a", "y_j")
)

res_rob2 <- build_stargazer_inputs(
  fit_rob2, sur_2sp,
  eq_names = c("y_s", "y_a"),
  dep_vars = c("y_s", "y_a")
)

res_rob3 <- build_stargazer_inputs(
  fit_rob3, sur_main,
  eq_names = c("y_s", "y_a", "y_j"),
  dep_vars = c("y_s", "y_a", "y_j")
)

```



```{r model_selection_table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}

cat("\\begin{table}[!htbp]\n")
cat("\\centering\n")
cat("\\caption{Model selection for SUR biomass dynamics}\n")
cat("\\label{tab:model_selection}\n")
cat("\\begin{tabular}{lcrrr}\n")
cat("\\toprule\n")
cat("Specification & K/eq & AIC & BIC & $\\Delta$AIC \\\\\n")
cat("\\midrule\n")

for (i in seq_len(nrow(model_sel))) {
  spec <- gsub("²", "$^2$", model_sel$Specification[i])
  cat(sprintf("%s & %d & %.1f & %.1f & %.1f \\\\\n",
      spec,
      model_sel$K_per_eq[i],
      model_sel$AIC[i],
      model_sel$BIC[i],
      model_sel$delta_AIC[i]))
}

cat("\\bottomrule\n")
cat("\\end{tabular}\n")
cat("\\end{table}\n")
```



```{r SUR_results, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}

stargazer(
  res_main[[1]]$lm, res_main[[2]]$lm, res_main[[3]]$lm,
  type = "latex",
  header = FALSE,
  float = TRUE,
  title = paste0("SUR estimates of biomass dynamics (main specification, N = ", nrow(sur_main), ")"),
  label = "tab:SUR_results",
  column.labels = c("Sardine", "Anchoveta", "Jack mackerel"),
  column.separate = c(1, 1, 1),
  dep.var.labels.include = FALSE,
  covariate.labels = c(
    "Own biomass ($x_{it}$)",
    "Own biomass$^2$ ($x_{it}^2$)",
    "SST",
    "SST$^2$",
    "Chlorophyll-a",
    "Constant"
  ),
  se = list(res_main[[1]]$se, res_main[[2]]$se, res_main[[3]]$se),
  digits = 3,
  single.row = TRUE,
  omit.stat = c("f"),
  no.space = TRUE,
  font.size = "footnotesize",
  notes = c(
    "$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01.",
    "Robust (Huber--White) SEs via MLR. Biomass in $10^5$ tons.",
    "Environmental variables centered at sample means.",
    "Jack mackerel uses Model B interpolation (see Section 3.1)."
  ),
  notes.append = FALSE,
  notes.align = "l"
)
```



```{r SUR_rob_panel_A, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}

stargazer(
  res_rob1[[1]]$lm, res_rob1[[2]]$lm, res_rob1[[3]]$lm,
  type = "latex",
  header = FALSE,
  float = TRUE,
  title = paste0("SUR robustness: observed jurel CS only (N = ", nrow(sur_obs), ")"),
  label = "tab:SUR_rob_obs",
  column.labels = c("Sardine", "Anchoveta", "Jack mackerel"),
  column.separate = c(1, 1, 1),
  dep.var.labels.include = FALSE,
  se = list(res_rob1[[1]]$se, res_rob1[[2]]$se, res_rob1[[3]]$se),
  digits = 3,
  single.row = TRUE,
  omit.stat = c("f"),
  no.space = TRUE,
  font.size = "footnotesize",
  notes = c("$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01. Robust SEs.",
            "Only years with observed (non-interpolated) jurel CS biomass.",
            "SST$^2$ dropped for parsimony given reduced sample."),
  notes.append = FALSE,
  notes.align = "l"
)
```
```{r SUR_rob_panel_B, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}

stargazer(
  res_rob2[[1]]$lm, res_rob2[[2]]$lm,
  type = "latex",
  header = FALSE,
  float = TRUE,
  title = paste0("SUR robustness: two-species, jurel as exogenous (N = ", nrow(sur_2sp), ")"),
  label = "tab:SUR_rob_2sp",
  column.labels = c("Sardine", "Anchoveta"),
  column.separate = c(1, 1),
  dep.var.labels.include = FALSE,
  se = list(res_rob2[[1]]$se, res_rob2[[2]]$se),
  digits = 3,
  single.row = TRUE,
  omit.stat = c("f"),
  no.space = TRUE,
  font.size = "footnotesize",
  notes = c("$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01. Robust SEs.",
            "Two-species SUR; jack mackerel biomass enters as exogenous covariate."),
  notes.append = FALSE,
  notes.align = "l"
)
```
```{r SUR_rob_interactions, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}

stargazer(
  res_rob3[[1]]$lm, res_rob3[[2]]$lm, res_rob3[[3]]$lm,
  type = "latex",
  header = FALSE,
  float = TRUE,
  title = paste0("SUR robustness: with cross-species interactions (N = ", nrow(sur_main), ")"),
  label = "tab:SUR_interactions",
  column.labels = c("Sardine", "Anchoveta", "Jack mackerel"),
  column.separate = c(1, 1, 1),
  dep.var.labels.include = FALSE,
  se = list(res_rob3[[1]]$se, res_rob3[[2]]$se, res_rob3[[3]]$se),
  digits = 3,
  single.row = TRUE,
  omit.stat = c("f"),
  no.space = TRUE,
  font.size = "footnotesize",
  notes = c("$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01. Robust SEs.",
            "Main specification augmented with pairwise biomass interactions.",
            "No interaction term is individually or jointly significant."),
  notes.append = FALSE,
  notes.align = "l"
)
```


## Trip level cost function

One undergrad students working on this module for his thesis...
Results by July 2026

\<\< NO RESULTS YET \>\>

## Total Annual trips


```{r poisson_dt, message=FALSE, warning=FALSE, include=FALSE}

library(dplyr)
library(fixest)
library(lubridate)


# (A) Define aquí tus especies SPF
spf_codes <- c(26, 33, 114)  

# Dfinir viajes
log_spf <- logbooks %>%
  arrange(COD_BARCO, FECHA_HORA_ZARPE, FECHA_LANCE) %>%
  group_by(COD_BARCO) %>%
  mutate(
    new_trip = case_when(
      NUMERO_LANCE_EX == 1 ~ 1L,
      is.na(lag(FECHA_HORA_ZARPE)) ~ 1L,
      FECHA_HORA_ZARPE != lag(FECHA_HORA_ZARPE) ~ 1L,
      TRUE ~ 0L
    ),
    trip_seq = cumsum(new_trip)
  ) %>%
  ungroup() %>%
  mutate(
    trip_id = paste(COD_BARCO, trip_seq, sep = "_")
  )


# Viajes por barco–año
trips_vy <- log_spf %>%
  distinct(COD_BARCO, year, trip_id) %>%
  count(COD_BARCO, year, name = "T_vy")


Mode <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA)
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

log_spf <- log_spf %>%
  mutate(
    TIPO_EMB_clean = if_else(
      TIPO_EMB %in% c("1", "2", "3"),
      NA_character_,
      TIPO_EMB
    )
  )


# Calculate harvest
harvest_vys <- log_spf %>%
  filter(COD_ESPECIE %in% spf_codes) %>%
  group_by(TIPO_FLOTA, COD_ESPECIE, COD_BARCO, year) %>%
  summarise(
    H_vys = sum(CAPTURA_RETENIDA, na.rm = TRUE)/1000,
    .groups = "drop"
  )

harvest_vy_wide <- harvest_vys %>%
  select(COD_BARCO, year, COD_ESPECIE, H_vys) %>%
  pivot_wider(
    names_from  = COD_ESPECIE,
    values_from = H_vys,
    names_prefix = "H_",
    values_fill = 0
  )

# shares_vsf <- log_spf %>%
#   filter(COD_ESPECIE %in% spf_codes) %>%
#   group_by(TIPO_FLOTA, COD_ESPECIE, COD_BARCO) %>%
#   summarise(
#     H_vsf = sum(CAPTURA_RETENIDA, na.rm = TRUE),
#     .groups = "drop"
#   ) %>%
#   group_by(TIPO_FLOTA, COD_ESPECIE) %>%
#   mutate(
#     share_vsf = H_vsf / sum(H_vsf)
#   ) %>%
#   ungroup()

# Get vessel characteristics
vessel_chars <- log_spf %>%
  group_by(COD_BARCO) %>%
  summarise(
    TIPO_FLOTA = Mode(TIPO_FLOTA),
    TIPO_EMB   = Mode(TIPO_EMB_clean),
    CAPACIDAD_BODEGA = median(CAPACIDAD_BODEGA, na.rm = TRUE),
    .groups = "drop"
  )

vessel_chars <- vessel_chars %>%
  mutate(
    TIPO_EMB = replace_na(TIPO_EMB, "UNK")
  )

poisson_df <- trips_vy %>%
  left_join(vessel_chars, by = "COD_BARCO") %>%
  left_join(harvest_vy_wide, by = c("COD_BARCO", "year"))

```

Because industrial and artisanal vessels operate under distinct technological and regulatory environments, we estimate separate effort models for each fleet segment. This approach allows the responsiveness of fishing trips to prices, quotas, and environmental conditions to differ across fleets, while remaining consistent with the structure of the bioeconomic simulation.


```{r est_poisson, message=FALSE, warning=FALSE, include=FALSE}
library(MASS)

pois_ind <- glm.nb(
  T_vy ~ CAPACIDAD_BODEGA + H_26 + H_33 + H_114 + TIPO_EMB,
  data = poisson_df %>% filter(TIPO_FLOTA == "IND")
)

pois_art <- glm.nb(
  T_vy ~ CAPACIDAD_BODEGA + H_26 + H_33 + H_114 + TIPO_EMB,
  data = poisson_df %>% filter(TIPO_FLOTA == "ART")
)

```

```{r tabla_poisson, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(stargazer)

cat("\\begin{table}[!htbp]\n")
cat("\\centering\n")
cat("\\caption{Annual Fishing Trips (Negative Binomial Models)}\n")
cat("\\label{tab:poisson_results}\n")

stargazer(
  pois_ind, pois_art,
  type = "latex",
  header = FALSE,
  float = FALSE,              # prevents stargazer from creating a duplicate table env
  title = "",
  label = "",
  column.labels = c("Industrial fleet", "Artisanal fleet"),
  column.separate = c(1, 1),
  dep.var.labels = "Number of trips per vessel-year",
  dep.var.caption = "",
  dep.var.labels.include = FALSE,
  digits = 3,
  single.row = TRUE,
  omit.stat = c("f"),
  no.space = TRUE,
  font.size = "footnotesize",
  notes = c("Negative binomial models estimated separately by fleet segment.", "Allocated harvest is computed from historical vessel shares and aggregate TAC."),
  notes.append = FALSE
)

cat("\\end{table}\n")


```


\<\< NO RESULTS YET \>\>

## Inverse demand model

\<\< NO RESULTS YET \>\>

Two undergrad students working on this module for their thesis...
Results by July 2026

## Numerical optimization

```{r simulations, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Ejemplo de estructura de datos simulados
sim_grid <- data.frame(
  vessel_id  = ...,    # id barco
  gear       = ...,    # arte
  year       = ...,    # año
  H_total    = ...,    # harvest anual óptimo del modelo dinámico (ton)
  p_diesel   = ...,    # precios insumos
  w_crew     = ...,
  # Z: características fijas de la embarcación
  length     = ...,
  hp         = ...,
  tonnage    = ...
)

#---------------------------------------------#
# 1) Función: Poisson + costo viaje -> costo anual
#---------------------------------------------#

library(dplyr)

compute_annual_cost <- function(sim_grid,
                                pois_model,
                                cost_model,
                                harvest_var   = "H_total",
                                trips_var     = "trips_hat",
                                htrip_var     = "h_trip",
                                cost_trip_var = "cost_trip_hat",
                                cost_ann_var  = "cost_annual_hat") {
  
  df <- sim_grid
  
  # 1. Predicción de viajes anuales con el Poisson
  #    Nota: para Poisson, E[T] = exp(η), donde η = Xβ
  eta_trips <- predict(pois_model, newdata = df, type = "link")
  df[[trips_var]] <- exp(eta_trips)     # T_hat
  
  # 2. Harvest por viaje (H_total / T_hat)
  #    Evita divisiones por cero poniendo un mínimo a T_hat
  df[[trips_var]][df[[trips_var]] < 1e-6] <- 1e-6
  
  df[[htrip_var]] <- df[[harvest_var]] / df[[trips_var]]
  
  # 3. Preparar 'newdata' para el modelo de costo por viaje
  #    Asumimos que el cost_model usa log(harvest_trip) como regressor.
  #    Creamos una columna 'harvest_trip' temporal para el predict().
  
  df_cost <- df
  
  df_cost$harvest_trip <- df_cost[[htrip_var]]
  
  # 4. Predecir costo por viaje
  #    Si tu modelo está en log(cost_trip), hay que exponentiar:
  cost_link <- predict(cost_model, newdata = df_cost, type = "response")
  # OJO: 
  # - Si el modelo es log(cost_trip) ~ ..., entonces type="response" ya devuelve exp(η).
  # - Si el modelo es cost_trip ~ ..., usa type="response" directamente.
  
  df[[cost_trip_var]] <- cost_link
  
  # 5. Costo anual = T_hat * costo por viaje
  df[[cost_ann_var]] <- df[[trips_var]] * df[[cost_trip_var]]
  
  return(df)
}


```

```{r modelo npv, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

#-------------------------------
# PARÁMETROS BÁSICOS
#-------------------------------

S <- 3          # número de especies (ej: sardina, anchoveta, jurel)
T_hor <- 20     # horizonte en años
rho <- 0.05     # tasa de descuento

species_names <- c("sardine", "anchovy", "jack")

# B0: biomasa inicial por especie (vector de longitud S)
B0 <- c(
  sardine = 500,   # reemplazar por dato real
  anchovy = 400,
  jack    = 600
)

# Aquí podrías tener un data.frame con clima por año (si quieres ya integrarlo)
# climate_df: year x variables SST, CHL, etc.
# climate_df[year, ] se usa dentro de biomass_dynamics()


#-------------------------------
# 1) DINÁMICA DE BIOMASA
#-------------------------------
# Esta función DEBE usar tus coeficientes SUR estimados.
# Aquí dejo una forma “genérica” para que la reemplaces.

biomass_dynamics <- function(B_t, H_t, climate_row, sur_params) {
  # B_t: vector biomasa en t (longitud S)
  # H_t: vector harvest en t (longitud S)
  # climate_row: fila con SST_t, CHL_t, etc.
  # sur_params: lista con coeficientes de tu SUR
  
  # EJEMPLO MUY SIMPLE (placeholder: REEMPLAZAR CON TU SUR)
  # B_{t+1} = B_t + r_s * B_t * (1 - B_t/K_s) - H_t + efecto_clima

  r  <- sur_params$r          # vector r_s
  K  <- sur_params$K          # vector K_s
  sst <- climate_row$sst
  chl <- climate_row$chl

  climate_effect <- sur_params$gamma_sst * sst + sur_params$gamma_chl * chl  # vector

  growth <- r * B_t * (1 - B_t / K) + climate_effect
  B_next <- pmax(B_t + growth - H_t, 1e-6)  # evitar biomasa negativa

  return(B_next)
}


#-------------------------------
# 2) DEMANDA INVERSA
#-------------------------------
# Debes enchufar aquí tu modelo estimado (anual o agregado).
# Recibe harvest total por especie en el año y devuelve precios.

inverse_demand <- function(H_t, demand_params) {
  # H_t: vector harvest anual (S)
  # demand_params: lista con coeficientes de tu modelo de demanda inversa
  
  # EJEMPLO: p_s = a_s + b_s * H_s + c_s * H_otros  (placeholder)
  a <- demand_params$a   # vector a_s
  b <- demand_params$b   # matriz (SxS) o algo similar
  
  p_t <- a + as.numeric(b %*% H_t)
  p_t <- pmax(p_t, 1e-4) # evitar precios <= 0
  
  return(p_t)
}


#-------------------------------
# 3) COSTO ANUAL C(H)
#-------------------------------
# Aquí usas el bloque del PASO 1:
# harvest anual óptimo por barco/flota -> Poisson -> costo anual total.
# Para el NPV, lo que importa es el costo agregado por especie o flota.

annual_cost <- function(H_t, year, scenario, cost_env) {
  # H_t: vector harvest anual (por especie o por flota, según tu diseño)
  # year: año t
  # scenario: etiqueta de clima si la usas
  # cost_env: lista con:
  #   - modelos poisson
  #   - modelos costo por viaje
  #   - shares por barco
  #   - precios de insumos, etc.
  
  # 1. Construir un data.frame "sim_grid" con:
  #    vessel_id, gear, year, H_total (por barco), precios insumos, Z...
  sim_grid <- build_sim_grid_for_year(
    H_t      = H_t,
    year     = year,
    scenario = scenario,
    cost_env = cost_env
  )
  
  # 2. Aplicar la función del PASO 1:
  sim_grid_out <- compute_annual_cost(
    sim_grid   = sim_grid,
    pois_model = cost_env$pois_model,
    cost_model = cost_env$cost_model
  )
  
  # 3. Sumar costos de todos los barcos/flotas:
  C_total <- sum(sim_grid_out$cost_annual_hat)
  
  return(C_total)
}


#-------------------------------
# 4) FUNCIÓN OBJETIVO NPV
#-------------------------------
# Los controles son las fracciones de explotación q_{s,t} = H_{s,t} / B_{s,t}
# OJO: usamos PARÁMETRO q_vec = vector (S*(T_hor)) con q_{s,t}.

npv_objective <- function(q_vec,
                          B0,
                          sur_params,
                          demand_params,
                          cost_env,
                          climate_df,
                          scenario = "base",
                          rho = 0.05) {
  
  # Reorganizamos q_vec (vector largo) en matriz S x T
  q_mat <- matrix(q_vec, nrow = S, ncol = T_hor,
                  dimnames = list(species_names, paste0("year", 1:T_hor)))
  
  # Inicialización
  B_t <- B0
  npv  <- 0
  
  for (t in 1:T_hor) {
    # 1. Harvest anual: H_t = q_{t} * B_t
    q_t <- q_mat[, t]
    q_t <- pmin(pmax(q_t, 0), 0.75)  # restringir 0 <= q <= 0.75 (ejemplo)
    
    H_t <- q_t * B_t
    
    # 2. Precios vía demanda inversa
    p_t <- inverse_demand(H_t, demand_params)
    
    # 3. Costo anual total (usando módulo de viaje)
    C_t <- annual_cost(H_t, year = t, scenario = scenario, cost_env = cost_env)
    
    # 4. Ingresos anuales
    revenue_t <- sum(p_t * H_t)
    
    # 5. Beneficio anual
    profit_t <- revenue_t - C_t
    
    # 6. Descuento
    npv <- npv + profit_t / ((1 + rho)^(t - 1))
    
    # 7. Actualizar biomasa usando SUR + clima
    climate_row <- climate_df[climate_df$year == t & climate_df$scenario == scenario, ]
    B_t <- biomass_dynamics(B_t, H_t, climate_row, sur_params)
  }
  
  # Optim usamos minimización, así que devolvemos -NPV
  return(-npv)
}


#-------------------------------
# 5) LLAMAR A OPTIM
#-------------------------------

# Valores iniciales para q (por ejemplo, 0.1 para todas las especies/años)
q0 <- rep(0.1, S * T_hor)

# Cotas para q: [0, 0.75]
lower_q <- rep(0,   S * T_hor)
upper_q <- rep(0.75, S * T_hor)

# Llamada a optim (L-BFGS-B permite bounds)
opt_res <- optim(
  par     = q0,
  fn      = npv_objective,
  method  = "L-BFGS-B",
  lower   = lower_q,
  upper   = upper_q,
  B0      = B0,
  sur_params    = sur_params,
  demand_params = demand_params,
  cost_env      = cost_env,
  climate_df    = climate_df,
  scenario      = "GFDL",      # o "base", "IPSL", lo que definas
  rho           = rho
)

opt_res$value        # -NPV óptimo
q_opt_vec <- opt_res$par
q_opt_mat <- matrix(q_opt_vec, nrow = S, ncol = T_hor,
                    dimnames = list(species_names, paste0("year", 1:T_hor)))


```

## Projections



```{r proyecciones, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

# climate_df
#   year scenario  sst   chl
# 1 2000  "hist"   ...   ...
# 2 2001  "hist"   ...   ...
# ...
# N 2025  "GFDL"   ...   ...
# N+1 2026 "GFDL"  ...   ...
# ...
# M 2050  "GFDL"   ...   ...

climate_df <- climate_df |>
  group_by(scenario) |>
  mutate(
    sst_anom = (sst - mean(sst[year %in% 1995:2020 & scenario == "hist"])) / sd(sst[year %in% 1995:2020 & scenario == "hist"]),
    chl_anom = (chl - mean(chl[year %in% 1995:2020 & scenario == "hist"])) / sd(chl[year %in% 1995:2020 & scenario == "hist"])
  )


npv_objective <- function(q_vec,
                          B0,
                          sur_params,
                          demand_params,
                          cost_env,
                          climate_df,
                          scenario = "GFDL",
                          rho = 0.05) {
  
  q_mat <- matrix(q_vec, nrow = S, ncol = T_hor,
                  dimnames = list(species_names, paste0("year", 1:T_hor)))
  
  B_t <- B0
  npv <- 0
  
  # Suponemos que t = 1 corresponde a un año base, por ejemplo 2025
  start_year <- 2025
  
  for (t in 1:T_hor) {
    
    year_t <- start_year + (t - 1)
    
    # 1. Harvest anual
    q_t <- q_mat[, t]
    q_t <- pmin(pmax(q_t, 0), 0.75)
    H_t <- q_t * B_t
    
    # 2. Precios vía demanda inversa
    p_t <- inverse_demand(H_t, demand_params)
    
    # 3. Costo anual (usa Poisson + costo por viaje por debajo)
    C_t <- annual_cost(H_t, year = year_t, scenario = scenario, cost_env = cost_env)
    
    revenue_t <- sum(p_t * H_t)
    profit_t  <- revenue_t - C_t
    
    npv <- npv + profit_t / ((1 + rho)^(t - 1))
    
    # 4. Obtener las variables ambientales de ese año y escenario
    climate_row <- climate_df[climate_df$year == year_t & 
                              climate_df$scenario == scenario, ]
    
    # 5. Actualizar biomasa usando SUR + clima proyectado
    B_t <- biomass_dynamics(B_t, H_t, climate_row, sur_params)
  }
  
  return(-npv)
}



biomass_dynamics <- function(B_t, H_t, climate_row, sur_params) {
  # Ejemplo:
  sst_anom <- climate_row$sst_anom
  chl_anom <- climate_row$chl_anom
  
  # g_s(B_t, SST, CHL) estimado con SUR
  # Aquí enchufas tus coeficientes reales:
  growth <- with(sur_params, {
    r * B_t * (1 - B_t / K) +
      beta_sst * sst_anom +
      beta_chl * chl_anom +
      beta_inter * (sst_anom * chl_anom) # solo si lo tienes
  })
  
  B_next <- pmax(B_t + growth - H_t, 1e-6)
  return(B_next)
}


opt_GFDL <- optim(..., scenario = "GFDL", ...)
opt_IPSL <- optim(..., scenario = "IPSL", ...)
opt_HADL <- optim(..., scenario = "HADL", ...)


```


\<\< NO RESULTS YET \>\>

Once I have results for the four models, then I can run optimizations

# Discussion

-   From @sumaila2011: "A shift in species’ geographic range will thus
    affect the distribution and composition of fisheries resources. This
    may affect fishing operations, the allocation of catch shares and
    the effectiveness of fisheries management measures"
-   For the ITQ fisheries (i.e., industrial jack mackerel?:
    -   The theoretical findings on multispecies harvest patterns in
        @Birkenbach2020-nh give rise to nuanced hypotheses about how
        behavior and outcomes will change after the adoption of catch
        shares. For example, a secure property right to catch fish at
        any time in the fishing season allows firms to spread the catch
        of stocks with high prices and downward-sloping demand over a
        longer fishing season. This minimizes market gluts that steer
        product toward lower priced frozen markets [@homans2005] and can
        result in higher prices for those species. Fishermen might also
        shift their efforts toward lower-priced species with cheaper
        quota or toward non-catch-share fisheries, intensifying the race
        to fish for those species during portions of the season
        [@asche2007; @cunningham2016spillovers].
        -   However, we do not include other species than jack mackerel,
            sardine and anchovy that might be caught by thise fleet.
            This would require too expand the model by **N** species,
            which would increase dimensionality of the model.
        -   We need permits to check if actually this happen (... still
            problem with Open-Access fisheries that do not require
            permits)
-   Although the present model captures the historical relationships
    between biomass, temperature, and productivity, future strengthening
    of coastal winds and warming of the Chilean coast (Fuenzalida et al.
    2007; Garreaud et al. 2009) could shift these dynamics, altering
    species overlap and fishery composition.
-   About using CHL for future predictions.
    -   From @sumaila2011: "However, our understanding of the magnitude
        and direction of climate change effects on primary productivity
        is uncertain. For example, projections of primary production
        based on an empirical relationship estimate increases of
        0.7–8.1% by 2050 relative to 2000 (ref. 39), whereas outputs
        from four Earth-system models suggest a possible decrease of
        2–20% by 2100 relative to pre-industrial conditions12. The
        opposite trends and the large regional differences of the
        estimates highlight the uncertainty in the projection of
        large-scale changes in primary productivity".
-   About prices [@sumaila2011]:
    -   "Everything being equal, when climate change reduces fish
        supply, fish price should increase, which could be large enough
        to balance out the loss in gross revenues due to reduced
        catches. However, consumers may seek substitutes as prices
        increase, thereby dampening the demand for fish and reducing the
        potential for price increases."
    -   "Ex-vessel revenues can be affected by climate change through
        effects first on the quantity, quality and distribution of
        catches, and then, on ex-vessel prices of fish."
-   Adaptation [@sumaila2011]:
    -   "When climate change affects the composition and productivity of
        exploited species in a region, some fishers can adapt by
        switching target species or gear type81 or by moving to
        marginally productive areas. Grafton, R. Q. Adaptation to
        climate change in marine capture fisheries. Mar. Policy 34,
        606–615 (2010)."
        -   "For example, new fisheries have already developed for
            several southern species in the UK (for example, red mullet)
            as these species have started to migrate to the North Sea
            because of an increase in sea temperature. Beare, D. J. et
            al. Long-term increase in prevalence of North Sea fishes
            having southern biogeographic affinities. Mar. Ecol. Prog.
            Ser. 284, 269–278 (2004)."
        -   "The ability of fishers and fishing enterprises to adapt
            depends on a number of factors, including the mobility of
            the fishing fleet. - "At present, the dominant hypothesis is
            that more technologically advanced fleets, usually located
            in rich northern countries, are more likely to be better
            prepared to adapt to climate change by moving to other
            fishing grounds and by shifting gears. MacNeil, M. A. et al.
            Transitional states in marine fisheries: adapting to
            predicted global change. Phil. Trans. R. Soc. B 365,
            3753–3763 (2011)."
-   we did not added production risk.... risk preferences? maybe using
    random coefficients? Discusse!!!

_ Julio Peña: Mejorar transferibilidad multispecie de las cuotas, y entre sectores artesanales e industrial

# Conclusions

NO YET!


## Future research question

-   Several other extensions to the model can be incorporated to be
    improved. For instance, the geographical space where fishermen
    operate is relevant, as depending on the location chosen and when to
    participate, the set of potential choices would vary
    [@Reimer2017-jw]. As I mentioned above, it is possible to extend the
    stock dynamic model by considering different locations. The model
    would also require that the participation decision, which is
    captured by the Poisson model on the annual number of trips, should
    then consider the decision to participate in a determined fishing
    ground, connecting the multispecies model of @Kasperski2015-jm to
    the literature of location choice modeling [e.g., @Dupont1993-jn;
    @Smith2005-us; @Hicks2020-mz].

  -   About the DCM?
    -   Including other economic activity would be relevant, as
        aquaculture is become more relevant in the country, where
        some fishermen have swith to small-scale aquaculture
        practice.
    -   Outside option...

-   Does higher quota allocation of jack mackerel, a predator for
    anchovy and sardine, helps small scale sector actually?
    
    
- From @KASPERSKI201655: "In this paper, optimality refers to maximizing the net present value of the harvests of the combined three fisheries and does not consider other ecological or social objectives which would be important to consider in an ecosystem based fisheries management approach."

# Repository

The source code for this project is available on
[GitHub](https://github.com/fquezadae/Impact-of-Environmental-Variability-on-Harvest)

# References
