---
title: "The Impact of Environmental Variability on Fishers' Harvest Decisions in Chile
  using a Multi-Species Approach"
author: "Felipe J. Quezada-Escalona"
date: "Departamento de Economía" #"Last update: `r Sys.Date()`"
output: 
  xaringan::moon_reader:
    css: ["udeconce.css", "default", "default-fonts"]
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```


```{r refmanager, include=FALSE}
library(RefManageR)
source(here::here("helper.R"))
BibOptions(check.entries = FALSE, 
           bib.style = "authoryear", 
           style = "markdown",
           dashed = TRUE)
bib <- ReadBib("bib_paperpile.bib")
citeopt <- list(max.names=2, longnamesfirst = FALSE)
```

# Introduction

- Marine resource distribution is changing due to climate variability `r Citep(bib, "Poloczanska2013-qq", .opts = citeopt)`.
- Fishers respond to change by:
  - Reallocating effort to other species/areas `r Citep(bib, "Gonzalez-Mon2021-kj", .opts = citeopt)`
  - Maintaining the current strategy
  - Exiting the fishery `r Citep(bib, "Powell2022-wj", .opts = citeopt)`
  
---

# Why a Multi-Species Model?

- Diversification improves income stability and climate resilience `r Cite(bib, "Kasperski2013-jz", "Finkbeiner2015-bs", .opts = citeopt)`
- Responses to availability vary by:
  - Port infrastructure
  - Market access
  - Regulations `r Citep(bib, "Powell2022-wj", .opts = citeopt)`
- Different fishers, different choices `r Citep(bib, "Jardine2020-um", .opts = citeopt)`

---

# Case Study: Chile’s Small Pelagic Fishery (SPF)

- Anchoveta, Jack mackerel, Sardine
- ~94% of national catch `r Citep(bib, "SUBPESCA2020", .opts = citeopt)`
- Climate change will impact:
  - Catch levels
  - Prices
  - Species composition

---

# Research Questions

- How will **fishing decisions, catch levels, and prices** evolve under different climate scenarios?
- How do fishers **substitute between species**?

---

# Hypotheses

- Reduced availability → Switch if expected revenue > expected cost in other fishery
- Otherwise → Decrease effort or exit
- Behavior is **heterogeneous**:
  - Geography
  - Gear type `r Citep(bib, "Reimer2017-jw", .opts = citeopt)`

---

# Methodology Overview

**Based on `r Citet(bib, "Kasperski2015-jm", .opts = citeopt)`:**

1. Estimate stock dynamics
2. Estimate trip-level costs
3. Estimate annual trips
4. Estimate inverse demand
5. Simulate climate change effects on profits/harvest

---

# Data Sources

### Requested (2012–2024)

- Stock abundance
- Annual landings
- Trip-level data
- Ex-vessel prices

### To Be Requested/Retrieved

- Sea surface temperature (SST), Chlorophyll, wind (Mostly satellite data)
- Wave height and speed (Chilean navy?)
- Crew wages (maybe INE?) 
- Fuel prices

---

# Model 1: Stock Dynamics

## Base model

`\[
x_{i,y+1} + h_{iy} = (1 + r_i)x_{iy} + \eta_i x_{iy}^2 + \sum_{j \neq i} \alpha_{ij} x_{iy} x_{jy}
\]`

- Follows logistic + interspecies interaction
- Seemingly unrelated regressions (SUR)

## Model with the environment

`\[
x_{i,y+1} + h_{iy} = (1 + r_i)x_{iy} + \eta_i x_{iy}^2 + \sum_{j \neq i} \alpha_{ij} x_{iy} x_{jy} + \rho_i Env_{iy} + \varepsilon_{iy}
\]`

- Adds environmental covariates (SST, chlorophyll)

---

# Model 2: Trip-Level Costs

`\[
\log(c_{iyp}) = \beta_0 + \beta_1 x_{iy} + \beta_2 d_{ip} + \beta_3 z_{p} + \varepsilon_{iyp}
\]`

- Dependent variable: log of average cost per trip
- Explanatory variables:
  - $x_{iy}$: stock abundance
  - $d_{ip}$: distance from port $p$ to fishing ground
  - $z_p$: port-level controls (e.g., infrastructure, fuel prices)

---

# Model 3: Annual Participation

`\[
T_{iy} = \gamma_0 + \gamma_1 R_{iy} + \gamma_2 c_{iy} + \gamma_3 \text{Lagged effort}_{iy} + u_{iy}
\]`

- Predicts number of trips \( T_{iy} \) made by fisher \( i \) in year \( y \)
- Key predictors:
  - Expected revenue \( R_{iy} \)
  - Cost per trip \( c_{iy} \)
  - Lagged behavior (habit persistence)

---

# Model 4: Inverse Demand

`\[
\log(p_{iy}) = \delta_0 + \delta_1 \log(q_{iy}) + \delta_2 w_y + \delta_3 \text{Import}_{iy} + \nu_{iy}
\]`

- Price as a function of quantity landed:
  - \(q_{iy}\): landings
  - \(w_y\): income/proxy demand shifter
  - \(\text{Import}_{iy}\): substitution with imports

---

# Integration and Simulation

- Use models to simulate:
  - Climate-driven stock projections
  - Trip-level and annual harvest adjustments
  - Market responses and price shifts
- Evaluate fisher income, species substitution, and supply chain outcomes


---

# Expected Results

- Climate affects:
  - **Stock** dynamics
  - **Fishing costs**
  - **Trip frequency**
- Catch composition shifts with climate
- Localized market effects

---

# Relevance

- Understand fishers’ adaptive capacity
- Inform climate-resilient fisheries policies in Chile
- Contribute to the sparse local multi-species economic modeling literature [^1]

[^1]: See @PeñaElNiño for ENSO effects in Jack Mackerel fishery using discrete choice.

---

# Thank you!

Felipe J. Quezada-Escalona  
Department of Economics
University of Concepción


**Questions?**

---

# References

```{r, results='asis', echo=FALSE, warning=FALSE}
print_bib_rmd(bib, start = 1, stop = 6)
```

---

# References

```{r, results='asis', echo=FALSE, warning=FALSE}
print_bib_rmd(bib, start = 7, stop = 10)
```