# Reduced-form stress tests and prior elicitation {#appendix-stress}

This appendix describes how we set the priors on the climate shifters
$(\rho_s^{SST}, \rho_s^{CHL})$ that enter the state-space model of the
Stock dynamics subsection of the main text, together with the
identifiability diagnostic that motivates the use of Bayesian methods
over direct maximum likelihood.

## Deterministic hindcast and cross-variant fit {#sec:appendix-stress-spec}

For each stock $s$ we estimate a deterministic Schaefer hindcast over
$2000$--$2024$, $\hat B_{s,t+1} = \hat B_{s,t} + r_{s,t}\hat B_{s,t}(1 - \hat B_{s,t}/K_s) - C_{s,t}$
with $r_{s,t} = r_s^{0} \exp(\rho_s^{SST}(SST_{t-1}-\overline{SST}) + \rho_s^{CHL}(\log CHL_{t-1}-\overline{\log CHL}))$,
stock-dynamics parameters $(r_s^{0}, K_s, B_{0,s})$ held fixed at the
IFOP/SPRFMO single-species prior means reported in Table \ref{tab:stress-structural-priors},
and shifters estimated by bounded least squares with each shifter
constrained to $[-3, 3]$,^[The bound is a biologically extreme upper
limit: $|\rho|=3$ implies that a one-unit anomaly changes productivity
by roughly $95\%$.] minimising median absolute percent error (MAPE).
The structural priors draw on the FishLife meta-analysis for
intrinsic growth rates [@Thorson2020-fishlife], the IFOP V--X
stock-assessment informes for anchoveta and sardine carrying
capacity [@IFOP2024-anchoveta-vx; @IFOP2022-sardina-vx], and the
range-wide SPRFMO assessment for jack mackerel [@SPRFMO2025-sc13].

\begin{table}[H]
\centering\small
\renewcommand{\arraystretch}{1.0}
\caption{Single-species priors on structural stock-dynamics parameters used by the deterministic hindcast and as priors in the state-space likelihood. Priors are log-normal on $(r_s^{0}, K_s, B_{0,s})$; the entry $\log\mathcal{N}(\mu,\sigma)$ denotes $\log\theta \sim \mathcal{N}(\log\mu,\sigma)$.\label{tab:stress-structural-priors}}
\begin{threeparttable}
\begin{tabular}{lccc}
\toprule
Stock & $r_s^{0}$ (yr$^{-1}$) & $K_s$ (kt) & $B_{0,s}$ (kt) \\
\midrule
Anchoveta              & $\log\mathcal{N}(0.6,\,0.25)$   & $\log\mathcal{N}(2{,}200,\,0.15)$ & $\log\mathcal{N}(376,\,0.10)$   \\
Sardine         & $\log\mathcal{N}(0.9,\,0.25)$   & $\log\mathcal{N}(3{,}000,\,0.15)$ & $\log\mathcal{N}(496,\,0.10)$   \\
Jack mackerel          & $\log\mathcal{N}(0.35,\,0.25)$  & $\log\mathcal{N}(8{,}000,\,0.25)$ & $\log\mathcal{N}(5{,}889,\,0.15)$ \\
\bottomrule
\end{tabular}
\begin{tablenotes}[para,flushleft]\footnotesize
\item \emph{Notes.} Intrinsic growth-rate priors ($r_s^{0}$) follow Thorson's FishLife meta-analysis. Carrying capacity ($K_s$) for anchoveta and sardine is derived from the unfished spawning biomass $B_{D0}$ reported in the IFOP V--X stock-assessment informes (anchoveta 2024, sardine 2022), scaled to total biomass; for jack mackerel, $K_s$ is re-scaled from the range-wide SPRFMO 2025 assessment. Initial-biomass priors ($B_{0,s}$) are not virgin biomass: they represent the state-space initial condition at $t=2000$ and are centred at the first observed value in the window (total biomass for the coastal stocks, IFOP acoustic survey for jack mackerel). This empirical-Bayes choice is necessary for identification of the initial state under $N=25$ annual observations; see Online Appendix A.2.
\end{tablenotes}
\end{threeparttable}
\end{table} Four nested variants are reported: base ($\rho=0$), $+$SST
only, $+\log$CHL only, and joint $+$SST$+\log$CHL.
\label{eq:stress-law}
\label{eq:stress-shifter}
\label{sec:appendix-stress-results}

```{r stress-mape-table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
mape_raw <- read.csv("data/bio_params/qa/hindcast_sst_comparison.csv",
 stringsAsFactors = FALSE)

stock_label <- c(
 anchoveta_cs = "Anchoveta",
 sardina_comun_cs = "Sardine",
 jurel_cs = "Jack mackerel"
)

stock_order <- c("anchoveta_cs", "sardina_comun_cs", "jurel_cs")

mape_tbl <- mape_raw %>%
 dplyr::arrange(match(stock_id, stock_order)) %>%
 dplyr::transmute(
 Stock = stock_label[stock_id],
 `Base` = sprintf("%.1f", median_err_pct_base),
 `+ SST` = sprintf("%.1f", median_err_pct_restricted),
 `+ log CHL` = sprintf("%.1f", median_err_pct_chl),
 `+ SST and log CHL` = sprintf("%.1f", median_err_pct_both)
 )

kable(mape_tbl, format = "latex", booktabs = TRUE, escape = FALSE,
 align = "lrrrr",
 caption = "Median absolute percent error of the deterministic Schaefer hindcast across shifter variants. \\label{tab:stress-mape}") %>%
 kable_styling(latex_options = c("hold_position"), font_size = 10)
```

No variant brings any of the three stocks below a $20\%$ MAPE
threshold. The two coastal stocks benefit from activating SST
(anchoveta: $95.9\%\to 30.8\%$, sardine: $98.5\%\to 23.1\%$), with
the joint specification delivering further marginal improvement.
Jack mackerel hindcast errors remain in the $44$--$49\%$ range across
all variants, with no shifter distinguishably better than the base.

The point estimates from the joint variant are interpretable for the
coastal stocks. For anchoveta both shifters are negative
($\rho^{SST} \approx -2.3$, $\rho^{CHL} \approx -2.3$); for sardine
$\rho^{SST} \approx -2.0$ is negative but $\rho^{CHL} \approx +2.1$
is positive---the opposite sign on CHL across
the two species reflects underlying ecological differences. For jack mackerel $\rho^{SST} \approx +0.6$
is small and $\rho^{CHL}$ binds at the lower boundary; we read this
as a failure of identification at the Centro-Sur scale rather than as
strong climate forcing.

## Identifiability diagnostic and prior elicitation {#sec:appendix-stress-diag}
\label{sec:appendix-stress-priors}

Two diagnostics motivate the Bayesian approach. First, the MLE's
difficulties are not due to collinearity: SST and CHL are uncorrelated
over the $2000$--$2024$ sample (Pearson $r = 0.030$, $p = 0.888$), so
the difference in hindcast error across variants cannot be blamed on
linear dependence between the shifters. Second, the jack mackerel MLE binds
at the boundary: the deterministic hindcast cannot separate the climate
signal from noise, so $\rho$ is not separately identifiable. Together
these point to a fundamental identification problem under $N=25$
observations, which the Bayesian state-space specification addresses
in a single estimation step.

Table \ref{tab:stress-priors} reports the priors on
$(\rho_s^{SST}, \rho_s^{CHL})$ used in the structural likelihood.
Prior centres for the two coastal stocks are the joint MLEs rounded to
one decimal place; jack mackerel priors are centred at zero so that posterior
identification is driven by the structural likelihood rather than by
the boundary solution. All priors are independent normal with unit
standard deviation. We do not impose hierarchical pooling across
species because $\rho^{CHL}$ has opposite signs for anchoveta and
sardine---pooling would shrink the two estimates toward a common
center and erase this ecological distinction.

This prior-elicitation strategy is a form of empirical Bayes: the
prior centres for anchoveta and sardine come from the same $2000$--$2024$
biomass series that enters the likelihood. Standard Bayesian practice
keeps prior and data independent; we accept this departure because
with $N=25$ annual observations it is necessary for identification.

We adopt this trade-off deliberately for two reasons. First, the prior
$\sigma = 1$ is wide enough that the data move the posterior
substantially when informative: for anchoveta, the posterior of
$\rho^{CHL}$ migrates to $-3.64$, more than one prior standard
deviation away from the prior centre of $-2.3$ (see the
posterior-identification table of the main text for full posterior
summaries). Second,
centring all priors at zero under $N=25$ would leave the joint
posterior of $(r_s^{0}, K_s, B_{0,s}, \rho_s^{SST}, \rho_s^{CHL})$
poorly identified, producing flat posteriors regardless of whether
climate coupling is real or absent. The empirical-Bayes choice
therefore makes the non-identification result for jack mackerel a
substantive finding, not a numerical artefact.

```{r stress-priors-table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
priors_tbl <- data.frame(
 Stock = c("Anchoveta", "Sardine", "Jack mackerel"),
 `rho_SST_mu` = c("-2.3", "-2.0", "0.0"),
 `rho_SST_sd` = c("1.0", "1.0", "1.0"),
 `rho_CHL_mu` = c("-2.3", "+2.1", "0.0"),
 `rho_CHL_sd` = c("1.0", "1.0", "1.0"),
 Source = c("Joint MLE", "Joint MLE", "Boundary solution; vague prior"),
 check.names = FALSE,
 stringsAsFactors = FALSE
)
names(priors_tbl) <- c("Stock",
 "$\\mu_{\\rho^{SST}}$", "$\\sigma_{\\rho^{SST}}$",
 "$\\mu_{\\rho^{CHL}}$", "$\\sigma_{\\rho^{CHL}}$", "Source")
kable(priors_tbl, format = "latex", booktabs = TRUE, escape = FALSE,
 align = "lrrrrl",
 caption = "Independent normal priors on the climate shifters used in the state-space likelihood. \\label{tab:stress-priors}") %>%
 kable_styling(latex_options = c("hold_position"), font_size = 10)
```
