# Predictive diagnostics {#appendix-predictive}

This appendix reports two out-of-sample diagnostics for the three nested
state-space specifications (T4b-ind, T4b-omega, T4b-full). The two
criteria answer different questions and are included precisely because
they do not always agree:

- **PSIS-LOO** measures how well each specification *describes* the observed
  2000--2024 biomass series after penalising for the model's effective
  dimensionality. It is informative for the in-sample identification
  reported in Section \ref{identification}, because a specification that
  misses systematic variation will be penalised.
- **PSIS-LFO** at the one-year horizon measures the true forecasting
  accuracy on withheld future observations. It is informative for whether
  the identified shifters translate into tangible *next-year* predictive
  gains over a climate-free baseline.

Because the paper uses the estimated shifters for long-run comparative
statics under CMIP6 regimes, rather than for near-term forecasting, it is
LOO that carries the primary identification-supporting role; LFO is
reported as an honesty check against the possibility that the shifters
are simply fitting idiosyncratic in-sample noise.

## B.1 PSIS-LOO cross-validation

We compute Pareto-smoothed importance sampling leave-one-out
cross-validation [@Vehtari2017-qe] with moment matching applied to
observations whose Pareto-$\hat{k}$ diagnostic exceeded 0.7. Results
are summarised in Table \ref{tab:loo-appendix}.

```{r loo-appendix-table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
loo_tbl <- read.csv("tables/loo_t4b_compare.csv", stringsAsFactors = FALSE) %>%
  dplyr::transmute(
    Model                                     = model,
    `$\\widehat{\\text{ELPD}}_{\\text{loo}}$` = sprintf("%.2f", elpd_loo),
    `SE`                                      = sprintf("%.2f", se_elpd_loo),
    `$\\Delta\\widehat{\\text{ELPD}}$`        = ifelse(elpd_diff == 0, "--",
                                                      sprintf("%.2f", elpd_diff)),
    `SE$_{\\Delta}$`                          = ifelse(se_diff == 0, "--",
                                                      sprintf("%.2f", se_diff)),
    `$\\hat{p}_{\\text{loo}}$`                = sprintf("%.1f", p_loo)
  )

kable(loo_tbl, format = "latex", booktabs = TRUE, escape = FALSE,
      align = "lrrrrr",
      caption = "PSIS-LOO comparison across state-space specifications. \\label{tab:loo-appendix}") %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10) %>%
  footnote(general = paste(
      "\u0394 ELPD is the pairwise difference relative to T4b-full (the leading model);",
      "the adjacent SE column reports the standard error of that difference.",
      "After moment matching, 65/68 observations satisfy Pareto-k \u2264 0.7 for T4b-full,",
      "64/68 for T4b-ind, and 67/68 for T4b-omega. The remaining high-Pareto-k points are",
      "the two censored jack-mackerel years (2012, 2015), which appear in all three models",
      "and therefore do not affect the ranking."),
      general_title = "Note:", escape = FALSE, threeparttable = TRUE)
```

The full specification attains the highest predictive density by a wide
margin, outperforming T4b-omega by
$\Delta\widehat{\text{ELPD}}_{\text{LOO}} = 18.02$ (SE $5.59$) and the
environment-free T4b-ind by
$\Delta\widehat{\text{ELPD}}_{\text{LOO}} = 14.59$ (SE $4.93$). Both gaps
exceed three standard errors. Notably, T4b-omega (covariance-rich but
climate-free) underperforms T4b-ind: introducing a residual cross-stock
correlation matrix without climate shifters is penalised by LOO. The
interpretation consistent with the identification results of
Section \ref{identification} is that the cross-stock residual correlation
captured by T4b-omega is a statistical footprint of shared climate
forcing rather than evidence of direct biological interaction; once SST
and CHL enter the growth equation explicitly (T4b-full), the marginal
value of $\Omega$ collapses.

A stock-by-stock decomposition (Figure \ref{fig:loo-pareto-k}) confirms
that the LOO gains of T4b-full are concentrated in the two coastal
pelagics: per-stock ELPD improves by $+4.68$ for anchoveta and $+9.12$
for sardina relative to T4b-ind, whereas the three specifications are
statistically indistinguishable for jack mackerel. The effective number
of parameters $\hat{p}_{\text{loo}} \approx 58$--$65$ is large relative
to $N = 68$, as expected for a state-space model with one latent state
per observation-year; this is similar across the three specifications
and therefore does not confound the pairwise comparison, but it does
motivate the additional forecasting test of Section B.2.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.80\textwidth]{../figs/t4b/loo_t4b_pareto_k.png}
\caption{Per-observation Pareto-$\hat{k}$ values after moment matching,
by model and stock. Dashed line: $\hat{k} = 0.5$; solid line: reliability
threshold $\hat{k} = 0.7$. Crosses denote censored jack-mackerel
observations.}
\label{fig:loo-pareto-k}
\end{figure}

## B.2 PSIS-LFO at one-year horizon

The leave-future-out diagnostic refits each specification on data through
year $t_{\text{cut}}$ and evaluates the log predictive density of the
biomass observations at $t_{\text{cut}} + 1$. We use five origin points,
$t_{\text{cut}} \in \{2011, 2014, 2017, 2020, 2022\}$, spanning the
second half of the estimation window. One-step-ahead marginal predictive
densities are obtained in closed form from the posterior draws of each
truncated fit, applying the log-space Schaefer step of the model and the
marginal predictive variance $\sigma_{\text{proc}}^2 + \sigma_{\text{obs}}^2$
stock by stock. For the two censored jack-mackerel targets (2012 and 2015)
the predictive contribution is the corresponding log-CDF tail term.
Table \ref{tab:lfo-appendix} reports the sum of one-step ELPD across the
five origins and three stocks (14 evaluation points per model, of which
two are censored).

```{r lfo-appendix-table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
lfo_tbl <- read.csv("tables/lfo_t4b_compare.csv", stringsAsFactors = FALSE) %>%
  dplyr::transmute(
    Model                                    = model_label,
    `Points`                                 = n_points,
    `Censored`                               = n_censored,
    `$\\widehat{\\text{ELPD}}_{\\text{LFO}}$` = sprintf("%.2f", elpd_lfo_sum),
    `Mean per point`                         = sprintf("%.2f", elpd_lfo_mean),
    `$\\Delta$ vs leader`                    = ifelse(elpd_diff == 0, "--",
                                                     sprintf("%.2f", elpd_diff))
  )

kable(lfo_tbl, format = "latex", booktabs = TRUE, escape = FALSE,
      align = "lcccrr",
      caption = "One-step-ahead LFO comparison across state-space specifications. \\label{tab:lfo-appendix}") %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10) %>%
  footnote(general = paste(
      "Sum of one-step predictive log densities over five origin years",
      "(2011, 2014, 2017, 2020, 2022) and three stocks (14 targets per model; two are censored).",
      "Each refit uses 4 chains \u00d7 1500 warmup + 1500 sampling iterations.",
      "All 15 refits satisfied R-hat \u2264 1.01 with at most one divergent transition",
      "(T4b-omega at the 2011 origin), which does not affect the reported ELPD."),
      general_title = "Note:", escape = FALSE, threeparttable = TRUE)
```

In sharp contrast to the LOO ranking, the three specifications are
statistically indistinguishable in LFO: the spread between the best and
worst model is only $1.36$ ELPD units over 14 target points (mean
$0.10$ per target). Restricting attention to the 12 uncensored targets---
that is, excluding the two jack-mackerel censored years whose predictive
contribution is dominated by a common log-CDF tail term---the differences
shrink further, to less than $0.3$ ELPD per model pair. Figure
\ref{fig:lfo-elpd-path} displays the per-target trajectory of the one-step
ELPD by stock.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.80\textwidth]{../figs/t4b/lfo_t4b_elpd_path.png}
\caption{One-step-ahead predictive log density by target year and stock,
across the three state-space specifications. Dashed line at $\text{ELPD}=0$
marks the threshold above which the model predicts the withheld observation
better than an uninformative uniform density over a one-log-unit interval.}
\label{fig:lfo-elpd-path}
\end{figure}

The LOO-LFO divergence is not a contradiction but a feature of the
research design. As noted in Section \ref{identification}, the historical
SST anomaly has standard deviation
$\hat{\sigma}_{SST} = 0.26\,^{\circ}\text{C}$, small relative to the
process-noise standard deviation
$\hat{\sigma}_{\text{proc}} \in \{0.17, 0.23, 1.22\}$ across the three
stocks. Year-to-year climate innovations therefore contribute little to
a one-year-ahead nowcast, even when the structural coupling between
climate and biological productivity is strong. The same coupling
accumulates over decadal horizons, which is the scale at which the
paper's projections operate and the scale at which the identified
$\rho_i$ are economically consequential. Stated differently, LOO
confirms that T4b-full captures systematic structure that the simpler
specifications miss (because it must explain observed 2000--2024
dynamics), while LFO confirms that the added structure is not an
artefact of overfitting to short-run noise (because it yields no
one-year advantage). The two diagnostics jointly support the structural
interpretation of the shifters adopted in the main text.
