# Reduced-form stress tests and prior elicitation {#appendix-stress}

This appendix documents the protocol used to elicit the prior densities
on the climate shifters $(\rho_i^{SST}, \rho_i^{CHL})$ that enter the
state-space specification of the Stock dynamics subsection of the main text, together
with the identifiability diagnostic that motivates the recourse to a
Bayesian likelihood rather than a direct maximum-likelihood treatment.

## Deterministic hindcast and cross-variant fit {#sec:appendix-stress-spec}

For each stock $i$ we estimate a deterministic Schaefer hindcast over
$2000$--$2024$, $\hat B_{i,t+1} = \hat B_{i,t} + r_{i,t}\hat B_{i,t}(1 - \hat B_{i,t}/K_i) - C_{i,t}$
with $r_{i,t} = r_i^{0} \exp(\rho_i^{SST}(SST_{t-1}-\overline{SST}) + \rho_i^{CHL}(\log CHL_{t-1}-\overline{\log CHL}))$,
biological parameters $(r_i^{0}, K_i)$ held at the centres of the
IFOP/SPRFMO single-species priors, and shifters estimated by bounded
least squares on $[-3,3]^2$ minimising median absolute percent error
(MAPE). Four nested variants are reported: base ($\rho=0$), $+$SST
only, $+\log$CHL only, and joint $+$SST$+\log$CHL.
\label{eq:stress-law}
\label{eq:stress-shifter}
\label{sec:appendix-stress-results}

```{r stress-mape-table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
mape_raw <- read.csv("data/bio_params/qa/hindcast_sst_comparison.csv",
 stringsAsFactors = FALSE)

stock_label <- c(
 anchoveta_cs = "Anchoveta",
 sardina_comun_cs = "Sardina com\\'un",
 jurel_cs = "Jack mackerel"
)

stock_order <- c("anchoveta_cs", "sardina_comun_cs", "jurel_cs")

mape_tbl <- mape_raw %>%
 dplyr::arrange(match(stock_id, stock_order)) %>%
 dplyr::transmute(
 Stock = stock_label[stock_id],
 `Base` = sprintf("%.1f", median_err_pct_base),
 `+ SST` = sprintf("%.1f", median_err_pct_restricted),
 `+ log CHL` = sprintf("%.1f", median_err_pct_chl),
 `+ SST and log CHL` = sprintf("%.1f", median_err_pct_both)
 )

kable(mape_tbl, format = "latex", booktabs = TRUE, escape = FALSE,
 align = "lrrrr",
 caption = "Median absolute percent error of the deterministic Schaefer hindcast across shifter variants. \\label{tab:stress-mape}") %>%
 kable_styling(latex_options = c("hold_position"), font_size = 10)
```

No variant brings any of the three stocks below a $20\%$ MAPE
threshold. The two coastal stocks benefit from activating SST
(anchoveta: $95.9\%\to 30.8\%$, sardine: $98.5\%\to 23.1\%$), with
the joint specification delivering further marginal improvement.
Jack mackerel hindcast errors remain in the $44$--$49\%$ range across
all variants, with no shifter distinguishably better than the base.

The joint-variant point estimates carry an internally coherent
biological reading for the coastal stocks. Anchoveta: both shifters
negative and similar in magnitude
($\rho^{SST} \approx -2.3$, $\rho^{CHL} \approx -2.3$). Sardina
com\'un: $\rho^{SST} \approx -2.0$ but $\rho^{CHL} \approx +2.1$
(opposite sign to anchoveta, consistent with the warm-water mesotrophic
affinity that distinguishes sardine from anchoveta in the Humboldt
Current biological literature). Jack mackerel: $\rho^{SST} \approx +0.6$
small and the $\rho^{CHL}$ estimate binds at the lower boundary of the
admissible box; we read the boundary solution as a failure of point
identification rather than a finding of strong climate forcing.

## Identifiability diagnostic and prior elicitation {#sec:appendix-stress-diag}
\label{sec:appendix-stress-priors}

Two diagnostics motivate the use of weakly informative priors rather
than direct maximum-likelihood estimation. First, the two environmental
forcings are statistically orthogonal over the $2000$--$2024$ sample
(Pearson $r = 0.030$, $p = 0.888$), so the cross-variant gap in
hindcast error cannot be attributed to collinearity in the design
matrix. Second, the jurel MLE binds at the boundary, indicating that
the climate signal is not separately identifiable from process and
observation noise absorbed into the residual under a deterministic
single-species hindcast; the Bayesian state-space specification
addresses both limitations within a single inferential framework.

Table \ref{tab:stress-priors} reports the priors on
$(\rho_i^{SST}, \rho_i^{CHL})$ used in the structural likelihood.
Prior centres for the two coastal stocks are the joint MLEs rounded to
one decimal place; jurel priors are centred at zero so that posterior
identification is driven by the structural likelihood rather than by
the boundary solution. All priors are independent normal with unit
standard deviation; no hierarchical pooling is imposed because
$\rho^{CHL}$ switches sign between anchoveta and sardina com\'un.

```{r stress-priors-table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
priors_tbl <- data.frame(
 Stock = c("Anchoveta", "Sardina com\\'un", "Jack mackerel"),
 `rho_SST_mu` = c("-2.3", "-2.0", "0.0"),
 `rho_SST_sd` = c("1.0", "1.0", "1.0"),
 `rho_CHL_mu` = c("-2.3", "+2.1", "0.0"),
 `rho_CHL_sd` = c("1.0", "1.0", "1.0"),
 Source = c("Joint MLE", "Joint MLE", "Boundary solution; vague prior"),
 check.names = FALSE,
 stringsAsFactors = FALSE
)
names(priors_tbl) <- c("Stock",
 "$\\mu_{\\rho^{SST}}$", "$\\sigma_{\\rho^{SST}}$",
 "$\\mu_{\\rho^{CHL}}$", "$\\sigma_{\\rho^{CHL}}$", "Source")
kable(priors_tbl, format = "latex", booktabs = TRUE, escape = FALSE,
 align = "lrrrrl",
 caption = "Independent normal priors on the climate shifters used in the state-space likelihood. \\label{tab:stress-priors}") %>%
 kable_styling(latex_options = c("hold_position"), font_size = 10)
```
