# Decomposition of trip-level projection uncertainty across the CMIP6 ensemble {#appendix-trips-ensemble}

The fleet-level comparative statics on trips reported in
Section \ref{sec:tripresults} (Table \ref{tab:trip_compstat}) integrate
over three sources of uncertainty: the posterior of the structural
shifters $(\rho^{SST}_s, \rho^{CHL}_s)$ given the historical record,
the heterogeneity of catch composition $(\omega_{v,s})$ and historical
quota allocation $(H^{\text{alloc}}_v)$ across vessels within fleet,
and the spread of climate-forced trajectories across the CMIP6 ensemble.
This appendix decomposes the total variance of the projected percentage
change in fleet-level annual trips, $\Delta T_f / T_f^{(0)}$, into a
within-model component (posterior uncertainty pooled with vessel
heterogeneity within fleet) and a between-model component (CMIP6
ensemble spread), using the law of total variance. It is the
companion to Appendix \ref{appendix-ensemble} and answers a different
diagnostic question: *whether the narrow cross-model interquartile
range observed for fleet-level trip responses in
Table \ref{tab:trip_compstat} reflects climate consensus or a
floor-effect saturation of the underlying biomass collapse.*

## Decomposition

Let $X_{f,m,d,v}$ denote the percentage change in annual trips for
vessel $v$ in fleet $f$ implied by climate model $m$ and posterior draw
$d$ under a fixed scenario--window pair $(SSP, w)$. The total variance
of $X_{f,\cdot,\cdot,\cdot}$ across $(m, d, v)$ admits the decomposition

\begin{equation}
\underbrace{\mathrm{Var}_{(m,d,v)}(X_{f,\cdot,\cdot,\cdot})}_{\text{total}}
\;=\;
\underbrace{\mathbb{E}_m\!\left[\mathrm{Var}_{(d,v)}(X_{f,m,\cdot,\cdot} \mid m)\right]}_{\text{within-model (posterior + vessels)}}
\;+\;
\underbrace{\mathrm{Var}_m\!\left[\mathbb{E}_{(d,v)}(X_{f,m,\cdot,\cdot} \mid m)\right]}_{\text{between-model (climate)}}
\label{eq:appG-decomp}
\end{equation}

The first term is the average variance within a model, taken over
posterior draws and vessels within fleet jointly: it captures
uncertainty about the structural elasticities and heterogeneity in
vessel exposure under a *fixed* climate forcing. The second term is
the variance across CMIP6 models of the within-model means: it captures
uncertainty about the magnitude of the climate forcing itself.

We pool posterior draws and vessel heterogeneity into the within-model
component for two reasons. First, structural parallelism with
Appendix \ref{appendix-ensemble}, which reports an exactly analogous
two-way decomposition for stock-level intrinsic productivity. Second,
vessel heterogeneity in $(\omega_{v,s}, H^{\text{alloc}}_v)$ is a fixed
feature of the historical fleet that does not shrink with longer
historical samples or larger climate ensembles; reporting it as part of
the within-model component is faithful to the reading that a longer
record or a richer ensemble would not, by construction, reduce that
share. A three-way decomposition that separates posterior, vessel,
and climate contributions can be obtained from the same
`factor_trips_dt` object generated by the trip-pipeline script and is
provided in the supplementary code; the qualitative pattern is
unchanged.

We compute Equation \eqref{eq:appG-decomp} by Monte Carlo from the
T4b-full posterior crossed with the six-model CMIP6 ensemble, propagated
through the Schaefer steady-state biomass equation under historical
average fishing pressure and the negative binomial trip equation, for
each of the 830 vessels in the panel.

## Results

```{r appG-load, echo=FALSE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
})

appG_raw <- readr::read_csv(
  "tables/appendix_g_trips_variance_decomposition_raw.csv",
  show_col_types = FALSE)

# Helpers para extraer una celda formatted
appG_pct <- function(fleet, scn, win, col) {
  v <- appG_raw %>%
    dplyr::filter(TIPO_FLOTA == fleet,
                  scenario   == scn,
                  window     == win) %>%
    dplyr::pull({{ col }})
  sprintf("%.0f%%", 100 * v[1])
}
appG_sd <- function(fleet, scn, win, col) {
  v <- appG_raw %>%
    dplyr::filter(TIPO_FLOTA == fleet,
                  scenario   == scn,
                  window     == win) %>%
    dplyr::pull({{ col }})
  sprintf("%.4f", v[1])
}

# Atajos para celdas que la prosa cita
art_245mid_within   <- appG_pct("ART", "ssp245", "mid", pct_within)
art_585end_within   <- appG_pct("ART", "ssp585", "end", pct_within)
ind_585end_within   <- appG_pct("IND", "ssp585", "end", pct_within)
sd_art_585end       <- appG_sd("ART",  "ssp585", "end", sd_total)
sd_ind_585end       <- appG_sd("IND",  "ssp585", "end", sd_total)
```

Table \ref{tab:appG-decomp} reports the decomposition for the four
scenario--window pairs and two fleets.

```{r appG-table, echo=FALSE, results='asis', message=FALSE, warning=FALSE}
appG_tab <- appG_raw %>%
  dplyr::transmute(
    Fleet                     = fleet_label,
    Scenario                  = scenario_label,
    `$n_m$`                   = n_models,
    `Mean $\\%\\Delta$`       = sprintf("$%+0.1f\\%%$", 100 * mean_pct_change),
    `$SD$ total`              = sprintf("%.4f", sd_total),
    `$\\%$ within`            = sprintf("$%.0f\\%%$", 100 * pct_within),
    `$\\%$ between`           = sprintf("$%.0f\\%%$", 100 * pct_between)
  )

knitr::kable(appG_tab,
             format    = "latex",
             escape    = FALSE,
             booktabs  = TRUE,
             align     = "llcrrrr",
             caption   = paste("Variance decomposition of",
                               "$\\Delta T_f / T_f^{(0)}$",
                               "into within-model (posterior pooled",
                               "with vessel heterogeneity) and",
                               "between-model (CMIP6) components.",
                               "$n_m$ is the number of CMIP6 models",
                               "contributing to the cell."),
             label     = "appG-decomp")
```

The decomposition shows that the within-model component dominates the
total variance across all eight scenario--window/fleet cells: it
accounts for `r art_245mid_within` of total variance for the artisanal
fleet under SSP2-4.5 mid-century and saturates at `r art_585end_within`
under SSP5-8.5 end-of-century; for the industrial fleet, the
within-model share rises from `r appG_pct("IND", "ssp245", "mid", pct_within)`
under SSP2-4.5 mid-century to `r ind_585end_within` under SSP5-8.5
end-of-century. The between-model component---the share of total
variance attributable to disagreement across CMIP6 models on the
magnitude of climate forcing---is at most $4\%$ of total variance in
the SSP2-4.5 mid-century cells and is numerically zero, to the nearest
percent, in the remaining six cells. This is *not* a statement that the
CMIP6 ensemble is in tight agreement on the climate signal that drives
fleet-level trips, in the sense that one would attribute to the
sub-ensemble of CMIP6 models with similar equilibrium climate
sensitivity; it is the mechanical signature of floor-effect saturation
of the underlying biomass collapse channel.

The mechanism is straightforward. Once the Schaefer steady-state
biomass factor $f^H_v$ collapses across most posterior draws, the
negative binomial trip equation maps that collapse into
$\exp(\beta_f \cdot H^{\text{alloc}}_v \cdot (f^H_v - 1)) \approx
\exp(-\beta_f \cdot H^{\text{alloc}}_v)$, a quantity determined by the
fleet-specific business mechanics ($\beta_f$ and the historical quota
allocation) rather than by the magnitude of the climate forcing. Once
all CMIP6 models drive the system into this saturation region, the
between-model spread of the within-model mean trip response collapses
to a narrow band---numerically near zero in the present sample---while
the within-model variance retains the dispersion induced by vessel
heterogeneity in $(\omega_{v,s}, H^{\text{alloc}}_v)$ and by the
structural posterior across that subset of draws in which the portfolio
survives. The total $SD$ values are correspondingly small in absolute
magnitude (`r sd_art_585end` for the artisanal fleet and
`r sd_ind_585end` for the industrial fleet under SSP5-8.5
end-of-century), confirming that what little variance remains is
within-model.

The means reported in Table \ref{tab:appG-decomp} are systematically
more negative than the medians reported in Table \ref{tab:trip_compstat}
of the main results section---for example, $-17.5\%$ for the artisanal
fleet under SSP5-8.5 end-of-century in this appendix versus $-10.2\%$ in
Table \ref{tab:trip_compstat}. This is the same left-skew mechanism
documented in Appendix \ref{appendix-ensemble} for the underlying
productivity response: posterior draws and vessels in which the
portfolio collapses pull the median up toward the floor at
$\exp(-\beta_f \cdot H^{\text{alloc}}_v) - 1$, while the mean integrates
over the long left tail of vessel-and-draw configurations in which the
collapse drives $f^H_v$ deep into negative territory and the trip
response down with it. Both summaries are valid; we report cross-model
medians in the main results table because the median is robust to the
extreme tail and is the natural complement to the cross-model
inter-quartile range, and we report cross-model means in this appendix
because the law of total variance in
Equation \eqref{eq:appG-decomp} is constructed on means rather than
medians.

This reading is consistent with the underlying biomass-level
decomposition reported in Appendix \ref{appendix-ensemble}: for
sardine---the stock that drives the artisanal fleet's portfolio
collapse---the SSP5-8.5 end-of-century cell exhibits a total $SD$ of
$0.010$ and an inverted within/between ratio that the appendix
explicitly reads as a floor effect at the lower bound of $-100\%$. The
fleet-level decomposition reported here inherits that floor effect: the
total $SD$ for the artisanal fleet under SSP5-8.5 end-of-century is
`r sd_art_585end` and the corresponding industrial cell is
`r sd_ind_585end`, both of which are small absolute magnitudes and
mechanically consistent with the saturated regime.

The substantive policy implication is that, for fleet-level trip
projections, the marginal information value of expanding the CMIP6
ensemble is small relative to that of tightening the structural
posterior or refining the description of vessel heterogeneity within
fleet. The dominant uncertainty is *not* "which climate model is
right" but "given that all of them imply portfolio collapse for the
two coastal stocks, how does the resulting heterogeneous economic
exposure across vessels translate into trip-level effort." This is the
econometric counterpart of the well-known result in the climate-impacts
literature that ensemble averaging masks first-order distributional
heterogeneity downstream of the climate signal [@Cline2017-dp;
@Free2019; @Kasperski2013-jz; @Oken2021-of].

Figure \ref{fig:appG-decomp} visualises the decomposition.

\begin{figure}[!ht]
\centering
\includegraphics[width=0.85\textwidth]{figs/t4b/appendix_g_trips_variance_decomposition.png}
\caption{Variance decomposition of $\Delta T_f / T_f^{(0)}$ across the
CMIP6 ensemble. Bars stack to one within each scenario--window pair
and split the total variance into within-model (posterior pooled with
vessel heterogeneity, blue) and between-model (CMIP6 spread, orange)
components. The within-model component dominates across all cells,
reflecting the floor-effect saturation of the underlying Schaefer
steady-state biomass collapse rather than tight cross-model agreement
on the climate signal.}
\label{fig:appG-decomp}
\end{figure}
