# Reduced-form stress tests and prior elicitation {#appendix-stress}

This appendix describes how we set the priors on the climate shifters
$(\rho_s^{SST}, \rho_s^{CHL})$ that enter the state-space model of the
Stock dynamics subsection of the main text, together with the
identifiability diagnostic that motivates the use of Bayesian methods
over direct maximum likelihood.

## Deterministic hindcast and cross-variant fit {#sec:appendix-stress-spec}

For each stock $s$ we estimate a deterministic Schaefer hindcast over
$2000$--$2024$, $\hat B_{s,t+1} = \hat B_{s,t} + r_{s,t}\hat B_{s,t}(1 - \hat B_{s,t}/K_s) - C_{s,t}$
with $r_{s,t} = r_s^{0} \exp(\rho_s^{SST}(SST_{t-1}-\overline{SST}) + \rho_s^{CHL}(\log CHL_{t-1}-\overline{\log CHL}))$,
stock-dynamics parameters $(r_s^{0}, K_s, B_{0,s})$ centred at the
IFOP/SPRFMO single-species priors, and shifters estimated by bounded
least squares on $[-3,3]^2$ minimising median absolute percent error
(MAPE). Four nested variants are reported: base ($\rho=0$), $+$SST
only, $+\log$CHL only, and joint $+$SST$+\log$CHL.
\label{eq:stress-law}
\label{eq:stress-shifter}
\label{sec:appendix-stress-results}

```{r stress-mape-table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
mape_raw <- read.csv("data/bio_params/qa/hindcast_sst_comparison.csv",
 stringsAsFactors = FALSE)

stock_label <- c(
 anchoveta_cs = "Anchoveta",
 sardina_comun_cs = "Sardina com\\'un",
 jurel_cs = "Jack mackerel"
)

stock_order <- c("anchoveta_cs", "sardina_comun_cs", "jurel_cs")

mape_tbl <- mape_raw %>%
 dplyr::arrange(match(stock_id, stock_order)) %>%
 dplyr::transmute(
 Stock = stock_label[stock_id],
 `Base` = sprintf("%.1f", median_err_pct_base),
 `+ SST` = sprintf("%.1f", median_err_pct_restricted),
 `+ log CHL` = sprintf("%.1f", median_err_pct_chl),
 `+ SST and log CHL` = sprintf("%.1f", median_err_pct_both)
 )

kable(mape_tbl, format = "latex", booktabs = TRUE, escape = FALSE,
 align = "lrrrr",
 caption = "Median absolute percent error of the deterministic Schaefer hindcast across shifter variants. \\label{tab:stress-mape}") %>%
 kable_styling(latex_options = c("hold_position"), font_size = 10)
```

No variant brings any of the three stocks below a $20\%$ MAPE
threshold. The two coastal stocks benefit from activating SST
(anchoveta: $95.9\%\to 30.8\%$, sardine: $98.5\%\to 23.1\%$), with
the joint specification delivering further marginal improvement.
Jack mackerel hindcast errors remain in the $44$--$49\%$ range across
all variants, with no shifter distinguishably better than the base.

The joint-variant point estimates are internally coherent for the
coastal stocks. Anchoveta: both shifters negative and similar in
magnitude ($\rho^{SST} \approx -2.3$, $\rho^{CHL} \approx -2.3$).
Sardina com\'un: $\rho^{SST} \approx -2.0$ but $\rho^{CHL} \approx +2.1$
(opposite sign for sardine, consistent with the established ecological
differentiation between sardine and anchoveta in the Humboldt system).
Jack mackerel: $\rho^{SST} \approx +0.6$ small and the $\rho^{CHL}$
estimate binds at the lower boundary of the admissible box; we read
the boundary solution as a failure of point identification rather
than a finding of strong climate forcing.

## Identifiability diagnostic and prior elicitation {#sec:appendix-stress-diag}
\label{sec:appendix-stress-priors}

Two diagnostics motivate the use of weakly informative priors rather
than direct maximum-likelihood estimation. First, the two environmental
forcings are statistically orthogonal over the $2000$--$2024$ sample
(Pearson $r = 0.030$, $p = 0.888$), so the cross-variant gap in
hindcast error cannot be attributed to collinearity in the design
matrix. Second, the jurel MLE binds at the boundary: the deterministic
hindcast cannot separate the climate signal from process and observation
noise, so $\rho$ is not separately identifiable. The Bayesian
state-space specification addresses both limitations within a single
estimation step.

Table \ref{tab:stress-priors} reports the priors on
$(\rho_s^{SST}, \rho_s^{CHL})$ used in the structural likelihood.
Prior centres for the two coastal stocks are the joint MLEs rounded to
one decimal place; jurel priors are centred at zero so that posterior
identification is driven by the structural likelihood rather than by
the boundary solution. All priors are independent normal with unit
standard deviation; no hierarchical pooling is imposed because
$\rho^{CHL}$ switches sign between anchoveta and sardine.

This prior-elicitation strategy is a form of empirical Bayes: the
prior centres for anchoveta and sardine are derived from the same
2000--2024 biomass series that enters the state-space likelihood,
which trades pure data-prior independence for tractable identification
under $N=25$ annual observations. We adopt this trade-off deliberately
for two reasons. First, the prior standard deviation $\sigma = 1$ is
large enough that the posterior is free to move substantially: the
posterior of $\rho_s^{CHL}$ for anchoveta migrates to $-3.65$
(prior mean $-2.3$), more than one prior standard deviation from
the centre, illustrating that the data does refine the prior when
informative. Second, the alternative of centring all priors at zero
would, under $N=25$, leave the joint posterior of
$(r_s^{0}, K_s, B_{0,s}, \rho_s^{SST}, \rho_s^{CHL})$ weakly identified
or multimodal, producing posteriors that are uninformative for
projections regardless of whether the underlying coupling is real or
absent. The empirical-Bayes choice therefore makes the non-identification
result for jack mackerel a substantive finding rather than a numerical
artefact of weak posterior geometry.

```{r stress-priors-table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
priors_tbl <- data.frame(
 Stock = c("Anchoveta", "Sardina com\\'un", "Jack mackerel"),
 `rho_SST_mu` = c("-2.3", "-2.0", "0.0"),
 `rho_SST_sd` = c("1.0", "1.0", "1.0"),
 `rho_CHL_mu` = c("-2.3", "+2.1", "0.0"),
 `rho_CHL_sd` = c("1.0", "1.0", "1.0"),
 Source = c("Joint MLE", "Joint MLE", "Boundary solution; vague prior"),
 check.names = FALSE,
 stringsAsFactors = FALSE
)
names(priors_tbl) <- c("Stock",
 "$\\mu_{\\rho^{SST}}$", "$\\sigma_{\\rho^{SST}}$",
 "$\\mu_{\\rho^{CHL}}$", "$\\sigma_{\\rho^{CHL}}$", "Source")
kable(priors_tbl, format = "latex", booktabs = TRUE, escape = FALSE,
 align = "lrrrrl",
 caption = "Independent normal priors on the climate shifters used in the state-space likelihood. \\label{tab:stress-priors}") %>%
 kable_styling(latex_options = c("hold_position"), font_size = 10)
```
