# Reduced-form stress tests and prior elicitation {#appendix-stress}

This appendix formalises the protocol used to elicit the prior densities
on the climate shifters $(\rho_i^{SST}, \rho_i^{CHL})$ that enter the
state-space specification of Section \ref{sec:stock-dynamics}. The
protocol has two purposes. First, it documents an independent,
reduced-form fit of the deterministic Schaefer hindcast augmented with
SST and log-CHL forcings; this fit motivates the choice of prior centres
and standard deviations. Second, it serves as a diagnostic for the
identifiability of $(\rho_i^{SST}, \rho_i^{CHL})$ under a non-Bayesian
point estimator, which we use to justify the recourse to a Bayesian
state-space specification with weakly informative priors rather than to
a maximum-likelihood treatment of the structural model.

## Hindcast specification {#sec:appendix-stress-spec}

For each stock $i \in \{\text{anchoveta}, \text{sardina común}, \text{jack mackerel}\}$ we estimate a deterministic Schaefer hindcast over the 2000--2024 estimation window,

\begin{equation}
\hat B_{i,t+1} \;=\; \hat B_{i,t} \,+\, r_{i,t}\, \hat B_{i,t}\!\left(1 - \frac{\hat B_{i,t}}{K_i}\right) \,-\, C_{i,t},
\qquad \hat B_{i,t_0} = B_{i,t_0}^{\text{obs}},
\label{eq:stress-law}
\end{equation}

with the climate-modulated growth rate

\begin{equation}
r_{i,t} \;=\; r_i^{0} \exp\!\Big( \rho_i^{SST}\, (SST_{t-1} - \overline{SST}) \;+\; \rho_i^{CHL}\, (\log CHL_{t-1} - \overline{\log CHL}) \Big).
\label{eq:stress-shifter}
\end{equation}

The biological parameters $(r_i^{0}, K_i)$ are held at the centres of the
stock-specific priors elicited from the official IFOP and SPRFMO
single-species assessments (anchoveta: $r_i^{0} = 0.6$,
$K_i = 2{,}200$~kt; sardina común: $r_i^{0} = 0.9$,
$K_i = 3{,}000$~kt; jack mackerel: $r_i^{0} = 0.35$,
$K_i = 8{,}000$~kt). Catches $C_{i,t}$ are observed and identical to
those used in the structural likelihood. The shifters
$(\rho_i^{SST}, \rho_i^{CHL})$ are estimated by bounded least squares on
the box $[-3, 3]^2$, minimising the median absolute percent prediction
error,

\begin{equation*}
\text{MAPE}_i(\rho_i^{SST}, \rho_i^{CHL}) \;=\; \operatorname{median}_t \left| \frac{\hat B_{i,t} - B_{i,t}^{\text{obs}}}{B_{i,t}^{\text{obs}}} \right|.
\end{equation*}

We estimate four nested variants for each stock: a base case with
$\rho_i^{SST} = \rho_i^{CHL} = 0$, two univariate cases that activate
SST or log-CHL alone, and a joint case that activates both. The
variants are not nested in an econometric sense; the comparison is
purely descriptive of how much hindcast error each shifter can absorb.
We adopt a $20\%$ MAPE threshold as a conventional benchmark for
acceptable deterministic hindcast performance in the small-pelagic
literature; this threshold plays no role in the structural likelihood
and is used here solely to summarise the cross-variant comparison.

## Cross-variant fit results {#sec:appendix-stress-results}

Table \ref{tab:stress-mape} reports the median absolute percent error
of the deterministic hindcast for each stock and each variant.

```{r stress-mape-table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
mape_raw <- read.csv("data/bio_params/qa/hindcast_sst_comparison.csv",
                     stringsAsFactors = FALSE)

stock_label <- c(
  anchoveta_cs       = "Anchoveta",
  sardina_comun_cs   = "Sardina común",
  jurel_cs           = "Jack mackerel"
)

stock_order <- c("anchoveta_cs", "sardina_comun_cs", "jurel_cs")

mape_tbl <- mape_raw %>%
  dplyr::arrange(match(stock_id, stock_order)) %>%
  dplyr::transmute(
    Stock              = stock_label[stock_id],
    `Base`             = sprintf("%.1f", median_err_pct_base),
    `+ SST`            = sprintf("%.1f", median_err_pct_restricted),
    `+ log CHL`        = sprintf("%.1f", median_err_pct_chl),
    `+ SST and log CHL` = sprintf("%.1f", median_err_pct_both)
  )

kable(mape_tbl, format = "latex", booktabs = TRUE, escape = FALSE,
      align = "lrrrr",
      caption = "Median absolute percent error of the deterministic Schaefer hindcast across shifter variants. \\label{tab:stress-mape}") %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10) %>%
  footnote(general = paste(
      "Median absolute percent error (in percent) of one-step-ahead deterministic predictions of biomass",
      "over the 2000-2024 estimation window, evaluated at the bounded least-squares estimate of",
      "(rho_SST, rho_CHL) restricted to [-3, 3]^2. The biological parameters (r_0, K) are held fixed at",
      "the centres of the official IFOP/SPRFMO priors. Anchoveta and sardina común have N = 24 yearly",
      "predictions; jack mackerel has N = 17 (the seven non-surveyed years and two left-censored years",
      "are excluded from the MAPE calculation). The 20% reference threshold is descriptive only and plays",
      "no role in the structural likelihood."),
      general_title = "Note:", escape = TRUE, threeparttable = TRUE)
```

The headline finding is that no variant brings any of the three stocks
below the $20\%$ MAPE threshold. The two coastal-upwelling stocks
benefit from activating SST: anchoveta moves from a base error of
$95.9\%$ to $30.8\%$ once $\rho_i^{SST}$ is allowed to vary, and
sardina común from $98.5\%$ to $23.1\%$. The further activation of
log-CHL improves both stocks marginally---to $26.6\%$ and $20.5\%$,
respectively---but neither closes the threshold. Activating log-CHL on
its own delivers essentially no improvement, in line with the absence
of contemporaneous correlation between the two environmental forcings
(see below). Jack mackerel hindcast errors remain in the
$44$--$49\%$ range across all variants, with no shifter variant
distinguishably better than the base.

The bounded least-squares coefficients of the joint variant deliver an
internally coherent biological reading for the two coastal-upwelling
stocks. For anchoveta, both shifters carry negative point estimates of
similar magnitude ($\rho_{\text{anch}}^{SST} \approx -2.3$,
$\rho_{\text{anch}}^{CHL} \approx -2.3$), consistent with a stock whose
short-run productivity is depressed by both warming and high
chlorophyll-a anomalies. For sardina común, the SST shifter is also
negative ($\rho_{\text{sard}}^{SST} \approx -2.0$) but the log-CHL
shifter switches sign relative to anchoveta
($\rho_{\text{sard}}^{CHL} \approx +2.1$), consistent with the warm-water,
mesotrophic affinity that distinguishes sardina from anchoveta in the
Humboldt-Current biological literature. For jack mackerel, the SST
shifter is small and positive ($\rho_{\text{jurel}}^{SST} \approx +0.6$)
and the log-CHL shifter is pinned at the lower boundary of the
admissible box, $\rho_{\text{jurel}}^{CHL} = -3$; we interpret the
boundary solution as a failure of point identification rather than as
a finding of strong climate forcing. Figure \ref{fig:stress-traj}
displays the hindcast trajectories under each variant alongside the
observed series.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.95\textwidth]{../data/bio_params/qa/hindcast_sst_trajectories.png}
\caption{Deterministic Schaefer hindcast trajectories under each
shifter variant, by stock. Solid line: observed biomass; dashed lines:
hindcast under base (no shifter), $+$SST only, $+\log$CHL only, and
joint $+$SST$+\log$CHL specifications. The joint specification
delivers the lowest median absolute percent error for the two
coastal-upwelling stocks; jack mackerel is essentially insensitive to
the shifter activation. Biological parameters $(r_i^{0}, K_i)$ are
fixed at the centres of the official IFOP/SPRFMO priors.}
\label{fig:stress-traj}
\end{figure}

## Identifiability diagnostic {#sec:appendix-stress-diag}

Two independent diagnostics support the use of weakly informative priors
on $(\rho_i^{SST}, \rho_i^{CHL})$ in the structural model rather than a
direct maximum-likelihood treatment.

First, the two environmental forcings are statistically orthogonal over
the 2000--2024 sample. The Pearson correlation between annual
Centro-Sur SST and log-CHL anomalies is $r = 0.030$ (two-sided
$p = 0.888$, $N = 25$). Collinearity between regressors therefore plays
no role in the difficulty of identifying the shifters; a variance
inflation diagnostic on the joint specification of
Eq. \eqref{eq:stress-shifter} returns values close to one for both
covariates. The coexistence of an essentially zero linear correlation
between the regressors and the wide cross-variant gap between
hindcast errors must therefore reflect features of the
non-linear-recursive law of motion in
Eq. \eqref{eq:stress-law}, not features of the design matrix.

Second, the joint MLE for jack mackerel binds at the lower boundary of
the admissible box. Across the 2000--2024 sample the climate
sensitivity of the Centro-Sur stock cannot be resolved within a
deterministic single-species hindcast: catches and the seven
non-surveyed years dominate the trajectory and the climate signal is
not separately identifiable from process and observation noise
absorbed into the residual. This pattern recurs in the initial fit of
the structural state-space specification reported in
Section \ref{identification}, where the marginal posterior of
$(\rho_{\text{jurel}}^{SST}, \rho_{\text{jurel}}^{CHL})$ tracks the
prior with no information added by the data. The structural model is
therefore the natural setting for jack mackerel: the absence of
information is correctly reported as a wide marginal posterior, rather
than as a boundary point estimate that a maximum-likelihood algorithm
would deliver. For the two coastal stocks, the joint MLE returns
interior solutions but with standard errors that are not separately
informative without an explicit accounting of process and observation
noise; the Bayesian state-space specification with priors elicited
from this hindcast addresses both limitations within a single
inferential framework.

## Translation to Bayesian priors {#sec:appendix-stress-priors}

The priors on $(\rho_i^{SST}, \rho_i^{CHL})$ used in the state-space
likelihood of Section \ref{sec:stock-dynamics} are reported in
Table \ref{tab:stress-priors}. For each coastal stock, the prior centre
is taken from the joint MLE of Section \ref{sec:appendix-stress-results},
rounded to one decimal place; for jack mackerel, where the joint MLE
binds at the boundary, the prior is centred at zero and is
deliberately uninformative. All priors are independent normal
densities with unit standard deviation, which is wide enough to allow
the structural likelihood to update the centre by more than one full
unit of semi-elasticity in either direction within a single posterior
standard deviation. We do not impose hierarchical pooling across
stocks: the sign reversal of $\rho_i^{CHL}$ between anchoveta and
sardina común is biologically interpretable and would be erased by a
pooled prior with a common hyper-mean.

```{r stress-priors-table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
priors_tbl <- data.frame(
  Stock        = c("Anchoveta", "Sardina común", "Jack mackerel"),
  `rho_SST_mu` = c("-2.3", "-2.0", "0.0"),
  `rho_SST_sd` = c("1.0",  "1.0",  "1.0"),
  `rho_CHL_mu` = c("-2.3", "+2.1", "0.0"),
  `rho_CHL_sd` = c("1.0",  "1.0",  "1.0"),
  Source       = c("Joint MLE",
                   "Joint MLE",
                   "Boundary solution; vague prior"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

names(priors_tbl) <- c(
  "Stock",
  "$\\mu_{\\rho^{SST}}$",
  "$\\sigma_{\\rho^{SST}}$",
  "$\\mu_{\\rho^{CHL}}$",
  "$\\sigma_{\\rho^{CHL}}$",
  "Source"
)

kable(priors_tbl, format = "latex", booktabs = TRUE, escape = FALSE,
      align = "lrrrrl",
      caption = "Independent normal priors on the climate shifters used in the state-space likelihood of Section \\ref{sec:stock-dynamics}. \\label{tab:stress-priors}") %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10) %>%
  footnote(general = paste(
      "Prior centres for the coastal-upwelling stocks are the joint maximum-likelihood point estimates",
      "of (rho_SST, rho_CHL) from the deterministic Schaefer hindcast of Appendix A.2, rounded to one",
      "decimal place. For jack mackerel the joint MLE binds at the lower boundary of the admissible box",
      "(rho_CHL = -3); the prior is centred at zero so that posterior identification is driven by the",
      "structural likelihood rather than by the boundary solution. All priors are independent normal",
      "with sd = 1.0; no hierarchical pooling is imposed because rho_CHL switches sign between anchoveta",
      "and sardina común."),
      general_title = "Note:", escape = TRUE, threeparttable = TRUE)
```

The interpretation of the resulting posterior is straightforward. For
the two coastal stocks, a posterior centred close to the prior centre
with a posterior standard deviation appreciably below one is evidence
that the structural state-space likelihood confirms the reduced-form
hindcast point estimate; a posterior shifted by more than one prior
standard deviation indicates that the structural model has extracted
information that the deterministic hindcast misses, most plausibly by
disentangling process noise from observation noise. For jack mackerel,
a posterior that essentially reproduces the prior is the correct
report of an absence of identifying information at the Centro-Sur
scale, as discussed in Section \ref{identification}.
