# Predictive diagnostics {#appendix-predictive}

This appendix reports the nested predictive comparison referenced in
Section \ref{sec:stock-dynamics}, comparing three state-space
specifications: (i) species-independent dynamics without climate
shifters (\emph{ind}), (ii) the same with a cross-stock residual
covariance $\Omega$ (\emph{omega}), and (iii) the full model
augmented with the climate shifter of Eq. \eqref{eq:shifter}
(\emph{full}). Two predictive metrics are reported: leave-one-out
PSIS-LOO (Pareto-smoothed importance sampling) and one-step-ahead LFO
(leave-future-out).

```{r loo-appendix-table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
loo_tbl <- read.csv("tables/loo_t4b_compare.csv", stringsAsFactors = FALSE) %>%
  dplyr::transmute(
    Model                                     = model,
    `$\\widehat{\\text{ELPD}}_{\\text{loo}}$` = sprintf("%.2f", elpd_loo),
    `SE`                                      = sprintf("%.2f", se_elpd_loo),
    `$\\Delta\\widehat{\\text{ELPD}}$`        = ifelse(elpd_diff == 0, "--",
                                                      sprintf("%.2f", elpd_diff)),
    `SE$_{\\Delta}$`                          = ifelse(se_diff == 0, "--",
                                                      sprintf("%.2f", se_diff)),
    `$\\hat{p}_{\\text{loo}}$`                = sprintf("%.1f", p_loo)
  )

kable(loo_tbl, format = "latex", booktabs = TRUE, escape = FALSE,
      align = "lrrrrr",
      caption = "PSIS-LOO comparison across state-space specifications. \\label{tab:loo-appendix}") %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10) %>%
  footnote(general = paste(
      "\u0394 ELPD is the pairwise difference relative to T4b-full (the leading model);",
      "the adjacent SE column reports the standard error of that difference.",
      "After moment matching, 65/68 observations satisfy Pareto-k <= 0.7 for T4b-full,",
      "64/68 for T4b-ind, and 67/68 for T4b-omega. The remaining high-Pareto-k points are",
      "the two censored jack-mackerel years (2012, 2015), which appear in all three models",
      "and therefore do not affect the ranking."),
      general_title = "Note:", escape = FALSE, threeparttable = TRUE)
```

PSIS-LOO favours the climate-augmented specification with a positive
$\Delta$ELPD of $14$--$18$ relative to the climate-free specifications,
indicating that the in-sample predictive performance improves when the
climate shifter is included.

```{r lfo-appendix-table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
lfo_tbl <- read.csv("tables/lfo_t4b_compare.csv", stringsAsFactors = FALSE) %>%
  dplyr::transmute(
    Model                                    = model_label,
    `Points`                                 = n_points,
    `Censored`                               = n_censored,
    `$\\widehat{\\text{ELPD}}_{\\text{LFO}}$` = sprintf("%.2f", elpd_lfo_sum),
    `Mean per point`                         = sprintf("%.2f", elpd_lfo_mean),
    `$\\Delta$ vs leader`                    = ifelse(elpd_diff == 0, "--",
                                                     sprintf("%.2f", elpd_diff))
  )

kable(lfo_tbl, format = "latex", booktabs = TRUE, escape = FALSE,
      align = "lcccrr",
      caption = "One-step-ahead LFO comparison across state-space specifications. \\label{tab:lfo-appendix}") %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10) %>%
  footnote(general = paste(
      "Sum of one-step predictive log densities over five origin years",
      "(2011, 2014, 2017, 2020, 2022) and three stocks (14 targets per model; two are censored).",
      "Each refit uses 4 chains \u00d7 1500 warmup + 1500 sampling iterations.",
      "All 15 refits satisfied R-hat <= 1.01 with at most one divergent transition",
      "(T4b-omega at the 2011 origin), which does not affect the reported ELPD."),
      general_title = "Note:", escape = FALSE, threeparttable = TRUE)
```

The one-step-ahead LFO comparison is essentially a tie across the three
specifications, indicating that the climate-augmented specification does
not deliver a detectable improvement in short-horizon forecast
performance over the climate-free alternatives. We interpret this as a
sample-size issue rather than a deficiency of the climate
specification: the structural climate elasticities are identified on
the full $25$-year window, not on one-step-ahead increments. The
posterior of the climate-augmented specification remains our primary
object of interpretation in the main text.
