## Model selection via leave-one-out cross-validation {#loo-comparison}

We compare three nested specifications of the Schaefer state-space
model to isolate the predictive contribution of residual cross-stock
covariance and environmental forcing. The three specifications share
identical biological priors (adopted from the IFOP and SPRFMO assessments)
and differ only in the joint structure of the process error:

- **T4b-ind** treats the three stocks as conditionally independent given
  their biological parameters. Process errors are mutually uncorrelated.
- **T4b-omega** introduces a full $3\times3$ residual correlation matrix
  $\Omega$ with an LKJ$(4)$ prior, allowing contemporaneous co-movement
  across species to be learned from the data.
- **T4b-full** adds stock-specific environmental shifters, modulating the
  intrinsic growth rate as
  $r_{i,t} = r_i^{0} \, \exp\!\big( \rho_i^{SST} (SST_t - \overline{SST})
  + \rho_i^{CHL} (\log CHL_t - \overline{\log CHL}) \big),$
  in addition to retaining the $\Omega$ structure.

Predictive performance is assessed by Pareto-smoothed importance sampling
leave-one-out cross-validation [PSIS-LOO; @Vehtari2017-qe] with moment
matching applied to observations whose Pareto-$\hat{k}$ diagnostic exceeded
0.7. The expected log pointwise
predictive density (ELPD) and the difference $\Delta\widehat{\text{ELPD}}$
across models, together with the standard error of the pairwise
difference, are reported in Table \ref{tab:loo-compare}.

```{r loo-compare-table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
loo_tbl <- read.csv("tables/loo_t4b_compare.csv", stringsAsFactors = FALSE) %>%
  dplyr::transmute(
    Model       = model,
    `$\\widehat{\\text{ELPD}}_{\\text{loo}}$` = sprintf("%.2f", elpd_loo),
    `SE`                                      = sprintf("%.2f", se_elpd_loo),
    `$\\Delta\\widehat{\\text{ELPD}}$`        = ifelse(elpd_diff == 0, "--",
                                                      sprintf("%.2f", elpd_diff)),
    `SE$_{\\Delta}$`                          = ifelse(se_diff == 0, "--",
                                                      sprintf("%.2f", se_diff)),
    `$\\hat{p}_{\\text{loo}}$`                = sprintf("%.1f", p_loo),
    `LOOIC`                                   = sprintf("%.2f", looic)
  )

kable(loo_tbl, format = "latex", booktabs = TRUE, escape = FALSE,
      align = "lrrrrrr",
      caption = "PSIS-LOO cross-validation of three nested state-space specifications. \\label{tab:loo-compare}") %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10) %>%
  footnote(general = paste(
      "Models ordered from highest to lowest $\\widehat{\\text{ELPD}}_{\\text{loo}}$.",
      "$\\Delta\\widehat{\\text{ELPD}}$ is the pairwise difference with respect to the leading model (T4b-full);",
      "SE$_{\\Delta}$ is the standard error of that difference.",
      "All three fits rely on moment-matched PSIS-LOO; 65/68 observations satisfy $\\hat{k} \\leq 0.7$ in T4b-full,",
      "64/68 in T4b-ind, and 67/68 in T4b-omega.",
      "The remaining high-$\\hat{k}$ observations correspond predominantly to the two censored jack-mackerel years (2012, 2015), which appear equally across all three specifications and therefore do not affect the ranking.",
      "$\\hat{p}_{\\text{loo}}$ is the effective number of parameters implied by LOO."),
      general_title = "Note:", escape = FALSE, threeparttable = TRUE)
```

The full environmental specification attains the highest predictive density
by a wide margin. T4b-full outperforms T4b-omega by
$\Delta\widehat{\text{ELPD}} = 18.02$ ($\text{SE} = 5.59$) and the
environment-free but covariance-rich T4b-ind by
$\Delta\widehat{\text{ELPD}} = 14.59$ ($\text{SE} = 4.93$). Both gaps
exceed three standard errors and therefore constitute decisive evidence
that direct environmental forcing carries predictive information beyond
what residual cross-stock correlation can absorb.

A second finding with substantive content is that **T4b-omega is worse than
T4b-ind** ($\Delta\widehat{\text{ELPD}} = -3.43$). Introducing the $\Omega$
covariance matrix without environmental shifters is penalised by LOO. The
natural reading is that the cross-stock residual correlation captured by
T4b-omega is not evidence of biological interaction but rather a statistical
footprint of shared environmental forcing. Once SST and CHL enter the growth
equation directly (T4b-full), the marginal value of $\Omega$ collapses and
its additional degrees of freedom are no longer paid for by the data.

A stock-by-stock decomposition of the ELPD (Table \ref{tab:loo-by-stock} and
Figure \ref{fig:loo-pareto-k}) shows that the full-specification gains are
concentrated in the two coastal pelagics. In T4b-full, the per-stock ELPD
rises by $+4.68$ for anchoveta and $+9.12$ for sardina relative to T4b-ind,
while for jack mackerel the three specifications are statistically
indistinguishable ($\text{ELPD} \in [-19.1, -18.0]$). This pattern is
mechanistically interpretable: anchoveta and sardina are upwelling-dependent
coastal stocks with demonstrated sensitivity to SST and chlorophyll at
annual scale, whereas jack mackerel is a migratory transboundary stock
whose dynamics at this resolution appear insensitive to local coastal
forcing.

```{r loo-by-stock-table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
by_stock <- read.csv("tables/loo_t4b_elpd_por_stock.csv", stringsAsFactors = FALSE) %>%
  dplyr::transmute(
    Model = model,
    Stock = stock_lab,
    `$N$` = n_obs,
    `$\\widehat{\\text{ELPD}}_{\\text{loo}}$` = sprintf("%.2f", elpd_loo),
    `SE` = sprintf("%.2f", se_elpd),
    `$\\hat{p}_{\\text{loo}}$` = sprintf("%.1f", p_loo),
    `$n(\\hat{k} > 0.7)$` = n_k_gt_07
  ) %>%
  dplyr::arrange(Stock, Model)

kable(by_stock, format = "latex", booktabs = TRUE, escape = FALSE,
      align = "llcrrrr",
      caption = "Per-stock PSIS-LOO decomposition. \\label{tab:loo-by-stock}") %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10) %>%
  footnote(general = paste(
      "ELPD is additive across observations, so the per-stock rows sum (up to rounding)",
      "to the aggregate $\\widehat{\\text{ELPD}}_{\\text{loo}}$ of Table \\ref{tab:loo-compare}.",
      "$n(\\hat{k} > 0.7)$ counts observations whose Pareto-$\\hat{k}$ exceeded the",
      "reliability threshold after moment matching; counts include the two censored",
      "jack-mackerel observations (2012, 2015), which appear in all three models."),
      general_title = "Note:", escape = FALSE, threeparttable = TRUE)
```

Figure \ref{fig:loo-pareto-k} displays the per-observation Pareto-$\hat{k}$
values against calendar year for each model and stock, with the
$\hat{k} = 0.7$ reliability threshold indicated. Figure
\ref{fig:loo-elpd-stock} shows $\Delta\widehat{\text{ELPD}}$ relative to
T4b-ind by stock, making the per-stock contribution of the full
specification explicit.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.85\textwidth]{../figs/t4b/loo_t4b_pareto_k.png}
\caption{Per-observation Pareto-$\hat{k}$ diagnostics of PSIS-LOO for the
three state-space specifications, faceted by stock and plotted against
calendar year. The dashed line marks $\hat{k} = 0.5$ (good) and the solid
line marks $\hat{k} = 0.7$ (reliability boundary). Crosses denote
censored jack-mackerel observations.}
\label{fig:loo-pareto-k}
\end{figure}

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.85\textwidth]{../figs/t4b/loo_t4b_elpd_stock.png}
\caption{$\Delta\widehat{\text{ELPD}}_{\text{loo}}$ by stock relative to
the independent-stocks baseline T4b-ind. Error bars show the per-stock
SE. The environmental specification T4b-full dominates for anchoveta and
sardina, while the three specifications are indistinguishable for jack
mackerel.}
\label{fig:loo-elpd-stock}
\end{figure}

### Diagnostic caveats

Two caveats are worth flagging for the reader. First, the effective
parameter count $\hat{p}_{\text{loo}}$ is large (approximately
$58$–$65$ for a sample of $68$ annual observations), which is expected for
state-space models with one latent process state per observation-year: LOO
requires re-estimation of neighbouring latent states when a single
observation is held out, and this inflates the effective dimensionality
[@Vehtari2017-qe, §6]. Because $\hat{p}_{\text{loo}}$ is similar across the
three specifications, this does not confound the comparison, but it does
signal that absolute $\widehat{\text{ELPD}}$ values should not be
interpreted as out-of-sample accuracy on genuinely future data. A
leave-future-out robustness check is reported in the Appendix.

Second, the two censored jack-mackerel observations (2012 and 2015) yield
$\hat{k} > 1$ across all three specifications, reflecting the known
difficulty of PSIS under leave-one-out for tail-probability contributions
(normal\_lcdf terms). Because these observations carry identical likelihood
weights across models, they neither alter the ranking nor the qualitative
interpretation.
